---
title: "MATH9102-W12-MLR Skeleton"
author: "Deirdre Lawless"
format: 
  html:
    self-contained: true
    toc: true      # Enables table of contents
    toc_float: true # Makes the TOC float on the side
    toc_depth: 3   # Specifies the depth of headers to include in TOC
editor: source

---


```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}


quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("lm.beta")   # For extracting standardized co-efficients from regression model
quiet_library("stargazer") # For formatting output of regression


quiet_library("performance")    # Loads tools for checking model assumptions and diagnostics (e.g., normality, collinearity)
quiet_library("car")            # Companion to Applied Regression – includes functions for influence tests and outlier detection
quiet_library("see")            # Needed for plots

quiet_library("ggResidpanel")   # Alternative for generating diagnostic plots for 4.5.x

# Code to load file
datalocation <- file.path("C:","Path", "to", "yourfiles")
datapath <- file.path(datalocation,'filename')
dataframename <- read.table(datapath) # change .table to .csv etc as needed


# Complete any manipulation/cleaning needed
```


# Step 1: Build the model
```{r}
#label: model1
#Create the model
model1<-lm(outcomevar~pred1+pred2+pred3, data=dataframname)
```

##  Step 2: Generate the Analysis of Variance Table
This provides the information needed to calculate the F statistic.
```{r}
#| label: anovam1
anova(model1)
```

## Step 3: Generate the summary statistics needed to assess model fit
```{r}
#| label: summm1
summary(model1)
```

## Step 4: Generate Standardized Co-efficients
In this example as the data is standardized these will not be different to the co-efficients produced by the summary.
```{r}
#| label: betam1
lm.beta::lm.beta(model1)
```

## Step 5 (optional): Use Stargazer to produce a tidy output of ANOVA and Summary
```{r stargazer}
stargazer(model1, type="text") #Tidy output of all the required stats
```

## Step 6: Report the results
Replace the content in italics with content relevant to your analysis (does not need to be italicised). 


A multiple linear regression model was created to _explain what the model was created for_.  A _statistically/non statitsically_ significant regression equation was found (F(_df1,df2_)=_FRatio_, p = _pvalue (if not significant)/p < relevant cut-of (if significant < 0.5, 0.1 or 0.001)_ with an adjusted R^2^ of _adjusted R2 from summary_. The regression equation is therefore:

  _Outcome Concept = intercept + b1 * pred1 + b2 * pred2 + b3 * pred3_
  
This indicates _explain what increases/changes in each predictor mean for the outcome_. _Write a statement about what this means for the hypothesis_.
 
 
Remember for categorical predictors you are comparing to the reference category.



| Statistic needed in written report                      | Where to find it in R output                                                    | Code chunk label |     |          |
| ------------------------------------------------------- | ------------------------------------------------------------------------------- | ---------------- | --- | -------- |
| **F-statistic**                                         | In the **ANOVA table** (`F value` column)                                       | `anovam1`        |     |          |
| **Model degrees of freedom (df1, df2)**                 | In **ANOVA** output (`Df`) <br> *df1 = model predictors* <br> *df2 = residuals* | `anovam1`        |     |          |
| **p-value for the model (overall test)**                | In **ANOVA** (`Pr(>F)`) — same row as F-statistic                               | `anovam1`        |     |          |
| **Adjusted R²**                                         | In the **Regression Summary** (`Adjusted R-squared`)                            | `summm1`         |     |          |
| **Unstandardized coefficients** (Intercept, b1, b2, b3) | In **Regression Summary** (`Estimate`)                                          | `summm1`         |     |          |
| **p-values for predictors**                             | In **Regression Summary** (`Pr(>|t| )`) | `summm1` |
| **Standardized coefficients** (β values)                | In standardized regression output                                               | `betam1`         |     |          |
| **Optional formatted version of the above**             | Stargazer formatted summary output                                              | `stargazer`      |     |          |



# Residual Analysis

## Step 1: Generate Summary Statistics/Non-Zero Variance

```{r}
#Generate summary statistics
#If using   pastecs::stat.desc for continous variables - make sure you include the basic switch=F to ensure you don't get scienfitic notation


```

## Step 2: Residual Analysis

### Step 2a: Conduct a Normality Test

```{r}
# Check normality - conducts a Shapiro-Wilks test and checks the standardized residuals for normal distribution
# A p-value < 0.05 indicates a significant deviation from normal distribution.

performance::check_normality(model1)

```

### Step 2b: Visualize residual distribution

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Residual Distribution Plots for Model 1"
# Visualize residual distribution


# Histogram of Residuals
ggplot(data.frame(residuals = resid(model1)), aes(x = residuals)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 20, 
                 fill = "skyblue", 
                 color = "white") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(resid(model1)), sd = sd(resid(model1))), 
                color = "red", 
                linewidth = 1) +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Density") +
  theme_minimal()

# A transformed version of a Q–Q plot displayed horizontally
plot(check_normality(model1))

# Q-Q plot of standardized residuals
qqnorm(rstandard(model1))
qqline(rstandard(model1), col = "red")

```

### Step 2c: Check outliers or influential points

```{r}
# Check for outliers or influential points

performance::check_outliers(model1)

```

### Step 2d: Examine the residuals

```{r}
# Extract residuals
residuals <- resid(model1)

# Standardize the residuals
std_residuals <- residuals / sd(residuals)

# Calculate max and min of the standardized residuals
max_std_residual <- max(std_residuals)
min_std_residual <- min(std_residuals)

# Print the results
cat("Max standardized residual:", max_std_residual, "\n")
cat("Min standardized residual:", min_std_residual, "\n")

#Identify potential outliers in the dataset based on the studentized residuals of the model.
car::outlierTest(model1) # Bonferonni p-value for most extreme observations - observations with a p value lower than the threshold are consider outliers.
# In this case there is only one row 

```

## Step 3: Check homoscedasticity (equal variance)

```{r}
performance::check_heteroscedasticity(model1)

```

## Step 4: Check multi-colinearity

```{r}
# Check multicollinearity

performance::check_collinearity(model1)

```

## Step 5: Report
Replace the content in italics with content relevant to your analysis (does not need to be italicised).

The data met the assumption of non-zero variances _report the variance for your continous predictors_. Examination of diagnostic plots indicated that the residuals were _approximately normally distributed/not normally distributed_, with a density plot _resembling normality/deviating from normality_, although the formal test suggested _what the test suggested)_ _(p=?)_. The standardized residuals ranged from _report max and min_  with _x/no_ observations exceeding ±3 standard deviations. The Bonferroni-adjusted outlier test (p = ?) indicated that _what this meant_, as Cook’s distance _remained below/exceeded_ the conservative threshold of 1.0. The residuals-versus-fitted values plot displayed _describe the scatter_  around the zero line, with _some/no_ evidence of heteroscedasticity _(p = ?)_, _supporting/deviating from_ the assumption of constant variance. Collinearity statistics showed _low/moderate/hig_ intercorrelations between predictors _(VIF and Tolerance for each predictor)_, confirming that multicollinearity _was/was not_ a concern. Overall, the model _met/did not meet_ the key assumptions of multiple linear regression—linearity, independence, homoscedasticity, and low multicollinearity—with only a minor departure from normality and a single non-influential outlier. The model can therefore be considered _to/not to_ provide a reliable estimate of the relationship between _list the outcome and the predictor concepts_.


| Assumption / Statistic required in report     | What to look for                                             | Where to find it                                                          | Code chunk label             |
| --------------------------------------------- | ------------------------------------------------------------ | ------------------------------------------------------------------------- | ---------------------------- |
| **Normality of residuals**                    | p-value from Shapiro–Wilk or check_normality result          | Output of `performance::check_normality(model1)`                          | first in Step 2a |
|                                               | Visual inspection: *distribution resembles normality or not* | Histogram + Q-Q plots                                                     | Step 2b plots                |
| **Standardized residuals range**              | Max and min standardized residual values                     | Printed result of `cat(...)`                                              | Step 2d                      |
| **Outlier test**                              | Bonferroni p-value for most extreme case                     | Output of `car::outlierTest(model1)`                                      | Step 2d                      |
| **Influence**                                 | Cook’s Distance < 1.0                                        | In diagnostic plots (`plot(model1)` → Cook’s) or within `check_model()`   | Step 6                       |
| **Homoscedasticity (Equal variance)**         | p-value from heteroscedasticity test                         | `performance::check_heteroscedasticity(model1)` Scale-Location Plot                           | Step 3                       |
| **Linearity**                                 | Evidence of non-random scatter around fitted line            | Residuals vs Fitted plot in `plot(model1, which = 1)` or `check_model()`  | Step 6                       |
| **Multicollinearity**                         | VIF and Tolerance values                                     | `performance::check_collinearity(model1)`                                 | Step 4                       |



## Step 6: Generate Summary Diagnostic Plots

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Diagnostic plots for Model 1"
# Overall diagnostic plots

# If using R 4.4.x use
performance::check_model(model1,
  check = c(
    "linearity",
    "homoscedasticity",
    "outliers",
    "collinearity",
    "normality"
  )
)

# If using R 4.5.x Show all 4 diagnostic plots in one figure using R basic plots
par(mfrow = c(2, 2))   # 2x2 grid layout
plot(model1)            # Generates: Residuals vs Fitted, Normal Q-Q, Scale-Location, Cook’s distance
par(mfrow = c(1, 1))    # Reset layout

```

```{r}
#| fig.cap: " Diagnostic plots for Model - Alternative approaches"
# Alternative for 4.5.x
car::residualPlots(model1)



ggResidpanel::resid_panel(model1)


```
