---
title: "Week 9: Checking Linear Regression Assumptions"
author: "MATH9102: Fundamentals of Data Analysis"
format:
  html:
    self-contained: true
    theme: cosmo
    toc: true
---

```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}

/* Center and bold table captions, force them below tables */
.table-caption, 
caption {
  caption-side: bottom;  /* caption below table */
  text-align: center;
  color: black;
  font-weight: bold;
}
</style>
```

```{r}
#| label: install-and-load
#| message: false
#| warning: false

# Set CRAN mirror for non-interactive sessions
options(repos = c(CRAN = "https://cloud.r-project.org"))

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}

quiet_library("tidyverse")      # Loads the tidyverse suite (ggplot2, dplyr, etc.) for data handling and visualisation
quiet_library("performance")    # Loads tools for checking model assumptions and diagnostics (e.g., normality, collinearity)
quiet_library("ppcor")          # Provides functions for calculating partial and semi-partial correlations
quiet_library("car")            # Companion to Applied Regression – includes functions for influence tests and outlier detection
quiet_library("see")            # Needed for plots
quiet_library("pastecs")        #Needed for descriptive stats


#Amend this to reflect your folder structure 
datalocation <- file.path("C:","Users", "deirdre.lawless", "datafiles")
datapath <- file.path(datalocation,'sleep.rdata')

#Read in the dataset - sleep dataset from Pallant
sleep<- read.table(datapath, header=TRUE)
sleep <-na.omit(sleep)

```

## Overview

In this exercise, you will learn **how to check the key assumptions of a linear regression model** using the `sleep.rdata` dataset. You’ll test each assumption step by step, interpret the results, and reflect on possible model improvements.

By the end of this activity, you should be able to:

- Verify **linearity** and **homoscedasticity** using residual plots  
- Check **normality of residuals**  
- Detect **influential observations**  
- Test for **multicollinearity**  

## How to Use This Exercise



This exercise mirrors the structure and tone of the Week 9 lecture notes. For each assumption, you will: 1) **Run** a short block of code, 2) **Inspect** the output/plot, and 3) **Answer** a brief interpretation prompt.

Keep your answers concise and evidence-based (reference the specific plot/test and its key indicators).
---

## 1. Fit a Linear Regression Model

Fit a regression model predicting **sleep quality** from relevant predictors. 
High stress and poor sleep behaviours are known predictors of daytime sleepiness, not typically its effects. In this task you will build two models:

Model 1 : "Does week night sleep predict the degree of sleepiness/fatigue experienced and does the effect of weeknight sleep duration on sleepiness or fatigue differ across stress levels?”

Model 2:"Does weeknight sleep duration predict levels of sleepiness or fatigue, and do stress level and gender moderate this relationship?"


| Variable             | Concept measured                                                                   | Type                | Typical role in regression                                        |
| -------------------- | ---------------------------------------------------------------------------------- | ------------------- | ----------------------------------------------------------------- |
| `totsas`             | Sleepiness, fatigue, low energy, tiredness, lethargy (aggregate of multiple items) | Continuous (5–50)   | **Outcome** – a holistic measure of sleep quality and its effects |
| `stressmo`           | Perceived stress level last mont                                                  | Categorical (1–10)   | **Predictor** – a psychological factor that may affect sleepiness |
                                                 |
| `hourwnit`           | Hours of sleep on weeknights                                                       | Continuous          | **Predictor**                                                     |

| `gender`     | Demographic controls                                                               | Binary / continuous | **Predictors or control variables**                               |


# Model 1
```{r}
# Example model (edit as appropriate)
model1 <- lm(totsas ~ stressmo + hourwnit,   data = sleep)


summary(model1)
```
A multiple linear regression model was conducted to examine whether perceived stress and average hours of sleep on weeknights could significantly predict daytime sleepiness and fatigue as measured by the totsas score. The overall model was statistically significant, F(6, 3) = 106.20, p = 0.001, explaining approximately 99.5% of the variance in sleepiness (Adjusted R² = 0.986).

The fitted regression equation was:

Predicted Total Sleepiness and Associated Sensation Score  = 33.44 
                          + 29.00 x Moderate perceived stress (level 5) 
                          + 23.47 × Moderately high perceived stress (level 6) 
                          + 26.84 × High perceived stress (level 7) 
                          + 30.00 × Very high perceived stress (level 8) 
                          + 6.81 × Extremely high perceived stress (level 9) 
                          – 3.06 × Average hours of sleep on weeknights

Higher perceived stress levels were associated with significantly greater total sleepiness and sensations scores (p < 0.01), whereas longer weekday sleep duration was linked to lower scores (p = 0.02). This suggests that participants reporting higher stress levels tend to experience greater sleepiness, while each additional hour of weeknight sleep corresponds to an estimated 3-point reduction in the total sleepiness score.


## Assess the Assumptions
### Step 1: Generate Summary Statistics

```{r}
pastecs::stat.desc(sleep$totsas, basic = FALSE)
pastecs::stat.desc(sleep$hourwnit, basic = FALSE)

# Frequency table
table(sleep$stressmo)
# Proportion table
prop.table(table(sleep$stressmo))

```

**Task:** Review means, ranges, and variances. Are the predictors measured on compatible scales? Any extreme values?

### Step 2: Residual Analysis

#### Step 2a. Normality Test

```{r}
performance::check_normality(model1)
```

**Task:** Does the p-value suggest that residuals deviate significantly from normality? What does this mean for the model?

#### Step 2b. Visualize Residual Distribution

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Residual Distribution Plots for Model 1"

ggplot(data.frame(residuals = resid(model1)), aes(x = residuals)) +
  geom_histogram(aes(y = ..density..), bins = 20, fill = "skyblue", color = "white") +
  stat_function(fun = dnorm, args = list(mean = mean(resid(model1)), sd = sd(resid(model1))), color = "red", linewidth = 1) +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Density") +
  theme_minimal()

plot(check_normality(model1))
qqnorm(rstandard(model1)); qqline(rstandard(model1), col = "red")
```

**Task:** Inspect the histogram and Q–Q plot. Do the residuals approximate a normal distribution?

#### Step 2c. Outliers or Influential Points

```{r}
performance::check_outliers(model1)
```

**Task:** Note any cases flagged as outliers. What threshold is used?

#### Step 2d. Examine Residuals

```{r}
residuals <- resid(model1)
std_residuals <- residuals / sd(residuals)
max(std_residuals); min(std_residuals)
car::outlierTest(model1)
```

**Task:** Identify any standardized residuals exceeding ±3.29. How might they influence model fit?

### Step 3: Check Homoscedasticity

```{r}
performance::check_heteroscedasticity(model1)
```

**Task:** Is there evidence of unequal variance across fitted values?

### Step 4: Check Multicollinearity

```{r}
performance::check_collinearity(model1)
```

**Task:** What are the VIF and Tolerance values? Are they within acceptable limits (\< 5 and \> 0.2)?

### Step 6: Summary Diagnostic Plots

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Diagnostic plots for Model 1"

# If using R 4.4.x
performance::check_model(model1,
  check = c(
    "linearity",
    "homoscedasticity",
    "outliers",
    "collinearity",
    "normality"
  )
)

# If using R 4.5.x Show all 4 diagnostic plots in one figure using R basic plots
par(mfrow = c(2, 2))   # 2x2 grid layout
plot(model1)            # Generates: Residuals vs Fitted, Normal Q-Q, Scale-Location, Cook’s distance
par(mfrow = c(1, 1))    # Reset layout
```
```{r}
#| fig.cap: " Diagnostic plots for Model - Alternative approaches"
# Alternative for 4.5.x
car::residualPlots(model1)

# Nicer alternative for 4.5.x
quiet_library("ggResidpanel")

ggResidpanel::resid_panel(model1)


```




**Task:** Review the combined diagnostics. Which assumptions are well met, and which may require attention?


### Task: Write the report
Visual inspection of the diagnostic plots indicated that the relationship between the predictors and the outcome was broadly linear. There was evidence of a broadly linear relationship between the predictors and the outcome, with only minor variation in the spread of residuals across predicted values. These small fluctuations suggest that the assumptions of linearity and approximate homocedasticity were largely satisfied.

The normality of residuals was examined both statistically and visually. The Shapiro–Wilk test indicated no significant deviation from normality (p = 0.959), and both the histogram and Q–Q plot suggested that residuals were approximately normally distributed around zero. The normality assumption was therefore considered satisfied.

No standardized residuals exceeded ±3.29, and the Bonferroni-adjusted outlier test did not identify any influential cases. Cook’s distance values were all well below 1.0, confirming that no single observation exerted undue influence on the regression estimates.

Variance Inflation Factors (VIFs) for the predictors were below 5 (VIF = 4.13), and tolerance values were above 0.2 (Tolerance = 0.24), indicating acceptable independence among predictors and no evidence of problematic multicollinearity.


Taken together, the diagnostic checks suggested that the model met the core assumptions of multiple linear regression. Linearity, normality, and multicollinearity conditions were well supported, and the variance of residuals was broadly consistent across predicted values. The model can therefore be regarded as suitable for interpreting the relationship between perceived stress, sleep duration, and daytime sleepiness.

------------------------------------------------------------------------

# 3. Model 2 — Add Gender as a Predictor

Now extend the model by adding `gender` as a control variable.

**Outcome variable:** `totsas`\
**Predictors:** `stressmo`, `hourwnit`, `anxiety`, `sex`

```{r}
# Ensure gender is coded as numeric (0/1) if not already
sleep$gender <- ifelse(sleep$sex == "male", 0, ifelse(sleep$sex == "female", 1, NA))

# Fit the extended model
model2 <- lm(totsas ~ stressmo + hourwnit + gender, data = sleep)
summary(model2)
```
A second multiple linear regression model was fitted to determine whether gender further improved prediction of daytime sleepiness and fatigue when combined with perceived stress and weeknight sleep duration. The overall model was statistically significant, F(7, 2) = 433.90, p = 0.002, explaining approximately 99.9% of the variance in sleepiness (Adjusted R² = 0.997).

The fitted regression equation was:

Sleepiness (totsas) = 36.84 
        + 29.00 × Moderate perceived stress (level 5)
        + 23.93 × Moderately high perceived stress (level 6)
        + 27.51 × High perceived stress (level 7)
        + 30.82 × Very high perceived stress (level 8)
        + 4.65 × Extremely high perceived stress (level 9)
        – 3.78 × Average hours of sleep on weeknights
        + 1.65 × Gender (1 = female, 0 = male)

Higher perceived stress continued to predict significantly greater total sleepiness and sensation scores (p < 0.01), and more hours of sleep on weeknights remained associated with reduced scores (p < 0.01). Gender was a marginal predictor (p = 0.07), suggesting that females (coded 1) reported slightly higher levels of total sleepiness after controlling for stress and sleep duration.

## Assess Assumptions
### Step 1: Generate Summary Statistics

```{r}
pastecs::stat.desc(sleep$gender, basic = FALSE)
table(sleep$gender, useNA = "ifany")
prop.table(table(sleep$gender))
```

**Task:** Are both genders adequately represented in the dataset?

### Step 2: Residual Analysis

#### Step 2a. Normality Test

```{r}
performance::check_normality(model2)
```

#### Step 2b. Visualize Residual Distribution

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Residual Distribution Plots for Model 2"
ggplot(data.frame(residuals = resid(model2)), aes(x = residuals)) +
  geom_histogram(aes(y = ..density..), bins = 20, fill = "skyblue", color = "white") +
  stat_function(fun = dnorm, args = list(mean = mean(resid(model2)), sd = sd(resid(model2))), color = "red", linewidth = 1) +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Density") +
  theme_minimal()

plot(check_normality(model2))
qqnorm(rstandard(model2)); qqline(rstandard(model2), col = "red")
```

#### Step 2c. Outliers or Influential Points

```{r}
performance::check_outliers(model2)
```

#### Step 2d. Examine Residuals

```{r}
residuals <- resid(model2)
std_residuals <- residuals / sd(residuals)
max(std_residuals); min(std_residuals)
car::outlierTest(model2)
```

### Step 3: Check Homoscedasticity

```{r}
performance::check_heteroscedasticity(model2)
```

### Step 4: Check Multicollinearity

```{r}
performance::check_collinearity(model2)
```

### Step 5: Summary Diagnostic Plots

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Diagnostic plots for Model 2"
# If using R 4.4.x use
performance::check_model(model2,
  check = c(
    "linearity",
    "homoscedasticity",
    "outliers",
    "collinearity",
    "normality"
  )
)
# If using R 4.5.x use
# If using R 4.5.x Show all 4 diagnostic plots in one figure using R basic plots
par(mfrow = c(2, 2))   # 2x2 grid layout
plot(model1)            # Generates: Residuals vs Fitted, Normal Q-Q, Scale-Location, Cook’s distance
par(mfrow = c(1, 1))    # Reset layout
```

```{r}
#| fig.cap: " Diagnostic plots for Model - Alternative approaches"
# Alternative for 4.5.x
car::residualPlots(model2)

# Nicer alternative for 4.5.x
quiet_library("ggResidpanel")

ggResidpanel::resid_panel(model2)


```
### Task: Write the report
The relationship between the predictors and the outcome appeared broadly linear, with only minor variation in the spread of residuals across predicted values. These small fluctuations suggest that the assumptions of linearity and approximately equal variance were largely satisfied.

The distribution of residuals was assessed using both statistical and descriptive approaches. The Shapiro–Wilk test indicated no significant deviation from normality (p = 0.389), and the residuals were approximately symmetric and centred around zero, supporting the assumption of normality.

No standardized residuals exceeded ±3.29, and no cases were identified as influential based on the Bonferroni-adjusted outlier test or Cook’s distance statistics. This suggests that the regression estimates were not disproportionately affected by any individual observation.

Variance Inflation Factors (VIFs) ranged from 1.82 to 7.44, and tolerance values remained above 0.1. These results indicate moderate but acceptable intercorrelations among predictors, with no evidence of severe multicollinearity.

Overall, the diagnostic checks indicated that Model 2 met the key assumptions of multiple linear regression. The relationships among predictors and the outcome were generally linear, residuals were approximately normally distributed, and predictor independence was adequate. The model can therefore be regarded as appropriate for examining the combined influence of stress, sleep duration, and gender on daytime fatigue.

------------------------------------------------------------------------

# Task: Reflection

1.  Compare Model 1 and Model 2. Does adding gender improve the model?\
2.  Which assumptions are most/least well met in each model?\
3.  Suggest one possible next step if violations were detected (e.g., transformation, re-specification, removal of outliers).


---


**End of Exercise**
