---
title: "MATH9102-W10-DimRed-Exercise"
format:
    html:
      self-contained: true
      toc: true      # Enables table of contents
      toc_float: true # Makes the TOC float on the side
      toc_depth: 3   # Specifies the depth of headers to include in TOC
---



# Introduction
In this lab you will be conducting a Principal Component Analysis (PCA) and an Exploratory Factor Analysis (EFA). 
You will use a wine dataset. It contains the results of a chemical analysis of wines grown in the same region in Italy but derived from three different cultivars. The analysis determined the quantities of 13 constituents found in each of the three types of wines. More information is available at https://search.r-project.org/CRAN/refmans/HDclassif/html/wine.html.


| **Variable Name**        | **Description**                                                                                                | **Type**             | **Higher Values Indicate…**                 |
| ------------------------ | -------------------------------------------------------------------------------------------------------------- | -------------------- | ------------------------------------------- |
| **Wine**                 | Wine class/cultivar identifier (1, 2, or 3), representing three grape varieties from the same region in Italy. | Categorical (factor) | Not applicable (group label)                |
| **Alcohol**              | Alcohol content (% by volume) of the wine sample.                                                              | Continuous           | Higher alcohol concentration                |
| **Malic.acid**           | Amount of malic acid, contributing to tartness and acidity.                                                    | Continuous           | More malic acid (greater sour/tart profile) |
| **Ash**                  | Total ash content (mineral residue after combustion).                                                          | Continuous           | Higher mineral content                      |
| **Acl**                  | Alkalinity of ash (buffering capacity/basicity measure).                                                       | Continuous           | Greater alkalinity level                    |
| **Mg**                   | Magnesium concentration (mg/L).                                                                                | Continuous           | Higher magnesium content                    |
| **Phenols**              | Total phenolic compounds affecting flavour and antioxidant properties.                                         | Continuous           | Higher total phenolic content               |
| **Flavanoids**           | Subclass of phenols contributing to flavour, colour, and antioxidant profile.                                  | Continuous           | Higher flavonoid concentration              |
| **Nonflavanoid.phenols** | Non-flavonoid phenolic compounds contributing to colour and astringency.                                       | Continuous           | Higher non-flavonoid phenol content         |
| **Proanth**              | Proanthocyanins (tannins) related to colour stability and astringency.                                         | Continuous           | Higher tannin content                       |
| **Color.int**            | Intensity of wine colour, measured spectrophotometrically.                                                     | Continuous           | Darker, more intense wine colouring         |
| **Hue**                  | Shade/tone of wine colour (ratio measure).                                                                     | Continuous           | Shift toward lighter/yellower hue           |
| **OD**                   | Optical density (OD280/OD315) indicating total polyphenol concentration.                                       | Continuous           | Higher polyphenol content                   |
| **Proline**              | Proline concentration (amino acid), often linked to ripeness and wine quality.                                 | Continuous           | Higher proline levels                       |


```{r echo=FALSE, include=FALSE}
# Set CRAN mirror for non-interactive sessions
options(repos = c(CRAN = "https://cloud.r-project.org"))

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}
# tidyverse: For data manipulation (dplyr), wrangling (tidyr), and plotting (ggplot2),
# essential for preparing datasets before running PCA/EFA.
quiet_library("tidyverse")

# psych: Core package for PCA, EFA, reliability analysis (Cronbach's alpha),
# factor extraction, rotation, and interpreting loadings.
quiet_library("psych")

# REdaS: Provides KMO (Kaiser-Meyer-Olkin) and Bartlett's test functions
# used to assess whether the dataset is suitable for dimension reduction.
quiet_library("REdaS")

# Hmisc: Contains correlation tools and functions for handling missing data,
# useful for computing correlation matrices required for PCA/EFA.
quiet_library("Hmisc")

# corrplot: Produces visual correlation matrix heatmaps,
# helping identify clusters of variables and collinearity patterns.
quiet_library("corrplot")

# ggcorrplot: ggplot-based correlation matrix visualisation,
# offering cleaner, publication-ready heatmaps.
quiet_library("ggcorrplot")

# factoextra: Visualisation tools for PCA and EFA outputs,
# including scree plots, biplots, eigenvalue charts, and clustering plots.
quiet_library("factoextra")

# nFactors: Helps determine the number of factors/components to retain
# using methods like parallel analysis and eigenvalue comparisons.
quiet_library("nFactors")

# knitr: Enables clean rendering of tables, outputs, and formatted results
# inside R Markdown or Quarto reports.
quiet_library("knitr")

quiet_library("readr")
```


```{r echo=FALSE, include=FALSE}
#| label: loadwine
datalocation <- file.path("./")
# Load the winedataset (update the path as needed)
datapath <- file.path(datalocation,'wine.csv')
wine <- read.csv(datapath)

# Remove the variables that won't be used in the dimension reduction
wine <- wine %>%
  dplyr::select(-Wine)
```





```{r}
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# Function to make a correlation matrix more readable
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
# Source: http://www.sthda.com/
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# Function to make a correlation matrix more readable
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
# Source: http://www.sthda.com/
flattenCorrMatrix <- function(cormat, pmat, cutoff = 0.05) {
  
  ut <- upper.tri(cormat)
  
  df <- data.frame(
    row    = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor    = cormat[ut],
    p      = pmat[ut],
    stringsAsFactors = FALSE
  )
  
  # Round correlation + p to 3 digits
  df$cor <- round(df$cor, 3)
  df$p   <- round(df$p, 3)
  
  # Format p for printing (non-bold)
  df$p_text <- ifelse(df$p < 1e-4,
                      format(df$p, scientific = FALSE, digits = 3),
                      sprintf("%.4f", df$p))
  
  # Markdown bold formatting for Quarto / Rmd tables
  df$p_md <- ifelse(df$p < cutoff,
                    paste0("**", df$p_text, "**"),
                    df$p_text)
  
  return(df)
}

```

# TASK ONE - Principal Component Analysis - Wine Quality dataset

Use PCA where you are certain there are underlying factors.
You are using the approach to reduce the number of variables you need to use to represent concepts.


## STEP 1: Create and screen the correlation matrix for the wine dataset
### Step 1a: Create the matrix
```{r corscreen}

#create a correlation matrix  - choose one of the approaches given in MATH9102-W10-DimRed.qmd
wineMatrix<-cor(wine)
round(wineMatrix, 2)
```




### Step 1b: Visualize the correlation matrix

```{r }
#| fig.cap:  "Figure 1: Basic Correlation Matrix"

# 1. Basic correlation matrix plot
ggcorrplot::ggcorrplot(wineMatrix)
```

```{r}
#| fig.cap: "Figure 1: Correlation Matrix - Wine Dataset"
# 5. Correlation matrix showing coefficients with labels
# Note: This can become cluttered with a high number of variables
#| fig-width: 16
#| fig-height: 16

ggcorrplot::ggcorrplot(
  wineMatrix,
  lab = TRUE,
  lab_size = 2.5,
  tl.cex = 8,
  tl.srt = 45,
  type = "lower"
)

```

## STEP 2: Check if the data is suitable for dimension reduction (there is enough variance of the right kind)
You want to do the following:

  * Check that your variables are correlated by doing Bartlett's test of Sphericity.
    + Check that the result is statistically significant.
  
  * Check that variables are not too highly correlated for PCA.
    + Calculate and inspect the determinant.
    + Check to see if it is greater than 0.00001.
  
  * Check that there is enough variance in your variables that could be caused by an underlying factor/dimension.
    + Use the Kaiser-Meyer-Olkin Measure of Sampling Adequacy (MSA) test.
    + High values close to 1 suggest that PCA/FA might be useful.
    + Values over 0.8 or over are considered strong.
    + Anything less than 0.5 suggests that PCA/FA won’t be useful.
    + 0.00 to 0.49 unacceptable, 0.50 to 0.59 miserable, 0.60 to 0.69 mediocre ,0.70 to 0.79 middling, 0.80 to 0.89 meritorious, 0.90 to 1.00 marvellous.

### Step 2a: Check Bartlett's test
```{r}
psych::cortest.bartlett(wine)
```


### Step 2b: Check the (KMO)

```{r}
#KMO 
REdaS::KMOS(wine)
```


### Step 2c:  Check the Determinant

```{r}
#Determinant (execute one of these):
det(wineMatrix)

```

##  Report
Using the guidance of Field (2024), the chemical composition data from the wine dataset was assessed for its suitability for dimension reduction. Bartlett’s test of sphericity produced a significant result, χ²(78) = 1317.181, p < .001, indicating that the observed correlation matrix contains sufficient shared variance for factor extraction or principal components analysis. The Kaiser–Meyer–Olkin (KMO) Measure of Sampling Adequacy was 0.78, which falls within the “middling” range according to Kaiser and Rice’s (1974) criteria, suggesting that the dataset is adequately suited for dimension reduction techniques. Examination of the determinant of the correlation matrix yielded a value of 0.0005, which is above the recommended minimum threshold of 0.00001, indicating that the data is not affected by problematic multicollinearity or singularity. Collectively, these diagnostics demonstrate that the wine dataset is appropriate for principal component analysis or exploratory factor analysis.


# TASK TWO Conduct PCA - Wine Dataset

Form: pcModel<-principal(dataframe/R-matrix, nfactors = number of factors, rotate = "method of rotation", scores = TRUE)

For PCA we know how many components it is possible to find.

The principal function will work out the loadings of each variable onto each component, i.e. the proportion each component explained and the cumulative proportion of variance explained.

## STEP 3: Execute the PCA
In this step, we run the PCA using all of the remaining variables in the wine dataset. The purpose here is simply to let R extract all possible components so we can see the complete structure of the data. You should focus on the loading matrix that is produced. 
A loading tells you how strongly each variable relates to a component. For example, you will notice that variables such as Flavanoids, Phenols, OD, and Proanth load very highly on the first component. This shows that these variables share a lot of variance and are likely measuring a similar underlying idea (in this case, something like “phenolic content”). 
Other components will have smaller or weaker loadings, and that’s expected — because PCA is extracting every possible component, not just the meaningful ones. At this stage we are not trying to interpret all components; instead, we are trying to understand the overall structure so that we can make decisions in the next step.


```{r}
#do the reduction
pc1 <-  principal(wine, nfactors = length(wine), rotate = "none")#Conduct the PCA extracting the same number of components as there are variables

pc1#output all details of the PCA

```

The initial unrotated PCA was performed using all possible components, extracting 13 components — one for each original variable. 

The unrotated loading matrix showed that several variables loaded strongly on the first component, including Flavanoids (.92), Phenols (.86), OD (.82) and Proanth (.68). These variables represent related aspects of the wines’ phenolic profile. Other components showed weaker and more diffuse patterns, reflecting the fact that PCA extracts all variance, including variance that is not meaningful.

The eigenvalues indicated that the first three components explained the majority of total variance:

PC1: 4.71 (36.2%)

PC2: 2.50 (19.2%)

PC3: 1.45 (11.1%)

Together, the first three components accounted for 66.5% of total variance, with subsequent components contributing very little (all < 0.92 eigenvalue; < 7% variance each).


## STEP 4:  Decide on which components to retain
Once the full PCA is run, we need to decide how many components to keep for interpretation. There are two main things to look at:

The scree plot – This helps you visually identify where the “elbow” or point of inflexion is. After this point, the components explain very little additional variance.

Eigenvalues – A common guideline is to keep components with eigenvalues greater than 1 (Kaiser’s criterion) or greater than 0.7 (Joliffe's criterion).



```{r}
#Create the scree plot
plot(pc1$values, type = "b", main="Scree Plot of Statistics Survey Questions - PCA") 

#Print the variance explained by each component
pc1$Vaccounted 

#Print the Eigenvalues
pc1$values


#Another way to look at the eigenvalues plus variance explained (need to use princomp function of PCA to get right class for use with factoextra functions)
pcf=princomp(wine)
factoextra::get_eigenvalue(pcf)
factoextra::fviz_eig(pcf, addlabels = TRUE, ylim = c(0, 50))#Visualize the Eigenvalues
factoextra::fviz_pca_var(pcf, col.var = "black")
factoextra::fviz_pca_var(pcf, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE # Avoid text overlapping
             )

#Print the loadings above the level of 0.3
psych::print.psych(pc1, cut = 0.3, sort = TRUE)

#Create a diagram showing the components and how the manifest variables load
fa.diagram(pc1) 

#Show the loadings of variables on to components
fa.sort(pc1$loading)
```


Communalities - tell us how much of each variable’s variance is captured by the components we keep.
A communality close to 1 means the retained components represent that variable very well.
A communality below about 0.4 may indicate that the variable is not well represented.

In the initial (unrestricted) PCA, each communality is exactly 1 because PCA uses all available variance. But once we extract only four components and apply rotation (in STEP 5), the communalities will drop to realistic values.

For this dataset, you will see that the first four components meet this threshold, and the scree plot shows a clear drop after the fourth component. Together, these components explain over 75% of the total variance, which means they capture most of the meaningful information in the dataset. Everything after Component 4 explains very little and mostly reflects noise. Therefore, in the next step, we extract **three** components and work with those.

```{r}
#Output the communalities of variables across components (will be one for PCA since all the variance is used)
pc1$communality 
#Visualize contribution of variables to each component
var <- factoextra::get_pca_var(pcf)
corrplot::corrplot(var$contrib, is.corr=FALSE) 

# Contributions of variables to PC1
factoextra::fviz_contrib(pcf, choice = "var", axes = 1, top = 10)
# Contributions of variables to PC2
factoextra::fviz_contrib(pcf, choice = "var", axes = 2, top = 10)
# Contributions of variables to PC3
factoextra::fviz_contrib(pcf, choice = "var", axes = 3, top = 10)
```

Two criteria were used:

1. Scree plot

The scree plot showed a clear “elbow” at Component 3, where the slope sharply flattened, indicating that meaningful reductions end at three components. Components 4–13 contributed trivial additional variance.

2. Eigenvalues

Kaiser’s criterion (eigenvalue > 1) supported retaining three components (4.71, 2.50, 1.45).
Joliffe’s criterion (eigenvalue > 0.7) would retain a fourth component (0.92), but inspection revealed that the fourth component was weak and conceptually incoherent.

Decision: A three-component solution was chosen as the most parsimonious and theoretically interpretable representation of the data.

## STEP 5: Repeat with reduced components and rotation

After deciding how many components to retain, we re-run the PCA asking for only those components (in this case, three). We then apply a rotation — usually varimax when doing PCA. Rotation does not change the amount of variance explained, but it helps make the components much easier to interpret. Rotation works by trying to maximise high loadings and minimise low ones so that each variable loads strongly on only one component whenever possible.

Once you apply rotation, you will find that the structure becomes much clearer. For example:

One component is mainly defined by phenolic variables (Flavanoids, Phenols, OD, Proanth).

Another is defined by Alcohol, Proline, and Colour Intensity.

A third captures variables related to acidity (Malic.acid, Acl, Hue).


The rotated solution gives you a clearer picture of the underlying dimensions in the dataset, and this is the version you would normally interpret.

```{r}
#Apply rotation to try to refine the component structure
pc2 <-  principal(wine, nfactors = 3, rotate = "varimax")#Extracting 3 components
#output the components
psych::print.psych(pc2, cut = 0.3, sort = TRUE)
#output the communalities
pc2$communality
#NOTE: you can do all the other things done for the model in STEP 4 if you want to
pc2$Vaccounted
```

Loadings after Rotation:

| Variable             | RC1 (Phenolics) | RC2 (Alcohol–Ripeness) | RC3 (Acidity–Colour Tone) |
| -------------------- | --------------- | ---------------------- | ------------------------- |
| Flavanoids           | .90             | —                      | —                         |
| OD                   | .88             | —                      | —                         |
| Phenols              | .82             | —                      | —                         |
| Proanth              | .66             | —                      | —                         |
| Nonflavanoid.phenols | −.56            | —                      | .33                       |
| Hue                  | .74             | —                      | —                         |
| Malic.acid           | −.56            | —                      | —                         |
| Alcohol              | —               | .86                    | —                         |
| Proline              | —               | .76                    | —                         |
| Color.int            | −.44            | .75                    | —                         |
| Mg                   | —               | .51                    | —                         |
| Ash                  | —               | —                      | .84                       |
| Acl                  | —               | −.32                   | .79                       |

Variance retained

Component 1: 33.4%

Component 2: 20.5%

Component 3: 12.6%

Total variance explained: 66.5%

Communalities (selected examples)

Flavanoids: .87

Phenols: .77

OD: .77

Proanth: .50

Alcohol: .74

Proline: .74

Color.int: .77

Hue: .62

Ash: .82

Acl: .81

Most variables were well represented (h² > .40), supporting adequacy of the solution.



## Report the outcomes - replace ? by values you obtain/your conclusions:

A principal component analysis (PCA) with varimax rotation was  conducted on the wine dataset to explore the underlying structure of its chemical composition variables. An initial unrotated PCA indicated that the first three components had eigenvalues exceeding Kaiser’s criterion and together explained 66.5% of the total variance. The scree plot showed a clear point of inflexion after the third component, supporting the retention of three components as the most interpretable and parsimonious solution.

The rotated PCA revealed a clear and coherent structure:

Component 1: Phenolic–Tannin Composition
This component showed strong positive loadings from Flavanoids (.90), OD (.88), Phenols (.82), Proanth (.66), and a moderate negative loading from Nonflavanoid.phenols (−.56). These variables represent polyphenolic concentration, colour density, antioxidant activity, and tannin content. High communalities (e.g., Flavanoids h² = .87; OD h² = .77) indicate that this component captures a substantial proportion of variance in these variables.

Component 2: Alcohol–Ripeness
Alcohol (.86), Proline (.76), and Colour Intensity (.75) loaded strongly onto this component, with a smaller positive loading for Magnesium (.51). These variables are closely linked to grape ripeness, fermentation, and concentration effects. Communalities for these indicators were high (e.g., Alcohol h² = .74; Proline h² = .74), supporting the component’s coherence.

Component 3: Acidity–Colour Tone
This component was defined by strong loadings from Ash (.84), Acl (.79), and Hue (.74), capturing the wines’ acidity, buffering capacity, and colour tone characteristics. Communalities for these variables ranged from .62 to .82, indicating good representation.

The components reflect three key chemical dimensions of the wines: phenolic richness, alcohol–ripeness characteristics, and acidity-related colour attributes. 



```{r}
#| fig.cap: "Figure 2 : Scree Plot Unrotated Solution"
#Create the scree plot
plot(pc1$values, type = "b", main="Scree Plot of Statistics Survey Questions - PCA") 
```

```{r}
#| fig.cap: "Figure 3 : Loading of Manifest Variables onto Factors"
#create a diagram showing the components and how the manifest variables load
fa.diagram(pc2) 
```


# TASK THREE: Conduct the Exploratory Factor Analysis

We now move from PCA (a data-reduction technique) to Exploratory Factor Analysis (EFA), which attempts to uncover *latent factors* that may be causing the observed correlations.
Here, we run an EFA using maximum likelihood extraction.
This method assumes multivariate normality and allows us to evaluate model fit.
Rotation (varimax) is applied AFTER deciding the number of factors to make interpretation clearer.


## STEP 3: Execute the EFA with rotation

Before choosing how many factors to keep, we perform a parallel analysis.
This compares the size of our eigenvalues to those expected by chance.
The number of factors where our data’s eigenvalues exceed random-data eigenvalues is recommended as the optimal number of factors.

We run an initial EFA extracting the maximum possible number of factors.
This unrotated solution is NOT used for interpretation.
Instead, we use it to inspect the eigenvalues, communalities, and loading structure before applying rotation and choosing a final factor model.

```{r}
#Factor Analysis - the default here is principal axis factoring fm=pa
#If we know our data going in is normally distributed we use maximum likelihood
# Determine the optimal number of factors using parallel analysis
psych::fa.parallel(wineMatrix)
# Do the factor reduction
facsol <- psych::fa(wineMatrix, nfactors=length(wine), obs=NA, n.iter=1, rotate="none", fm="ml")
```

**Understanding the initial unrotated maximum-likelihood EFA** 

The first three eigenvalues from the ML solution were:

ML1: 4.41

ML2: 2.16

ML3: 1.09

These correspond to 33.3%, 16.0%, and 8.0% of variance respectively. The cumulative variance explained by the first three unrotated factors was 57.3%.

The remaining eigenvalues were all below 1 and contributed very little, supporting the likelihood of a three-factor solution.

## STEP 4: Decide which factors/components to retain (FACTOR ANALYSIS)
### STEP 4a: Look at the loadings

Generate a scree plot to see where the “elbow” occurs.
This helps determine how many factors explain meaningful variance.
Higher eigenvalues indicate stronger factors.

The factor diagram visualises how variables relate to each factor.
Thicker lines = stronger loadings. Red indicates negative loading
This helps you see which variables group together naturally.

```{r}
#Create your scree plot
plot(facsol$values, type = "b") #scree plot

#Print the Variance accounted for by each factor/component
facsol$Vaccounted
#Output the Eigenvalues
facsol$values 

#Print the components with loadings
psych::print.psych(facsol,cut=0.3, sort=TRUE)

#Print sorted list of loadings
fa.sort(facsol$loading)

#create a diagram showing the factors and how the manifest variables load
fa.diagram(facsol)

```

### STEP 4b: Look at the communalities
```{r}
#Output the communalities of variables across components 
facsol$communality 

# Contributions of variables to factor solution
factoextra::fviz_contrib(pcf, choice = "var", axes = 1, top = 10)
# Contributions of variables to factor solution
factoextra::fviz_contrib(pcf, choice = "var", axes = 2, top = 10)
```

1. Parallel analysis indicated:
  
  Factors to retain (FA eigenvalues): 3
  
  Components to retain (PCA eigenvalues): 2

  This is a common outcome in real data: the FA solution  identifies more latent factors than PCA components.

2. Kaiser’s criterion (eigenvalues > 1)

  Suggested up to three factors, the fourth factor had:
    
    A small eigenvalue (0.52)
    
    A loading structure dominated by a single variable
    
    This indicated poor interpretability.

3. Scree plot showed:

  A steep decline after Factor 2
  
  A small but interpretable inflexion at Factor 3
  
  A complete flattening from Factor 4 onwards

Decision:

  A three-factor solution was selected because:
  
    It aligns with parallel analysis
    
    It avoids weak “single-variable factors”
    
    It provides a conceptually coherent structure


## STEP 5: Apply rotation for EFA

Parallel analysis compares:

Actual Data (solid blue lines)
– PC Actual Data: eigenvalues from PCA
– FA Actual Data: eigenvalues from Factor Analysis

Simulated Data (dotted red lines)
– these simulate what eigenvalues would look like if the data contained NO underlying structure (i.e., random noise)

A component/factor should be retained only if the actual eigenvalue is larger than the simulated (random) eigenvalue at the same component number.

In this plot:
The blue triangles exceed the red line for only the first two factors.

The line drops below the simulated values at factor 3 and onwards.

After deciding how many factors to retain, we rerun the EFA using varimax rotation.
Rotation does NOT change total variance explained—it only makes patterns clearer.
Varimax rotation tries to ensure each variable loads strongly onto ONE factor, simplifying interpretation of the factor structure.

```{r}
#Apply rotation to try to refine the component structure
# Determine the optimal number of factors using parallel analysis
psych::fa.parallel(wineMatrix)

# Use the determined nfactors value (e.g., 3) in the factor analysis
facsolrot <- psych::fa(wineMatrix, nfactors=3, obs=NA, n.iter=1, rotate="varimax", fm="ml")

#output the components
psych::print.psych(facsolrot, cut = 0.3, sort = TRUE)
#output the communalities
facsolrot$communality

```
Loadings after rotation:

| Variable             | RC1 (Phenolics) | RC2 (Alcohol–Ripeness) | RC3 (Acidity–Colour Tone) |
| -------------------- | --------------- | ---------------------- | ------------------------- |
| Flavanoids           | .90             | —                      | —                         |
| OD                   | .88             | —                      | —                         |
| Phenols              | .82             | —                      | —                         |
| Proanth              | .66             | —                      | —                         |
| Nonflavanoid.phenols | −.56            | —                      | .33                       |
| Hue                  | .74             | —                      | —                         |
| Malic.acid           | −.56            | —                      | —                         |
| Alcohol              | —               | .86                    | —                         |
| Proline              | —               | .76                    | —                         |
| Color.int            | −.44            | .75                    | —                         |
| Mg                   | —               | .51                    | —                         |
| Ash                  | —               | —                      | .84                       |
| Acl                  | —               | −.32                   | .79                       |

Variance retained

Component 1: 33.4%

Component 2: 20.5%

Component 3: 12.6%

Total variance explained: 66.5%

Communalities (selected examples)

Flavanoids: .87

Phenols: .77

OD: .77

Proanth: .50

Alcohol: .74

Proline: .74

Color.int: .77

Hue: .62

Ash: .82

Acl: .81

Most variables were well represented (h² > .40), supporting adequacy of the solution.

## STEP 6: Report EFA:
Report the findings:

An exploratory factor analysis (EFA) using maximum-likelihood extraction and varimax rotation was conducted to identify latent structures underlying the chemical composition of the wines. Parallel analysis supported the retention of three factors, and inspection of the scree plot and eigenvalue patterns confirmed this solution as the most interpretable and theoretically coherent.

The rotated solution revealed three distinct and meaningful factors:

Factor 1: Phenolic–Tannin Structure

This factor exhibited strong loadings from Flavanoids (.93), OD (.86), Phenols (.82), and Proanth (.62), along with a moderate negative loading from Nonflavanoid.phenols (–.53). These variables collectively capture phenolic concentration, antioxidant properties, colour density, and tannin structure. Communalities for these variables were high (e.g., Flavanoids h² = .93; Phenols h² = .80), indicating that this factor accounts for a substantial proportion of their variance. This dominant factor reflects the wines’ polyphenolic richness.

Factor 2: Alcohol–Ripeness

Factor 2 showed strong loadings on Alcohol (.78), Proline (.69), and Color.int (.75), with a smaller contribution from Mg (.37). These variables index characteristics related to grape ripeness, fermentation processes, and extraction, positioning this factor as representing alcohol content and ripeness-related chemical properties. Communalities were generally high for the core variables (e.g., Proline h² ≈ .62; Color.int h² ≈ .75).

Factor 3: Acidity–Colour Tone

This factor was defined by high loadings from Acl (.86) and Ash (.63), along with a substantial loading from Malic.acid (–.47) and Hue (.65). Together these represent measures of acidity, buffering capacity, mineral content, and spectral colour tone. Communalities ranged from moderate to high (Hue h² ≈ .50; Acl h² ≈ .93), supporting the interpretation of a coherent acidity–colour dimension.

Overall structure

The three factors together explained approximately 58% of the total variance, a substantial proportion for a chemical composition dataset of this complexity. Each factor aligned closely with known principles of wine chemistry:

- Polyphenolic structure

- Alcohol and ripeness indicators

- Acidity and colour balance

Only one variable (Mg) showed a low communality (h² = .16), suggesting it is not strongly governed by the latent factors identified.

```{r}
#| fig.cap: "Figure 2 : Parallel Analysis"
# Determine the optimal number of factors using parallel analysis
psych::fa.parallel(wineMatrix)
```

```{r}
#| fig.cap: "Figure 3 : Loading of Manifest Variables onto Factors"
#create a diagram showing the components and how the manifest variables load
fa.diagram(facsolrot) 
```

# TASK FOUR:  Reliability Analysis

Now that we have identified components (from PCA) and factors (from EFA), we need to check whether the variables within each proposed group consistently measure the same underlying construct.

Cronbach's alpha is used to evaluate internal consistency:
   α > .80 = good
   α ~ .70 = acceptable
   α < .60 = poor

 Note: Dimension reduction does NOT guarantee reliability.
 A component/factor may be statistically valid but not reliable as a scale.
 If we are just trying to reduce the number of variables we may still be able to use. If however we want to use the component/factor as a reliable measure of a concept it will cause use problems if it is not considered reliable.


## STEP 6: Reliability Analysis PCA 
```{r}
#If you know that certain sections should be treated as groups, test these separately
#Output Cronbach alpha values - use pysch::alpha

# Component 1: Phenolic–Tannin Composition
comp1_vars <- wine[, c("Flavanoids", "Phenols", "OD", "Proanth", "Nonflavanoid.phenols")]

# Component 2: Alcohol–Ripeness
comp2_vars <- wine[, c("Alcohol", "Proline", "Color.int")]

# Component 3: Acidity–Colour Tone
comp3_vars <- wine[, c("Malic.acid", "Acl", "Hue")]

# Compute Cronbach's Alpha for each PCA-derived component
psych::alpha(comp1_vars)   # Phenolic–Tannin Composition
psych::alpha(comp2_vars)   # Alcohol–Ripeness
psych::alpha(comp3_vars)   # Acidity–Colour Tone


```
Internal consistency was evaluated for each of the three PCA-derived components using Cronbach’s alpha. Reliability was assessed to determine whether the variables loading on each component form a coherent measurement scale rather than merely a statistical grouping. This distinction is important because PCA identifies patterns of shared variance but does not guarantee that the resulting components represent reliable constructs.

Component 1: Phenolic–Tannin Composition included Flavanoids, Phenols, OD, Proanth, and Nonflavanoid.phenols. Cronbach’s alpha for this set of variables was: α = .81. This indicates acceptable internal consistency. Inspection of the item-deleted statistics showed: The four positive-loading items (Flavanoids, Phenols, OD, Proanth) correlated strongly with the underlying component. Nonflavanoid.phenols showed a negative loading, which initially reduced internal consistency.

Component 2: Alcohol–Ripeness consisted of Alcohol, Proline, and Color.int. The reliability was: α ≈ .01. This value indicates extremely poor internal consistency. Although these variables loaded together in the PCA due to statistical covariance, the low alpha suggests that these variables do not behave consistently enough to form a unified scale. Even though the component is meaningful in explaining shared variance, it should not be interpreted as a stable latent construct.

Component 3: Acidity–Colour Tone contained Malic.acid, Acl, and Hue. This group also showed low reliability: α = .16.  While these variables load together in the PCA to form a meaningful statistical dimension, they do not form a reliable measurement scale.

In summary, only Component 1 shows acceptable internal coherence and could be used as a reliable composite scale.
Components 2 and 3, although statistically interpretable and useful for dimension reduction, do not possess the internal consistency required for use as a valid, reliable measurement scale.  

## STEP 6: Reliability Analysis EFA
This is essentially the same but demonstrates the use of check.keys
```{r}
# -------------------------------------------------
# Factor 1: Phenolic–Tannin Composition
# Variables with strong F1 loadings:
# Flavanoids, Phenols, OD, Proanth, Nonflavanoid.phenols
# -------------------------------------------------
efa_comp1 <- wine[, c("Flavanoids", "Phenols", "OD", "Proanth", "Nonflavanoid.phenols")]

psych::alpha(efa_comp1)                # Raw alpha
psych::alpha(efa_comp1, check.keys=TRUE)  # Auto-reverse if needed


# -------------------------------------------------
# Factor 2: Alcohol–Ripeness
# Variables with strong F2 loadings:
# Alcohol, Proline, Color.int, Mg
# -------------------------------------------------
efa_comp2 <- wine[, c("Alcohol", "Proline", "Color.int", "Mg")]

psych::alpha(efa_comp2)
psych::alpha(efa_comp2, check.keys=TRUE)

# -------------------------------------------------
# Factor 3: Acidity–Colour Tone
# Strong F3 variables:
# Malic.acid, Acl, Hue
# -------------------------------------------------
efa_comp3 <- wine[, c("Malic.acid", "Acl", "Hue")]

psych::alpha(efa_comp3)
psych::alpha(efa_comp3, check.keys=TRUE)

```
Cronbach’s alpha was computed for each of the three factors identified in the maximum-likelihood EFA with varimax rotation. Although factor analysis identifies clusters of variables that share common variance, this does not guarantee that these variables form reliable measurement scales. Reliability analysis therefore evaluates whether each factor represents a coherent and internally consistent latent construct.

Factor 1: Phenolic–Tannin Structure,  consisted of Flavanoids, Phenols, OD, Proanth, and Nonflavanoid.phenols. The initial reliability analysis indicated the presence of a negatively oriented item (Nonflavanoid.phenols), which is expected because it loads inversely on the factor. After applying automatic reverse-keying, Cronbach’s alpha was: α = .85. This indicates good internal consistency. All five variables contributed positively to the scale once reverse-keying was applied, with item-deleted statistics showing that no variable substantially reduced reliability. These results indicate that Factor 1 reflects a coherent and internally reliable polyphenolic–tannin construct, consistent with the strong and interpretable loading pattern identified in the EFA.

Factor 2: Alcohol–Ripeness included Alcohol, Proline, Color.int, and Mg. Cronbach’s alpha for these four variables was: α = .06. This indicates poor internal consistency.  Thee low reliability suggests that although the variables loading onto this factor share variance statistically, they do not behave as a unified measurement scale. Thus, Factor 2 should be interpreted as a meaningful statistical dimension but not as a coherent latent construct.

Factor 3: Acidity–Colour Tone consisted of Malic.acid, Acl, and Hue (with Ash optionally cross-loading onto this factor). The reliability estimate for the three-variable grouping was: α = .28 (after reverse-keying Hue). This reflects questionable internal consistency. Although the variables load together in the EFA and collectively represent aspects of acidity, buffering capacity, and colour tone, their inter-item correlations are inconsistent. The factor therefore represents a valid statistical dimension but does not meet the thresholds required to be considered a reliable latent construct.



# Summary comparison
The results highlight that although PCA and EFA yield similar structural patterns, EFA indicates weaker reliability for two of the three components, emphasising the need for cautious interpretation.

