---
title: "MATH9102-W2-Soln"
author: "Deirdre Lawless"
format: 
  html:
    self-contained: true   # Embed images rather than create sub-folder
    code-fold: true         # Fold code blocks, user can click to expand
    code-tools: true        # Show copy/download buttons for code
editor: visual
---

#TU059/TU060/TU256/PhD School of Computer Science, TU Dublin
#MATH9102 - Week Two Solutions

## 1. Install and load packages and Load Data
```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}

quiet_library("tidyverse")
#Amend this to reflect your folder structure 
datalocation <- file.path("C:","Users", "deirdre.lawless", "datafiles")
datapath <- file.path(datalocation,'experim.dat')

#Read in the dataset
exper <- read.table(datapath, header=TRUE)
names(exper)

```



# HISTOGRAMS Fear of Statistics (fost1, fost2)

```{r}
ggplot(exper, aes(fost1)) +
  geom_histogram(binwidth = 0.4, fill="red") +
  labs(x = "Fear of Statistics (Time 1)", y = "Frequency")


ggplot(exper, aes(fost2)) +
  geom_histogram(binwidth = 0.4, fill="red") +
  labs(x = "Fear of Statistics (Time 2)", y = "Frequency")


```

# BAR CHARTS (mean scores Fear of Statistics by gender)

```{r}
ggplot(exper, aes(sex, fost1)) +
  stat_summary(fun = mean, geom="bar") +
  labs(x = "Gender", y = "Mean Fear of Statistics (Time 1)")

    
ggplot(exper, aes(sex, fost2)) +
  stat_summary(fun = mean, geom="bar") +
  labs(x = "Gender", y = "Mean Fear of Statistics (Time 2)")
```


# STEM AND LEAF Confidence in Coping with Statistics

```{r}
stem(exper$confid1)
stem(exper$confid2)
```




# BOXPLOTS Depression

## by time only
```{r}
ggplot(exper, aes(y = depress1)) +
  geom_boxplot() + labs(y = "Depression (Time 1)")


ggplot(exper, aes(y = depress2)) +
  geom_boxplot() + labs(y = "Depression (Time 2)")
```

## by gender
```{r}

ggplot(exper, aes(sex, depress1)) +
  geom_boxplot() + labs(x = "Gender", y = "Depression (Time 1)")


ggplot(exper, aes(sex, depress2)) +
  geom_boxplot() + labs(x = "Gender", y = "Depression (Time 2)")
```




# DENSITY PLOTS Fear of Statistics

## overall
```{r}
ggplot(exper, aes(fost1)) +
  geom_density(fill="blue", alpha=0.4) +
  labs(x = "Fear of Statistics (Time 1)", y = "Density")

  
ggplot(exper, aes(fost2)) +
  geom_density(fill="blue", alpha=0.4) +
  labs(x = "Fear of Statistics (Time 2)", y = "Density")
```

## by gender
```{r}
ggplot(exper, aes(fost1, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Fear of Statistics (Time 1)", y = "Density")


ggplot(exper, aes(fost2, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(x = "Fear of Statistics (Time 2)", y = "Density")

```




# EXAM RESULTS

```{r}
summary(exper$exam)
hist(exper$exam, main="Distribution of Exam Results",
     xlab="Exam Score", col="lightgreen", border="white")
```

