---
title: "Title"
subtitle: "Include module code"
author: "Include your student ID and name"
format:
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    toc: true
editor: source
---


```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}

quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("pastecs")    # Tools for descriptive statistics, time series analysis, and data exploration (e.g., stat.desc)
quiet_library("psych")   # Descriptive statistics by group
quiet_library("FSA")        # Includes general tools for data analysis (e.g., nonparametric tests, bootstrapping)
quiet_library("semTools")   # Supplementary functions for Structural Equation Modeling (SEM), including model fit, reliability, and measurement invariance tests
quiet_library("VIM") # Visualization and Imputation of Missing Values
quiet_library("naniar") #(naniar short for NA + naniar) - Missing data wrangling plus Little's test
quiet_library("simputation") # For imputing variables

quiet_library("sjstats")  # Provides convenient functions for effect sizes
quiet_library("gmodels")  # Includes 'CrossTable()' for contingency tables.

quiet_library("car")   # Includes Levene's test for homogeneity
quiet_library("effectsize")   # Includes calculations for effectsize
quiet_library("rstatix")   # Includes calculations for effectsize and Dunn's test

quiet_library("gt") # Formatting tables
quiet_library("flextable") # Formatting tables

```


```{r}
#| label: loaddata
#Amend this to reflect your folder structure 
datalocation <- file.path("C:","Users", "deirdre.lawless", "datafiles")
datapath <- file.path(datalocation,'experim_clean.dat')

#Read in the dataset
experim <- read.table(datapath, header=TRUE)

#Setting the column names to be that used in the dataset but in lowercase to make life a bit easier
colnames(experim) <- tolower(colnames(experim))

```

```{r}
#| label: add-agegrp
#| output: false


#The goal of this code is to divide participants into three age groups (Low, Mid, High) based on the distribution of ages in the dataset. This is purely for illustration purposes - it is not something you should do in the real world.

# Calculate quantile-based breaks including min and max
q_breaks <- quantile(experim$age, probs = seq(0, 1, length.out = 4), na.rm = TRUE)
#quantile(experim$age, probs = seq(0, 1, length.out = 4)) calculates the boundaries that divide the age distribution into three equal parts.

#The first value (0%) is the minimum age,
#the second (33%) is the first tertile,
#the third (67%) is the second tertile,
#and the last (100%) is the maximum age.

# Create age groups
experim$age_group <- cut(
  experim$age,
  breaks = q_breaks,
  include.lowest = TRUE,
  labels = c("Undergraduate", "Postgraduate", "Mature Student")
)

# cut() then uses these boundaries to categorize each age into:
#"Low" for the youngest third; "Mid" for the middle third; "High" for the oldest third
#include.lowest = TRUE ensures the smallest age value is included in the first group.
```

# 1. Introduction
* Include in this section:
* **Context and rationale:** Some background
* **Research aim:** One or two sentences summarising what the report investigates.
* **Research question and hypotheses:**

* Example:

This analysis uses data from a study investigating the impact of two different types of interventions designed to help students cope with anxiety about an upcoming statistics course.

Students were randomly assigned to one of two equal groups and asked to complete several psychological scales at **Time 1 (pre-intervention)**. These included:

- *Fear of Statistics Test*  
- *Confidence in Coping with Statistics Scale*  
- *Depression Scale (Measured using the Beck Depression Inventory (BDI) (Beck, Steer, & Brown, 1996)*

Following this baseline measurement:

- **Group 1:** participated in sessions designed to *improve mathematical skills*  
- **Group 2:** participated in a program designed to *build confidence* in coping with statistics  

After the intervention (**Time 2**) and again at a three-month follow-up (**Time 3**), students completed the same scales. Their **performance on a statistics exam** was also recorded as a measure of learning outcome.

The purpose of this analysis is to explore the extent to which **type of support** (maths skills vs. confidence-building), **age group**, and **students’ depression levels at the end of the course** can be considered predictors of **exam performance**.

Before developing a predictive model, a series of preliminary analyses were conducted to establish statistical evidence supporting the inclusion of these variables as potential predictors. A set of hypotheses were derived which together test whether each potential predictor (depression, type of support, and age group) shows a meaningful association with exam performance (see Table 1). Evidence from the analysis of these hypotheses will provide the basis for including these variables in a subsequent **multiple linear regression model** predicting exam outcomes.


```{=html}
<table style="width:100%; border-collapse:collapse;">
  <thead>
    <tr style="background-color:#f2f2f2;">
      <th style="text-align:left; padding:8px;">Analysis Focus</th>
      <th style="text-align:left; padding:8px;">Null Hypothesis (H₀)</th>
      <th style="text-align:left; padding:8px;">Alternative Hypothesis (H₁)</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td style="padding:8px;"><strong>Depression and Exam Performance</strong></td>
      <td style="padding:8px;">A student’s level of depression at the end of the course (Time 3) is <strong>not related</strong> to their final exam performance.</td>
      <td style="padding:8px;">A student’s level of depression at the end of the course (Time 3) <strong>is related</strong> to their final exam performance.</td>
    </tr>
    <tr>
      <td style="padding:8px;"><strong>Type of Support and Exam Performance</strong></td>
      <td style="padding:8px;">The type of support provided (maths skills vs. confidence-building) <strong>does not affect</strong> students’ final exam performance.</td>
      <td style="padding:8px;">The type of support provided (maths skills vs. confidence-building) <strong>does affect</strong> students’ final exam performance.</td>
    </tr>
    <tr>
      <td style="padding:8px;"><strong>Age Group and Type of Support</strong></td>
      <td style="padding:8px;">A student’s age group and the type of support provided are <strong>independent</strong> (unrelated).</td>
      <td style="padding:8px;">A student’s age group and the type of support provided are <strong>associated</strong>.</td>
    </tr>
    <tr>
      <td style="padding:8px;"><strong>Age Group and Exam Performance</strong></td>
      <td style="padding:8px;">Students’ exam performance <strong>does not differ</strong> across age groups.</td>
      <td style="padding:8px;">Students’ exam performance <strong>differs</strong> across age groups.</td>
    </tr>
  </tbody>
</table>

  <div style="text-align:center; font-weight:bold; padding-top:8px;">
  Table 1. Hypotheses for Planned Analysis.
  
</div>
  
```

In the remainder of this report, Section 2 describes the dataset used for the analysis, including participant demographics, variable descriptions, and data preparation steps. Section 3 outlines the analytical approach, detailing the statistical methods and rationale for their selection. Section 4 presents the results of the analyses, organised by hypothesis. Section 5 discusses the main findings in the context of the research aims, highlighting their implications and limitations. Finally, Section 6 provides a concise conclusion, summarising the key insights and potential directions for future research. References and any supplementary materials are provided in Section 7 and the Appendix, respectively.



# 2. Dataset

* **Participants:** number of cases, key demographics.
* **Missing data / cleaning:** discuss levels of missingness and implications; mention if any cases or variables were removed or recoded or imputed.
* **Amendments:** discuss any recoding of variables e.g. in this case age group was created
* ** Describe the variables relevant to the analysis **
  * **Continuous variables** this would include your normality assessment
  * **Grouping variables** 
  * **Include visuals**

Divide into sub-sections as you wish.
This section can include a small table summarising variable names, types, and ranges.
You can include the visuals in your main report or in an appendix as you prefer.

* Example:

The dataset used in this study consists of 300 observations. The concepts of interest are outlined in Table 2. Each associated variable was examined to determine whether the level of missing data could introduce bias. No missing data were identified.

Of the participants, 51.33% identified as male and 48.67% as female. Regarding the support groups, 50% of students were assigned to the maths skills group and 50% to the confidence skills groups. A new variable was created to recode students’ ages into categorical age groups. In this classification, 36.67% of the students were identified as undergraduates (19–22 years), 33.67% as postgraduates (23–30 years), and 29.66% as mature students (30+ years).

```{=html}
<table style="width:100%; border-collapse:collapse;">
  <thead>
    <tr style="background-color:#f2f2f2;">
      <th style="text-align:left; padding:8px;">Concept</th>
      <th style="text-align:left; padding:8px;">Statistical Type</th>
      <th style="text-align:left; padding:8px;">Description</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td style="padding:8px;"><strong>Depression</strong></td>
      <td style="padding:8px;">Continuous (Ratio)</td>
      <td style="padding:8px;">A student’s level of depression at the end of the course. Single measure calculated from student anwsers to Beck Depression Inventory.</td>
    </tr>
    <tr>
      <td style="padding:8px;"><strong> Exam Performance</strong></td>
      <td style="padding:8px;"> Continuous (Ratio)</td>
      <td style="padding:8px;">The student's numerical result on their final examination.</td>
    </tr>
    <tr>
      <td style="padding:8px;">Type of Support</strong></td>
      <td style="padding:8px;">Nominal (Binary).</td>
      <td style="padding:8px;">A grouping variable to indicate the type of support provided: maths skills or condfidence skills.</td>
    </tr>
    <tr>
      <td style="padding:8px;"><strong>Age Group</strong></td>
      <td style="padding:8px;">Nominal (Multi-category).</td>
      <td style="padding:8px;">A categorisation of students age: Undergraduate (19-22), Postgraduate (23-25), Mature Student(25+)) </td>
    </tr>
  </tbody>
</table>

<div style="text-align:center; font-weight:bold; padding-top:8px;">
  Table 2. Concepts of Interest for Planned Analysis.
  
</div>

```

Before statistical tests were conducted, the data representing the relevant concepts was inspected to determine the correct test to be used. 



Students' level of depression at the end of the course were assessed for normality. Visual inspection of the histogram and QQ-Plot (see Figure 1 and Figure 2) suggested some potential deviations due to skewness and kurtosis.. However, the standardized score for skewness (0.66) and for kurtosis (-2.47) can be considered acceptable using the criteria proposed by Curran, West and Finch (1996). Further as  as over 95% of standardized scores fall within the bounds of +/-1.96 and 100% within the bounds of +/- 3.29, using the guidance of Field, Miles and Field (2012), there are insufficient outliers to be considered problematic. Therefore the data for depression can be considered to approximate a normal distribution (m=11.48, sd=6.23,n=300).

```{r}
#| fig-cap: "Figure 1: Histogram of Depression at the end of the Course"
#Create histogram
ggplot(experim, aes(x = depress3)) +
  labs(x = "Depression at the end of the Course") +
  geom_histogram(
    binwidth = 2,
    colour = "black",
    aes(y = after_stat(density), fill = after_stat(count))
  ) +
  scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C") +
  stat_function(
    fun = dnorm,
    color = "red",
    args = list(
      mean = mean(experim$depress3, na.rm = TRUE),
      sd = sd(experim$depress3, na.rm = TRUE)
    )
  )
```
```{r}
#| fig.cap: "Figure 2: QQPlot for Depression at the end of the Course"
#Create a qqplot
qqnorm(experim$depress3)
qqline(experim$depress3, col=2) #show a line on the plot
```



Students' results on an examination at the end of the course were assessed for normality. Visual inspection of the histogram and QQ-Plot (see Figure 3 and Figure 4) suggested some potential deviations due to skewness and kurtosis. However, the standardized score for skewness (-1.18) and for kurtosis (1.47) can be considered acceptable using the criteria proposed by Curran, West and Finch (1996). While only 94% of standardised scores fell within  the bounds of +/-1.96, 99.7% fall within the bounds of +/- 3.29. Therefore using the guidance of Field, Miles and Field (2012), there are insufficient outliers to be considered problematic. Therefore the data for exam performance can be considered to approximate a normal distribution (m=67.1,sd=11.97,n=300).


```{r}
#| fig-cap: "Figure 3: Histogram of Exam Performance at the end of the Course"
#Create histogram
ggplot(experim, aes(x = exam)) +
  labs(x = "Exam Performance at the end of the Course") +
  geom_histogram(
    binwidth = 2,
    colour = "black",
    aes(y = after_stat(density), fill = after_stat(count))
  ) +
  scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C") +
  stat_function(
    fun = dnorm,
    color = "red",
    args = list(
      mean = mean(experim$exam, na.rm = TRUE),
      sd = sd(experim$exam, na.rm = TRUE)
    )
  )
```
```{r}
#| fig.cap: "Figure 4: QQPlot for Exam Performance at the end of the Course"
#Create a qqplot
qqnorm(experim$exam)
qqline(experim$exam, col=2) #show a line on the plot
```






--- We will complete this section during class --

# 3. Analysis Approach

Explain what you did statistically:

* State the cut-off level you are working with for statistical significance
* State the types of tests and why they’re appropriate:

  * Correlation (for continuous–continuous relationships)
  * Independent samples *t*-test/Mann-Whitney (for two groups)
  * Chi-square test (for two categorical variables)
  * One-way ANOVA/Kruskal Wallis (for differences across three groups)
  
* State the effect size conventions you will be using.

* Example:
An alpha level of .05 was used as the threshold for statistical significance, consistent with conventional standards in psychological and behavioural research. Results with p values less than .05 were considered statistically significant. To evaluate the practical significance of findings, effect sizes were reported alongside p values. For parametric and non-parametric difference tests, eta squared (η²) was used as the measure of effect size, interpreted according to Cohen’s (1988) conventions, where values of approximately .01, .06, and .14 indicate small, medium, and large effects, respectively. For correlational analyses, Pearson’s r (or its non-parametric equivalent, Spearman’s rho/Kendall's tau) was reported, with effect sizes interpreted using Cohen’s benchmarks of .10, .30, and .50 for small, medium, and large associations, respectively. Reporting both statistical significance and effect size facilitates a more comprehensive interpretation of the data’s substantive importance.


-- We will complete this section during class --

# 4. Results

## 4.1 Relatioship between Depression and Exam Performance	
H0:A student’s level of depression at the end of the course (Time 3) is not related to their final exam performance.	
H1: A student’s level of depression at the end of the course (Time 3) is related to their final exam performance.

The relationship between depression and exam performance was investigated using a Pearson correlation. No statistically significant, correlation was found (r =-0.01, n=298, p=0.87). There is therefore no evidence to reject the null hypothesis in favour of the alternate. We can conclude therefore that there appears to be no relationship between depression and exam performance, see Figure 5.

```{r}
#| label: "scatter"
#| fig-cap: "Figure 5: Scatterplot Depression and Exam Performance At the of the Course"
#Create a Simple scatterplot of feeling of control and exam performance
#aes(x,y)
scatter <- ggplot(experim, aes(depress3, exam))
scatter + geom_point() + labs(x = "Depression at the end of the course", y = "Exam performance at the end of the course") 

#Add a regression line
scatter + geom_point() + geom_smooth(method = "lm", colour = "Red", se = F) + labs(x = "Depression at the end of the course", y = "Exam performance at the end of the course")
```

## 4.2 Does the Type of Support Provided Impact  Exam Performance	
H0: The type of support provided (maths skills vs. confidence-building) does not affect students’ final exam performance.	
H1: The type of support provided (maths skills vs. confidence-building) does affect students’ final exam performance.


An independent-samples t-test was conducted to compare exam performance scores for students who had maths skills support and those who had confidence skills support. Levene’s test for equality of variances was not statistically significant, indicating that the assumption of homogeneity of variance was met (p > .05). No significant difference in the exam results was found (M=67.40, SD= 12.77 for students who had confidence skills support, M= 66.79, SD= 11.16 for students who had maths skills support), (t(298)= 0.49, p = 0.66. A very small effect size was also indicted by the eta squared value (0.001). There is therefore no evidence to support rejecting the null hypothesis in favour of the alternate. There appears to be no significant difference in exam performance for different support offered.


# 4.3 Assoication between Age Group and Type of Support	
H0: A student’s age group and the type of support provided are independent (unrelated).	H1: A student’s age group and the type of support provided are associated.


A Chi-square test of independence was conducted to examine the relationship between being age group category and support group assigned. The association was not statistically significant (χ²(2, N = 298) = 0.93, p =.63). A student's age group did not impact the support group to which they were assigned. The effect size, measured by Phi (φ = 0.05), indicates a small association. There is therefore evidence to support rejecting the null hypothesis in favour of the alternate hypothesis.

## 4.4 Difference in Exam Performance	for Different Age Groups
H0: Students’ exam performance does not differ across age groups.	
H1: Students’ exam performance differs across age groups.


### Reporting the results with eta squared effect

A one-way between-groups analysis of variance (ANOVA) was conducted to explore the impact of exam performance. Participants were divided into three groups according to their age (Group 1: 22 yrs or less; Group 2: 23 to 30 yrs; Group 3: 40yrs and above). No statistically significant difference in exam results was found for the three age groups: (F(2, 297)= 2.05, p=0.13.  The effect size, calculated using eta squared was (η² = 0.01), indicated a small effect. The mean score for Group 1 (M=67.54,SD=12.33) was not significantly different to that for Group 2 (M=68.43,SD=11.26) or  Group 2 (M=65.03,SD=12.17). Group 2 and 3 did not differ significantly from each other. Therefore, the null hypothesis cannot be rejected in favour of the alternative hypothesis. Age does not appear to impact exam performance. 


Organise by  hypothesis- a sub-section for each. For each:

1. Briefly restate the hypothesis
2. Present the inferential test result (test statistic, df, *p*-value, and effect size)
3. Interpret in plain language


-- We will complete this section during class --

# 5. Discussion

* **Summarise findings:** What patterns emerged?
* **Interpret results:** Why might these differences or relationships exist?
* **Discuss limitations:** small sample, sources of bias, etc.
* **Implications:** 

# 6. Conclusion

A concise paragraph that brings everything together — what was learned overall, and what it suggests for future work.


# 7. References
List any articles, books or sources cited.

Beck, A. T., Steer, R. A., & Brown, G. K. (1996). Manual for the Beck Depression Inventory–II. San Antonio, TX: Psychological Corporation.
Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd ed.). Hillside, NJ: Lawrence Erlbaum Associates.
Curran, Patrick J., Stephen G. West, and John F. Finch. (1996). "The robustness of test statistics to nonnormality and specification error in confirmatory factor analysis."
Field, A., Field Z. & Miles J.(2012).  Discovering statistics using IBM SPSS statistics. Sage publications limited.

# Appendix
Optional
If you choose to include your visuals in the appendix.

* Example:
## General Descriptives

### Gender Proportions:
```{r}
#| label: desc-gender
# Proportions of gender
prop.table(table(experim$sex))
```
### Age Group Proportions:
```{r}
#| label: desc-age

#Get the numbers in each age group
table(experim$age_group)

# Show the range of ages in each group
prop.table(table(experim$age_group))
```

### Group Proportions:
```{r}
#| label: desc-group

#Get the numbers in group
table(experim$group)

# Show the range of ages in each group
prop.table(table(experim$group))
```


## Missing Data

```{r}
#| label: missing
varsint<-c("depress3", "exam", "group", "age_group")
esub<-experim[varsint]
#Adjust the margins so that we can display the labels
par(mar = c(10, 4, 4, 2))  # more space at bottom
# rename columns to human-friendly labels - we are restricted with the length we can use
colnames(esub) <- c("Depression", "Exam", "Group", "AgeGrp")
# run VIM::aggr
res <- summary(VIM::aggr(esub, sortVar = TRUE, combined = FALSE))$combinations

# Include Little's MCAR if you have data missing
```

## Normality

```{r}
#|label: desc-depress
#Generate regular summary statistics 
psych::describe(experim$depress3)

#Generate standardized scores for skewness and kurtosis
tpskew<-semTools::skew(experim$depress3)
tpkurt<-semTools::kurtosis(experim$depress3)
tpskew[1]/tpskew[2]
tpkurt[1]/tpkurt[2]

#Calculate the percentage of standardized scores outside acceptable limits
zdepress3<- abs(scale(experim$depress3))

FSA::perc(as.numeric(zdepress3), 1.96, "gt")
FSA::perc(as.numeric(zdepress3), 3.29, "gt")
# Below we are reporting the normality assessment - embedding code within a paragraph to get the statistics generated ( you can just write it out)

```

```{r}
#|label: desc-exam
#Generate regular summary statistics 
psych::describe(experim$exam)

#Generate standardized scores for skewness and kurtosis
tpskew<-semTools::skew(experim$exam)
tpkurt<-semTools::kurtosis(experim$exam)
tpskew[1]/tpskew[2]
tpkurt[1]/tpkurt[2]

#Calculate the percentage of standardized scores outside acceptable limits
zexam<- abs(scale(experim$exam))

FSA::perc(as.numeric(zexam), 1.96, "gt")
FSA::perc(as.numeric(zexam), 3.29, "gt")
# Below we are reporting the normality assessment - embedding code within a paragraph to get the statistics generated ( you can just write it out)

```






```{r}
#| label: "correlation test"
#Pearson Correlation - this is the correct test for this data
stats::cor.test(experim$depress3, experim$exam, method='pearson')

```

## t-test




```{r}
#| fig.cap: "Figure 6: Histogram of Exam Performance Grouped by student's Type of Support"

# Create a separate histogram showing the distribution for the two (potentially) separate populations (the groups)
# Create the plot with both histogram and density plot
ggplot(cleaned_experim, aes(x=exam, color=group, fill=group))  + 
  labs(x="Total exam performance") + 
  # Histogram with alpha for transparency
  geom_histogram(aes(y = ..density..), position="identity", alpha=0.5, binwidth=1, color="black") +
  # Density plot with transparency for overlapping curves
  geom_density(alpha=0.4) +
  # Customize the color and fill scales
  scale_color_manual(values=c("blue", "red"), labels=c("Maths Skills", "Confidence Skills")) +
  scale_fill_manual(values=c("lightblue", "lightcoral"), labels=c("Maths Skills", "Confidence Skills"))

```

```{r}
#| fig.cap: "Table 4: Descriptive Statistics of Exam performance Grouped by Respndent's Type of Support"
#Get descriptive statistics by group - output as a matrix
psych::describeBy(experim$exam, experim$group, mat=TRUE)

```


## Independent t-test

```{r}
#Conduct Levene's test for homogeneity of variance in library car - the null hypothesis is that variances in groups are equal so to assume homogeneity we would expect probability to not be statistically significant.
car::leveneTest(exam ~ group, data=experim)
#Pr(>F) is your probability - in this case it is not statistically significant so we can assume homogeneity


#Conduct the t-test from package stats - basic r package. Prefixing it to make sure that is the one used rather than t-test from another package.

#In this case we can use the var.equal = TRUE option to specify equal variances and a pooled variance estimate
stats::t.test(exam~group,var.equal=TRUE,data=experim)
#No statistically significant difference was found

res <- stats::t.test(exam~group,var.equal=TRUE,data=experim)
#Calculate Cohen's d
#artithmetically
effcd=round((2*res$statistic)/sqrt(res$parameter),2)

#Alternate approach -Using function from effectsize package to convert t to Cohen's d
effectsize::t_to_d(t = res$statistic, res$parameter)

#Calculate eta squared
effes=round((res$statistic*res$statistic)/((res$statistic*res$statistic)+(res$parameter)),3)
effes

```

## Nominal - chi

## Approach 1 - use the Crosstable function from gmodels

```{r}
#CrossTable(predictor, outcome, fisher = TRUE, chisq = TRUE, expected = TRUE)
#Note: for this approach there must be the same number of values for both variables so
#missing data is important. 

# Remove rows where either variable is NA
experim_clean <- experim %>%
  filter(!is.na(age_group) & !is.na(group))

# Now run CrossTable requesting both Fisher and Chi Square so we have full information
# We request the format in SPSS to make it easier to interpret.
gmodels::CrossTable(
  experim_clean$age_group,
  experim_clean$group,
  fisher = TRUE,
  chisq = TRUE,
  expected = TRUE,
  sresid = TRUE,
  format = "SPSS"
)

```

## Approach 2: Create your own contingency table and use the chi square test from the stats package

```{r}
#Alternate approach: Create your own contingency table
mytable<-xtabs(~age_group+group, data=experim)
#You can use the chisq.test function in the stats package to do the test
#set the flag correct=TRUE to get Yates correction needed for 2x2 table
ctest<-stats::chisq.test(mytable, correct=TRUE)#chi square test

ctest#will give you the details of the test statistic and p-value
ctest$expected#expected frequencies
ctest$observed#observed frequencies
ctest$p.value
#Calculate effect sizes
sjstats::phi(mytable)
sjstats::cramer(mytable)

```
## ANOVA


```{r}
#Get descriptive statistics by group - output as a matrix
psych::describeBy(experim$exam, experim$age_group, mat=TRUE)
#Store the output to use in our final reporting of the outcomes of ANOVA

optdesc<-psych::describeBy(experim$exam,experim$age_group,mat=TRUE)
```

```{r}
#| label: ANOVA-optage_group
#Conduct Bartlett's test for homogeneity of variance  - the null hypothesis is that variances in groups are equal so to assume homogeneity we would expect probability to not be statistically significant.
stats::bartlett.test(exam~ age_group, data=experim)
#p value is > 0.05 so the result is not statistically significant so we can assume homogeneity

#Conduct ANOVA using the one-way ANOVA from the stats package

# Compute the analysis of variance using the oneway.test from the stats package- we set VAR.equal to be true because we have homogeneity
res <- stats::oneway.test(exam ~ as.factor(age_group), data = experim, var.equal = TRUE)
# Summary of the analysis
res

#We use Tukey for post-hoc testing
rstatix::tukey_hsd(experim, exam ~ age_group)

#Compute our Eta squared
effes=effectsize::effectsize(res)
effes

```
