---
title: "Water Quality Metric Data Analysis"
subtitle: "TU060 - Determining water quality metrics rlationships and predictors"
author: "Eamonn Kelly  - Student ID: D24127620"
format:
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    toc: true
editor: source
---

```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}

quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("pastecs")    # Tools for descriptive statistics, time series analysis, and data exploration (e.g., stat.desc)
quiet_library("psych")   # Descriptive statistics by group
quiet_library("FSA")        # Includes general tools for data analysis (e.g., nonparametric tests, bootstrapping)
quiet_library("semTools")   # Supplementary functions for Structural Equation Modeling (SEM), including model fit, reliability, and measurement invariance tests
quiet_library("VIM") # Visualization and Imputation of Missing Values
quiet_library("naniar") #(naniar short for NA + naniar) - Missing data wrangling plus Little's test
quiet_library("simputation") # For imputing variables

quiet_library("sjstats")  # Provides convenient functions for effect sizes
quiet_library("gmodels")  # Includes 'CrossTable()' for contingency tables.

quiet_library("car")   # Includes Levene's test for homogeneity
quiet_library("effectsize")   # Includes calculations for effectsize
quiet_library("rstatix")   # Includes calculations for effectsize and Dunn's test

quiet_library("gtsummary") # Formatting tables
quiet_library("flextable") # Formatting tables
quiet_library("dplyr") # Formatting tables
quiet_library("knitr") # Formatting tables
quiet_library("tidyr") # Formatting tables


```

```{r}
#| label: loaddata
#Amend this to reflect your folder structure 
datalocation <- file.path("./data")
datapath <- file.path(datalocation,"WQI Results on Dataset.csv")

#Read in the csv file dataset
wqi_df <- read.csv(datapath, header=TRUE)

#Setting the column names to be that used in the dataset but in lowercase to make life a bit easier
colnames(wqi_df) <- tolower(colnames(wqi_df))


```

```{r}
#| label: craete_data_subset
#
# Define variables that we want to use in or analysis
# we'll craete a subset of the data with just those to keep ru dataset as light as we can.
varsint <- c("waterbodyname",  "years",  "sampledate",  "alkalinity.total..as.caco3.",  "bod...5.days..total.",  "conductivity..25.c",  "dissolved.oxygen",  "ph",  "temperature",  "ccme_wqi",  "wawqi",  "horton.wqi",  "brown.wqi",  "srdd.wqi")

# create data subset of variables we want to lok at
wqi_dsubset <- wqi_df[varsint]


# just get human friendly labels
colnames(wqi_dsubset) <- c( "waterbodyname", "years", "month", "alkalinity",  "biochem_ox_dmnd", "conductivity", "dissolved.oxygen", "ph",  "temperature", "ccme_wqi",  "wawqi",  "horton.wqi",  "brown.wqi",  "srdd.wqi")



```
## Introdction


This analysis uses water quality data which was taken over a period of 16 years from 2007 to 2023 across 16 locations in Ireland by the Environmental Protection Agency(EPA). The data was then used to classify water quality at these locations.

Water samples were taken from 160 sites within these 16 geographical areas, at various times throughout the year, tested in labs, and calculations were then carried out to determine Water Quality according to different water quality standards, a number of which exist.


The purpose of this analysis is to evaluate the relationships among key physical, geographic and time based patterns to gain a better understanding how individual factors influence overall water quality across these Irish monitoring sites. The aim will be to attempt to identify key predictors of water quality and patterns determining water quality. The analysis examines whether variables such as **ph**, **temperature**, **biochemical oxygen demand *BOD**, **dissolved oxygen** vary by location, season, and which have the greatest impact in determining water quality.

Before developing a predictive model, a series of preliminary analyses were conducted to establish statistical evidence supporting the inclusion of these variables as potential predictors and major determining factors. A set of hypotheses were derived which together test potential predictors (**ph**, **temperature**, **biochemical oxygen demand *BOD**, **dissolved oxygen**) and show a significant association with water quality. Evidence from the analysis of these hypotheses will provide the basis for including these variables in a subsequent **multiple linear regression model** predicting exam outcomes.


## Dataset

The data set was sourced from the following location 
[Water Quality Monitoring Dataset (Ireland)](https://figshare.com/articles/dataset/Water_Quality_Monitoring_Dataset_Ireland_/25002131). Thsi dataset is a merging of multiple csv files from the EPA into a single source. 

We have reduced the dataset down further into a subset of variables which are of interest to us for our analysis.


## Description

 Our dataset has `29,159` rows, and `14` variables. Then variables and statistical types are listed in the below table.


| variable         | r_type  | is_factor | factor_summary                           | stat_type             |
| ---------------- | ------- | --------- | ---------------------------------------------------------------- | 
| waterbodyname    | factor  | TRUE      | 160 levels (e.g., ABBEYTOWN_010, Allua,) | Categorical (Nominal) |
| years            | integer | FALSE     | NA                                       | Discrete (Ratio)      |
| month            | factor  | TRUE      | 12 levels (e.g., Jan, Feb, …)            | Categorical (Ordinal) |
| alkalinity       | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| biochem_ox_dmnd  | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| conductivity     | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| dissolved.oxygen | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| ph               | double  | FALSE     | NA                                       | Continuous (Interval) |
| temperature      | double  | FALSE     | NA                                       | Continuous (Interval) |
| ccme_wqi         | factor  | TRUE      | 5 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| wawqi            | factor  | TRUE      | 6 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| horton.wqi       | factor  | TRUE      | 6 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| brown.wqi        | factor  | TRUE      | 5 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| srdd.wqi         | factor  | TRUE      | 6 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |

Table 2. Variable and Statistical Datatypes present in the data set

```{r}
summary(wqi_dsubset)
head(wqi_dsubset)
```

| Variable | Range (Min–Max) | Mean | SD | Median | IQR (1st Qu. - 3rd Qu| 
|-----------|-----------------|------|----|---------|--------------|
| **Year** | 2007–2023 | 2015 | 4.5 | 2015 | 2011–2018 | 
| **Alkalinity (mg/L)** | 0–442 | 139.9 | 100.2 | 127 | 55–214 |
| **Biochemical Oxygen Demand (BOD, mg/L)** | 0.0–16.0 | 1.28 | 0.65 | 1.20 | 1.2–1.2 | 
| **Conductivity (µS/cm)** | 33–4200 | 363 | 270.0 | 356 | 229–501 |  saline influences. |
| **Dissolved Oxygen (mg/L)** | 0–198 | 58.29 | 22.5 | 54.9 | 49.4–75.0 |
| **pH** | 4.7–9.8 | 7.55 | 0.7 | 7.7 | 7.2–8.0 | 
| **Temperature (°C)** | 0.6–637.0* | 10.85 | 2.5 | 10.7 | 8.2–13.1 |

Table 3. Continuous (Interval/Ratio) Data Descriptions

**Some General observations:**
- **year** - taken over 16 year period.
- **Biochemical Oxygen Demand (BOD, mg/L)** - very stable data, 75% of data below 1.2 value, there is a potential outlier present though
- **ph** - mean values
- **conductivity** - big spread of data of data
| **Temperature (°C)** - range from just above freezing and 3rd quantile approx 13 C. There is an outlier here.

There do appear to be some outliers, particularly in temperature, with a temperature of 637 C not be a real value. This ms be due to a broken thermometer, or perhaps data entry error. Conductivity, has a 3rd quantile value of 501 yet has a max value of 42000. This could eb an outlier however we need to determine if exrteme values are outliers or actual real data due to extreme environmental conditions

Needs further investigation, as such decided to remove this variable for the moment, and proceed more reliable data from others. Environments can have extreme values when extreme environmental conditions occur, this could  be problematic when determining if a data vlue is an error or actual real measurement. 

### Missing Data

There was no mising data present in the dataset. Analysis was performed on the dataset and no missing data was found. Various ypes of elements were search for including *?*, *unknown*, *none*, *na*, *n/a*, *missing*, *null*, *NULL*, *NaN*, *-*, *--*, * *, and none of these were found to exist. See Appendix for more details.



### Continuous Variables - Normality and Outliers

#### ph value

```{r}
#| label: hist_ph
#| fig.cap: "Figure 1: Histogram of ph"

# create a histogram of ph with density curve over it
# some referenes taken from hare https://stackoverflow.com/questions/5688082/overlay-histogram-with-density-curve
gg <- ggplot(wqi_dsubset_clean, aes(x = ph))
gg <- gg + labs(
  title = "Histogram of ph",
  x = "ph",
  y = "Density"
)
# draw the histogram with bin width and colour defined
gg <- gg + geom_histogram(
  binwidth = 0.2,
  colour = "black",
  aes(y = ..density.., fill = ..count..)
)
# sets the colour gradient for the histogram
gg <- gg + scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C")
# creates the normal distribution curve using probability density function
# using the mean and standard deviation from our dataset ph variable
gg <- gg + stat_function(
  fun = dnorm,
  color = "red",
  args = list(
    mean = mean(wqi_dsubset_clean$ph, na.rm = TRUE),
    sd = sd(wqi_dsubset_clean$ph, na.rm = TRUE)
  )
)
# set the minimal style
gg <- gg + theme_minimal()
# call to dislpay the plot
gg

```


```{r}
#| label: qq_ph
#| fig.cap: "Figure 2: QQ Plot of ph values"

# craete qq plot for temperature
# plots the data quantiles on y axis and theoritical values on x axis
qqnorm(wqi_dsubset_clean$ph,
       main = "QQ Plot of ph values",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col = "steelblue",
       pch = 19)
# drawa the comparison line which is standard normal
qqline(wqi_dsubset_clean$ph, col = "red", lwd = 2)
```
```{r}
#| label: desc_ph
#| fig.cap: "Table 3: Descriptive Statistics for pH Values"

# 
# use pastec stats describe to get basic statistics
pastecs::stat.desc(wqi_dsubset_clean$ph, basic = FALSE)

# will give us mean, sd, median, skew, kurtosis, etc.
psych::describe(wqi_dsubset_clean$ph)

# Standardized skewness and kurtosis
ph_skew <- semTools::skew(wqi_dsubset_clean$ph)
ph_kurt <- semTools::kurtosis(wqi_dsubset_clean$ph)
ph_skew_z <- ph_skew[1] / ph_skew[2]
ph_kurt_z <- ph_kurt[1] / ph_kurt[2]

cat("Standardised skewness (z):", round(ph_skew_z, 2), "\n")
cat("Standardised kurtosis (z):", round(ph_kurt_z, 2), "\n")

#Calculate percentage of standardized scores for total sleep satisfcation
# are within our limits of +/- 1.96, +/- 3.29
zph <- abs(scale(wqi_dsubset_clean$ph))

# % beyond ±1.96 (approx 5% expected in normal data)
FSA::perc(as.numeric(zph), 1.96, "gt")

# % beyond ±3.29 (approx 0.1% expected in normal data)
FSA::perc(as.numeric(zph), 3.29, "gt")
```
Visual inspection of the ph histogram and Q–Q plot indicated a roughly symmetrical, unimodal distribution of pH values. Although standardized skewness (z = –65.23) and kurtosis (z = 26.11) were statistically significant, only 6.1 % of standardized scores exceeded ±1.96 and 0.39 % exceeded ±3.29, using the criteria proposed by Curran, West and Finch (1996). As 99.6% of standardized scores fall within the bounds of +/- 3.29, using the guidance of Field, Miles and Field (2012), the data can be considered to approximate a normal distribution.

#### temperature values

We identified a potential issue earlier with temperature concerngin outliers. we will address that now and check for normality

```{r}
#| label: 1_del_outliers_temp_view 

# sort tempe in descending order and show top 20 values
# craete a d to hold these topw 20 avlues, arraneg i desc order, and use dply packge to select the appropriate columns so we can see all the data affected and if there are any other factors that come into play
top20_temp <- wqi_dsubset_clean %>%
  arrange(desc(temperature)) %>%
  dplyr::select(waterbodyname, years, month, temperature, ph, biochem_ox_dmnd, dissolved.oxygen, conductivity) %>%
  head(20)

# use knitr to create a table of values
# details sbased around examples from here https://www.geeksforgeeks.org/r-language/kable-method-in-r/
knitr::kable(top20_temp, caption = "top 20 highest temp values")

```


We see that the top 5 temp values are in excess of any possible real values in Ireland. We also note that the top 5 values are from 3 locations, so some repeats. This could eb a thermomter issue or a data entry issue. Given the size of our dataset at 29152 rows of data we will delete these 5 rows. We will do so by removing any temp value above 40 degrees, to ensure we have realistic temp values.  

```{r}
#| label: 2_del_outliers_del

# get a count how many valeus are > than 40
sum(wqi_dsubset_clean$temperature > 40, na.rm = TRUE)

# delete these rows 
wqi_dsubset_clean_t <- wqi_dsubset_clean %>% filter(temperature <= 40 | is.na(temperature))

# compare how many rows were dropped
n_removed <- nrow(wqi_dsubset_clean) - nrow(wqi_dsubset_clean_t)
cat("Rows removed due to temperature > 40°C:", n_removed, "\n")

# pritn otu nrows left - shoudl eb 5 less than 29159 - 5  = 29154, just to check nothing untoward happened
cat("Rows remaining in the dataset =", nrow(wqi_dsubset_clean_t), "\n")

```



```{r}

#| label: hist_temp
#| fig.cap: "Figure 3: Histogram of Temp values"
# using clean_t datafrane now from here on. new cleaned data fram - temp outliers
# create a histogram of ph with density curve over it
gg <- ggplot(wqi_dsubset_clean_t, aes(x = temperature))
gg <- gg + labs(
  title = "Distribution of Water Temperature Values",
  x = "Temperature (°C)",
  y = "Density"
)
# draw the histogram with bin width and colour defined
gg <- gg + geom_histogram(
  binwidth = 1, 
  colour = "black",
  aes(y = ..density.., fill = ..count..)
)
# sets the colour gradient for the histogram
gg <- gg + scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C")
# creates the normal distribution curve using probability density function
# using the mean and standard deviation from our dataset ph variable
gg <- gg + stat_function(
  fun = dnorm,
  color = "red",
  args = list(
    mean = mean(wqi_dsubset_clean_t$temperature, na.rm = TRUE),
    sd = sd(wqi_dsubset_clean_t$temperature, na.rm = TRUE)
  )
)
# set the minimal style
gg <- gg + theme_minimal()
# call to dislpay the plot
gg


```
```{r}
#| label: qq_temp
#| fig.cap: "Figure 4: QQ Plot of Temp values"

# craete qq plot for temperature
# plots the data quantiles on y axis and theoritical values on x axis
qqnorm(wqi_dsubset_clean_t$temperature,
       main = "Q–Q Plot of Water Temperature Values",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col = "steelblue",
       pch = 19)
# drawa the comparison line which is standard normal
qqline(wqi_dsubset_clean_t$temperature, col = "red", lwd = 2)

```


```{r}
#| label: desc_temp
#| fig.cap: "Table 4: Descriptive Statistics for Temp"

# 
# use pastec stats describe to get basic statistics
pastecs::stat.desc(wqi_dsubset_clean_t$temperature, basic = FALSE)

# will give us mean, sd, median, skew, kurtosis, etc.
psych::describe(wqi_dsubset_clean_t$temperature)

# Standardized for skewness and kurtosis
temp_skew <- semTools::skew(wqi_dsubset_clean_t$temperature)
temp_kurt <- semTools::kurtosis(wqi_dsubset_clean_t$temperature)
temp_skew_z <- temp_skew[1] / temp_skew[2]
temp_kurt_z <- temp_kurt[1] / temp_kurt[2]

# print otu the standadized skew and kurtosis
cat("Standardised skewness (z):", round(temp_skew_z, 2), "\n")
cat("Standardised kurtosis (z):", round(temp_kurt_z, 2), "\n")

#Calculate percentage of standardized scores for total sleep satisfcation
# are within our limits of +/- 1.96, +/- 3.29
ztemp <- abs(scale(wqi_dsubset_clean_t$temperature))

# % beyond ±1.96 (approx 5% expected in normal data)
FSA::perc(as.numeric(ztemp), 1.96, "gt")

# % beyond ±3.29 (approx 0.1% expected in normal data)
FSA::perc(as.numeric(ztemp), 3.29, "gt")
```

Visual inspection of the temperature histogram and Q–Q plot indicated a symmetrical, unimodal distribution of temp values with the QQ plot shows minor deviations from normal at the tails. Standardized skewness (z = 17.06) and kurtosis (z = -7.39) were not statistically significant, only 4.2 % of standardized scores exceeded ±1.96 and 0.065 % exceeded ±3.29, using the criteria proposed by Curran, West and Finch (1996). As 99.35% of standardized scores fall within the bounds of +/- 3.29, using the guidance of Field, Miles and Field (2012), the data can be considered to approximate a normal distribution.

####  Biochemical Oxygen Demand (BOD)

```{r}

#| label: hist_bod
#| fig.cap: "Figure 5: Histogram of Biochemical Oxygen Demand (BOD) Values"

# create a histogram of ph with density curve over it
gg <- ggplot(wqi_dsubset_clean_t, aes(x = biochem_ox_dmnd))
gg <- gg + labs(
  title = "Distribution of Biochemical Oxygen Demand (BOD) Values",
  x = "BOD (mg/L)",
  y = "Density"
)
# draw the histogram with bin width and colour defined
gg <- gg + geom_histogram(
  binwidth = 0.2,               
  colour = "black",
  aes(y = ..density.., fill = ..count..)
)
# set the colour gradient for the histogram
gg <- gg + scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C")
# creates the normal distribution curve using probability density function
# using the mean and standard deviation from our dataset ph variable
gg <- gg + stat_function(
  fun = dnorm,
  color = "red",
  args = list(
    mean = mean(wqi_dsubset_clean_t$biochem_ox_dmnd, na.rm = TRUE),
    sd = sd(wqi_dsubset_clean_t$biochem_ox_dmnd, na.rm = TRUE)
  )
)
# set the minimal style
gg <- gg + theme_minimal()
# call to dislpay the plot
gg
```



```{r}

#| label: qq_bod
#| fig.cap: "Figure 6: QQ Plot of Biochemical Oxygen Demand (BOD) Values"

# craete qq plot for temperature
# plots the data quantiles on y axis and theoritical values on x axis
qqnorm(wqi_dsubset_clean_t$biochem_ox_dmnd,
       main = "Q–Q Plot of Biochemical Oxygen Demand (BOD) Values",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col = "steelblue",
       pch = 19)
# drawa the comparison line which is standard normal
qqline(wqi_dsubset_clean_t$biochem_ox_dmnd, col = "red", lwd = 2)

```

```{r}
#| label: desc_bod
#| fig.cap: "Table 5: Descriptive Statistics for Biochemical Oxygen Demand (BOD) Values (Untransformed)"

# use pastec stats describe to get basic statistics
pastecs::stat.desc(wqi_dsubset_clean_t$biochem_ox_dmnd, basic = FALSE)

# will give us mean, sd, median, skew, kurtosis, etc.
psych::describe(wqi_dsubset_clean_t$biochem_ox_dmnd)

# Standardized scores for skewness and kurtosis
bod_skew <- semTools::skew(wqi_dsubset_clean_t$biochem_ox_dmnd)
bod_kurt <- semTools::kurtosis(wqi_dsubset_clean_t$biochem_ox_dmnd)
bod_skew_z <- bod_skew[1] / bod_skew[2]
bod_kurt_z <- bod_kurt[1] / bod_kurt[2]

# print otu the standadized skew and kurtosis 
cat("Standardised skewness (z):", round(bod_skew_z, 2), "\n")
cat("Standardised kurtosis (z):", round(bod_kurt_z, 2), "\n")

#Calculate percentage of standardized scores for total sleep satisfcation
# are within our limits of +/- 1.96, +/- 3.29
zbod <- abs(scale(wqi_dsubset_clean_t$biochem_ox_dmnd))

# % beyond ±1.96 (approx 5% expected in normal data)
FSA::perc(as.numeric(zbod), 1.96, "gt")

# % beyond ±3.29 (approx 0.1% expected in normal data)
FSA::perc(as.numeric(zbod), 3.29, "gt")

```


Visual inspection of the Biochemical Oxygen Demand (BOD) histogram indicated a highly positive skew with standardised skewness (z = 564.87) with pronounced leptokurtosis with standardised kurtosis (z = 4163.14). The Q–Q plot indicated a strong right tail, showing a small number of high BOD readings (up to 16) among predominantly low values (mean = 1.27 and median = 1.20 and a standard deviation = 0.55).

To address the skewness and non-normal distribution, we applied a transform to the log10.


##### Transformation)

```{r}
# Transform the data and save as a column in the survey dataset
# we need to add +1 as we have some zero values and wil result in an erro otherwsie
# we will create a new column'bod_log10' for the log values =, retaining the original values
wqi_dsubset_clean_t_bod <- wqi_dsubset_clean_t %>%
  mutate(bod_log10 = log10(biochem_ox_dmnd + 1))  # adds +1 to handle zeros safely

summary(wqi_dsubset_clean_t_bod )
```

```{r}

#| label: hist_bod_2
#| fig.cap: "Figure 7: Histogram of Biochemical Oxygen Demand (BOD) Values"

# create a histogram of ph with density curve over it
# we will sue our new dataframe and new log 10 column from thw transfform
gg <- ggplot(wqi_dsubset_clean_t_bod, aes(x = bod_log10))
gg <- gg + labs(
  title = "Distribution of Biochemical Oxygen Demand (BOD) Values",
  x = "BOD (mg/L)",
  y = "Density"
)
# draw the histogram with bin width and colour defined
gg <- gg + geom_histogram(
  binwidth = 0.2,               
  colour = "black",
  aes(y = ..density.., fill = ..count..)
)
# set the colour gradient for the histogram
gg <- gg + scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C")
# creates the normal distribution curve using probability density function
# using the mean and standard deviation from our dataset ph variable
gg <- gg + stat_function(
  fun = dnorm,
  color = "red",
  args = list(
    mean = mean(wqi_dsubset_clean_t_bod$bod_log10, na.rm = TRUE),
    sd = sd(wqi_dsubset_clean_t_bod$bod_log10, na.rm = TRUE)
  )
)
# set the minimal style
gg <- gg + theme_minimal()
# call to dislpay the plot
gg
```

```{r}
#| label: qq_bod_2
#| fig.cap: "Figure 8: QQ Plot of Biochemical Oxygen Demand (BOD) Values"

# craete qq plot for temperature
# plots the data quantiles on y axis and theoritical values on x axis
qqnorm(wqi_dsubset_clean_t_bod$bod_log10,
       main = "Q–Q Plot of Biochemical Oxygen Demand (BOD) Values",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col = "steelblue",
       pch = 19)
# drawa the comparison line which is standard normal
qqline(wqi_dsubset_clean_t_bod$bod_log10, col = "red", lwd = 2)

```
```{r}
#| label: desc_bod_2
#| fig.cap: "Table 6: Descriptive Stats for BOD Values - Transformed Log10"

# use pastec stats describe to get basic statistics
pastecs::stat.desc(wqi_dsubset_clean_t_bod$bod_log10, basic = FALSE)

# will give us mean, sd, median, skew, kurtosis, etc.
psych::describe(wqi_dsubset_clean_t_bod$bod_log10)

# Standardized skewness and kurtosis
bod_skew <- semTools::skew(wqi_dsubset_clean_t_bod$bod_log10)
bod_kurt <- semTools::kurtosis(wqi_dsubset_clean_t_bod$bod_log10)
bod_skew_z <- bod_skew[1] / bod_skew[2]
bod_kurt_z <- bod_kurt[1] / bod_kurt[2]

# print otu the standadized skew and kurtosis
cat("Standardised skewness (z):", round(bod_skew_z, 2), "\n")
cat("Standardised kurtosis (z):", round(bod_kurt_z, 2), "\n")

#Calculate percentage of standardized scores for total sleep satisfcation
# are within our limits of +/- 1.96, +/- 3.29
zbod <- abs(scale(wqi_dsubset_clean_t_bod$bod_log10))

# % beyond ±1.96 (approx 5% expected in normal data)
FSA::perc(as.numeric(zbod), 1.96, "gt")

# % beyond ±3.29 (approx 0.1% expected in normal data)
FSA::perc(as.numeric(zbod), 3.29, "gt")
```

Visual inspection of the histogram and Q–Q plot of the Biochemical Oxygen Demand (BOD) values, which have been transformed by Log10, show a correction to the previous positive skewness in the histogram, and while there is still deviation from the tails we now have Standardised skewness (z = 184.15) and standardised kurtosis (z = 676.68). However, aproximately 93.6% of standardized scores lay within ±1.96 , and 97,96% are within ±3.29 z, (Curran, West, & Finch, 1996). Following the guidance of Field, Miles, and Field (2012), the data can be considered to approximate a normal distribution.

#### Dissolved Oxygen (DO)

```{r}
#| label: hist_do
#| fig.cap: "Figure 9: Distribution of Dissolved Oxygen (DO) Values"

# create a histogram of ph with density curve over it
gg <- ggplot(wqi_dsubset_clean_t_bod, aes(x = dissolved.oxygen))
gg <- gg + labs(
  title = "Distribution of Dissolved Oxygen (DO) Values",
  x = "Dissolved Oxygen (mg/L)",
  y = "Density"
)
# draw the histogram with bin width and colour defined
gg <- gg + geom_histogram(
  binwidth = 2,                 # adjust as needed for smoother bins
  colour = "black",
  aes(y = ..density.., fill = ..count..)
)
# sets the colour gradient for the histogram
gg <- gg + scale_fill_gradient("Count", low = "#DCDCDC", high = "#7C7C7C")
# creates the normal distribution curve using probability density function
# using the mean and standard deviation from our dataset ph variable
gg <- gg + stat_function(
  fun = dnorm,
  color = "red",
  args = list(
    mean = mean(wqi_dsubset_clean_t_bod$dissolved.oxygen, na.rm = TRUE),
    sd = sd(wqi_dsubset_clean_t_bod$dissolved.oxygen, na.rm = TRUE)
  )
)
# set the minimal style
gg <- gg + theme_minimal()

# call to dislpay the plot
gg

```


```{r}

#| label: qq_do
#| fig.cap: "Figure 10: Q–Q Plot of Dissolved Oxygen (DO) Values"

# craete qq plot for do
# plots the data quantiles on y axis and theoritical values on x axis
qqnorm(wqi_dsubset_clean_t_bod$dissolved.oxygen,
       main = "Q–Q Plot of Dissolved Oxygen (DO) Values",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col = "steelblue",
       pch = 19)
# drawa the comparison line which is standard normal
qqline(wqi_dsubset_clean_t_bod$dissolved.oxygen, col = "red", lwd = 2)

```
```{r}
#| label: desc_do
#| fig.cap: "Table 11: Descriptive Statistics for Dissolved Oxygen (DO) Values"

# use pastec stats describe to get basic statistics
pastecs::stat.desc(wqi_dsubset_clean_t_bod$dissolved.oxygen, basic = FALSE)

# will give us mean, sd, median, skew, kurtosis, etc.
psych::describe(wqi_dsubset_clean_t_bod$dissolved.oxygen)

# Standardized skewness and kurtosis
do_skew <- semTools::skew(wqi_dsubset_clean_t_bod$dissolved.oxygen)
do_kurt <- semTools::kurtosis(wqi_dsubset_clean_t_bod$dissolved.oxygen)
do_skew_z <- do_skew[1] / do_skew[2]
do_kurt_z <- do_kurt[1] / do_kurt[2]

cat("Standardised skewness (z):", round(do_skew_z, 2), "\n")
cat("Standardised kurtosis (z):", round(do_kurt_z, 2), "\n")

# --- Percent of cases outside typical z-score thresholds ---
# Absolute z-scores for Dissolved Oxygen
zdo <- abs(scale(wqi_dsubset_clean_t_bod$dissolved.oxygen))

# % beyond ±1.96 (approx 5% expected in normal data)
FSA::perc(as.numeric(zdo), 1.96, "gt")

# % beyond ±3.29 (approx 0.1% expected in normal data)
FSA::perc(as.numeric(zdo), 3.29, "gt")

```
Visual inspection of the Dissolved Oxygen (DO) histogram and Q–Q plot indicated a bimodal or a potential  multi-modal distribution of values with the QQ plot shows minor deviations from normal at the tails, and somewhat through the centre. Standardized skewness (z = -1.88) and kurtosis (z = -4.66) were not statistically significant, and only 93% of standardized scores exceeded ±1.96 and 99.9% exceeded ±3.29, using the criteria proposed by Curran, West and Finch (1996). Using vidual plot data and the guidance of Field, Miles and Field (2012), the data can be considered to bimodal or potentially multimodal.

We will not apply a transformation to that data, rather consider it non-parametric.

### Visualization

The following are some facetted plots of variables of interest for quality standard. We are using the water quality measurement of **CCME WQI**, an its categories *Excellent* and *Poor** and the corresponding variable values that correspond to these.


```{r}
#| fig.cap: "Figure 12: Boxplot of Water Quality variance per variable"
# keep onyl quality rows that contain v aleu of interest
wqi_subset <- wqi_long %>%
  filter(ccme_wqi %in% c("Excellent", "Poor"))

# boxplots for each parameter
ggplot(wqi_subset, aes(x = ccme_wqi, y = value, fill = ccme_wqi)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  facet_wrap(~ parameter, scales = "free_y", ncol = 2) +
  theme_minimal() +
  labs(
    title = "Comparison of Key Water Quality Parameters\nBetween 'Excellent' and 'Poor' CCME WQI Classes",
    x = "CCME WQI Category",
    y = "Measured Value"
  ) +
  theme(legend.position = "none")

```

Bocplots of water quality classification across variables 
- **bod**: Excellent Water Quality is achieved between 1-2.5 mg/L, not above this, but poor water quality can occur at any range of values.
- **disolved oxygen**: Excellent water quality is achieved at narrow specific range of lower values while poor water quality is achieved at very narrow range of higher values circa 50+. 
- **ph**: Clear distinction between poor and excellent values.Higher ph values gviign pooer water quality.
- **temperature**: There is a sweet spot between approx 9 and 12 where excellent quality occurs. Values higher (between approx 13C-18C) or lower (between approx 0C and 7C) lead to poorer water quality.



```{r}
#| fig.cap: "Figure 13: Water Quality variance per variable"
# create  facetted plot of excellent and poor for our data types t
# pipe the data into long format fo plotting specifyign what variables to sue
# and store in wqi_long dataframe
wqi_long <- wqi_dsubset_clean_t_bod %>%
  tidyr::pivot_longer(
    cols = c(ph, temperature, biochem_ox_dmnd, dissolved.oxygen),
    names_to = "parameter",
    values_to = "value"
  )

# keep onyl quality rows that contain v aleu of interest
wqi_subset <- wqi_long %>%
  filter(ccme_wqi %in% c("Excellent", "Poor"))

# create the facetted plot, set transparency to 0.5
ggplot(wqi_subset, aes(x = value, fill = ccme_wqi)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ parameter, scales = "free_x", ncol = 2) +
  theme_minimal() +
  labs(
    title = "Density Distributions of Key Parameters\nfor 'Excellent' and 'Poor' Water Quality Classes",
    x = "Measured Value",
    y = "Density",
    fill = "CCME WQI Category"
  )

```

“Density plots highlight differences in distribution shape between water quality classes. 
- **bod**: Excellent Water Quality is achieved between 1-2.5 mg/L, not above this, but poor water quality can occur at any range of values.
- **disolved oxygen**: Excellent water quality is achieved at lower values while poor water quality is achieved at hgher values circa 50+. This pattern is contrary to the expected ecological relationship between DO and water quality, where higher oxygen typically indicates better conditions. The reversal may reflect dataset-specific factors such as differences between coastal or marine sampling locations, possible sensor calibration differences, or the weighting of dissolved oxygen within the CCME WQI calculation. This variable has also been classied as multimodal earlier. This would require further investigation. 
- **ph**: We see that the higher the ph value the poorer the water quality.
- **temperature**: There is a sweet spot between approx 9 and 12 where excellent quality occurs. Values higher (between approx 13C-18C) or lower (between approx 0C and 7C) lead to poorer water quality




```{r}
#| fig.cap: "Figure 14: Comparative of numebr of variable values for water Quality Index systems"

# Select the relevant columns using the exact names you showed
wqi_compare <- wqi_dsubset_clean_t_bod %>%
  select(waterbodyname, ccme_wqi, wawqi, horton.wqi, brown.wqi, srdd.wqi) %>%
  pivot_longer(
    cols = -waterbodyname,
    names_to = "WQI_System",
    values_to = "Class"
  )

# Plot a comparison of classification counts
ggplot(wqi_compare, aes(x = Class, fill = WQI_System)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Comparison of Water Quality Classifications Across WQI Systems",
    x = "Water Quality Class",
    y = "Number of Sites",
    fill = "WQI System"
  )
```


Below is a table of classification systems and how the same data for oru dataset is classified across each of them.

| Observation | Interpretation|
|----------------|------------|
| **CCME_WQI** | Most values classified as **“Fair”**, very few reach the criteria **Excellent**.|
| **WAWQI**   | Most values classified as**“Good”**  and even spread for others.|
| **Brown_WQI** | Most values classified as**“Good”** and then **Excellent**.|
| **Horton_WQI**| Fairly even classification
| **SRDD_WQI**  | Even distributi acrss top 3 to 4 categories|

There is a large variety of standards. We will continue to use **CCME_WQI**, however using Fair as a more appropriate category for good water quality rather than excellent. as it appears to be very difficul to achieve that standard for Irish waters. This would require further investigation.


## Section 4 : Results

### Section 4.1 Hypothesis 1:

Ho = There is no statistically significant relationship between Biochemical Oxygen Demand (BOD) and temperature
Ha = There is a rstatistically significant elatinoshipt between Biochemical Oxygen Demand (BOD) and temperature

This is a two tailed test, as we do not indicate if is a positive or negaive relationship.

Question 1 > Biochemical Oxygen Demand (BOD) and temperature (both interval / ratio) - Question: Is there a relationship between water pH and temperature?





```{r}

#| label: scatter_temp_bod
#| fig.cap: "Figure 15: Scatterplot of Temperature and Biochemical Oxygen Demand (BOD)"
#| fig.align: center

ggplot(wqi_dsubset_clean_t_bod, aes(x = temperature, y = bod_log10)) +
  geom_point(alpha = 0.5, colour = "steelblue") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  labs(
    title = "Relationship Between Water Temperature and Biochemical Oxygen Demand (BOD)",
    x = "Temperature (°C)",
    y = "Biochemical Oxygen Demand (log₁₀ mg/L)"
  ) +
  theme_minimal()
```

Checking for *Homoscedasticity* and *Heterocedasciticity*. Scatterplot is not exact rectangle or square, but it is not a cone shape, as such we have Homoscedasticity.

Theer is also a very weak linear relationship.


```{r}

#| label: corr_bod_temp
#| fig.cap: "Table 12: Pearson Correlation Between Water Temperature and Biochemical Oxygen Demand (BOD)"

# Pearson correlation between Temperature and BOD (log10)
pearcor_bod_temp <- stats::cor.test(wqi_dsubset_clean_t_bod$temperature, wqi_dsubset_clean_t_bod$bod_log10, method = "pearson")

pearcor_bod_temp


# calculate effect size Using function from effectsize package to convert t to Cohen's d

effectsize::t_to_d(t = pearcor_bod_temp$statistic, pearcor_bod_temp$parameter)
```

The relationship between Biochemical Oxygen Demand (BOD) and temperature was investigated using a Pearson correlation. While a statistically Significant result was found (r =0.1, n=29152. p=<0.01), it is very weak. Cohen's d also indicated a very small effect size (`r effcd`). There is therefore no evidence to reject the null hypothesis in favour of the alternate. We can conclude that there is no statistically significant relationship between Biochemical Oxygen Demand (BOD) and temperature, theresfore the null hypothesis is true and hholds.



### Section 4.2 Hypothesis 1 :


## Appendix


Dataset analysis
```{r}
head(wqi_df)
nrow(wqi_df)
ncol(wqi_df)
str(wqi_df)
```

Missing Data:
  
There are no missing values. No placeholders which may indicate null data is presidential.
As a result of no missing data being present there was no need to run Little's test, to identify MCAR data, or dig subsequently into MAR or MNAR options. Thsi did not come into play as the data set was clean.

```{r}

# Missing Data
# carete a list of valeus that can be opresent in data, to ID them
placeholders <- c("?", "unknown", "none", "na", "n/a", "missing", "null", "NULL", "NaN", "-", "--", " ")

# check if placeholder list avlues above exist in the dataset
# convert all occurrences of the list elements in datframe to NA to standardise missing or misleadingg data values
# and allow yo deal wiht them together cleanly
# sourced from here https://www.rdocumentation.org/packages/naniar/versions/1.1.0/topics/replace_with_na_all
wqi_dsubset_clean <- replace_with_na_all(data = wqi_dsubset, condition = ~ tolower(as.character(.x)) %in% placeholders)


# get summary table of numebrs and percentages for missing non real values 
missing_data_summary <- miss_var_summary(wqi_dsubset)

# View results
print(missing_data_summary)

```

## Future areas of study

water body types
need to d

extreme data values - error or real data
Perisitence over time, persistence in location. so some more detailed geographical knowledge is required.

- Water quality standards,  **CCME_WQI** ,  **WAWQI**   , | **Brown_WQI** ,| **Horton_WQI**, | **SRDD_WQI**


## References
EPA - Summary report - Water qaulity in ireland
https://www.epa.ie/publications/monitoring--assessment/freshwater--marine/EPA-Water-Quality-in-Ireland-Summary-Report-ACC.pdf 

- Curran, West and Finch (1996).
- Field, Miles and Field (2012),
- knitr to create a table of values
# details sbased around examples from here https://www.geeksforgeeks.org/r-language/kable-method-in-r/
- ggplot # create a histogram of ph with density curve over it
# some referene taken from ehre https://stackoverflow.com/questions/5688082/overlay-histogram-with-density-curve
- Water quality standards,  **CCME_WQI** ,  **WAWQI**   , | **Brown_WQI** ,| **Horton_WQI**, | **SRDD_WQI**

