---
title: "MATH9102-W8-RegressionExercise"
author: "Deirdre Lawless"
format: 
  html:
    toc: true      # Enables table of contents
    toc_float: true # Makes the TOC float on the side
    toc_depth: 3   # Specifies the depth of headers to include in TOC
editor: source
---


```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}

quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("foreign")   # For opening SPSS files
quiet_library("lm.beta")   # For extracting standardized co-efficients from regression model
quiet_library("stargazer") # For formatting output of regression
quiet_library("rstatix")   # Effect size calculation plus Post hoc Tukey Test


#Amend this to reflect your folder structure 
datalocation <- file.path("C:","Users", "deirdre.lawless", "datafiles")
datapath <- file.path(datalocation,'sleep.rdata')

#Read in the dataset - sleep dataset from Pallant
sleep<- read.table(datapath, header=TRUE)
```

```{r}
summary(sleep)

```
# Exercise 1: Build a simple regression model
Using the Sleep dataset from Pallant (sleep.sav)build a linear regression model to investigate whether:
Total Sleep and Associated Sensation score (totsas) can be considered to be predicted by a respondent’s level of anxiety (anxiety) (simple linear regression)

## Step 1: Build the model
```{r}
#| label: bmodel1
model1<-lm(totsas~anxiety, data=sleep)#Replace the variables 

```

## Step 2:  Generate the Analysis of Variance Table
This provides the information needed to calculate the F statistic.
```{r}
#| label: anovam1
anova(model1)
```


## Step 3: Generate the summary statistics needed to assess model fit
```{r}
#| label: summm1
summary(model1)
```

## Step 4: Generate Standardized Co-efficients
```{r}
#| label: betam1
lm.beta::lm.beta(model1)
```

## Step 5 (optional: Use Stargazer to produce a tidy output of ANOVA and Summary
```{r}
#| label: stargazerm1
stargazer(model1, type="text") #Tidy output of all the required stats
```

## Step 6: Report the model

# Exercise 2: Multiple Linear Regression - Investigating a differential effect
Extend your previous model to investigate whether a differential effect exists for respondents who report trouble sleeping v those who do not.
Include a dummy variable for trouble sleeping (trubslep)

## Step 1: Build the model
```{r}
#| label: buildm2
model2<-lm(totsas~anxiety+trubslep, data=sleep)

```

## Step 2:  Generate the Analysis of Variance Table
This provides the information needed to calculate the F statistic.
```{r}
#| label: anovam2
anova(model2)
```


## Step 3: Generate the summary statistics needed to assess model fit
```{r}
#| label: summm2
summary(model2)
```

## Step 4: Generate Standardized Co-efficients
```{r}
#| label: betam2
lm.beta::lm.beta(model2)
```

## Step 5 (optional: Use Stargazer to produce a tidy output of ANOVA and Summary
```{r}
#| label: stargazerm2
stargazer(model2, type="text") #Tidy output of all the required stats
```

## Step 6: Report the model

## Step 7: Compare with Simple Linear Regression Model


```{r}
#| label: stargazercomp
#Create the simple regression model
stargazer(model1, model2, type="text") #Quick model comparison
```

# Exercise 3: Multiple Linear Regression - Including an Interactive Term
Extend your previous MLR to include an interaction term -number of caffeine drinks per day (caffeine) and the number of alcoholic drinks per day (alchohol – it is misspelt in the dataset).


## Step 1: Build the model
```{r}
#| label: buildm3
model3<-lm(totsas~anxiety+trubslep+caffeine:alchohol, data=sleep)

```

## Step 2:  Generate the Analysis of Variance Table
This provides the information needed to calculate the F statistic.
```{r}
#| label: anovam3
anova(model3)
```


## Step 3: Generate the summary statistics needed to assess model fit
```{r}
#| label: summm3
summary(model3)
```

## Step 4: Generate Standardized Co-efficients
```{r}
#| label: betam3
lm.beta::lm.beta(model3)
```

## Step 5 (optional: Use Stargazer to produce a tidy output of ANOVA and Summary
```{r}
#| label: stargazerm3
stargazer(model3, type="text") #Tidy output of all the required stats
```

## Step 6: Report the model

## Step 7: Compare with the Regression Model incorporating Differential Effect

```{r}
#| label: stargazercomp1
#Create the simple regression model
stargazer(model2, model3, type="text") #Quick model comparison
```

# Exercise 4: 
Extend your previous MLR model (model2) which used a dummy variable to represent trouble sleeping to include education level (highest education level completed) as a predictor.


## Step 1: Build the model
```{r}
#| label: buildm4

sleep$edlevelf <- factor(sleep$edlevel) # Create a factor from edlevel
sleep$edlevelf <- relevel(sleep$edlevelf, ref = "primary school") # set the reference level to be primary
model4<-lm(totsas~anxiety+trubslep+edlevelf, data=sleep)

```

## Step 2:  Generate the Analysis of Variance Table
This provides the information needed to calculate the F statistic.
```{r}
#| label: anovam4
anova(model4)
```


## Step 3: Generate the summary statistics needed to assess model fit
```{r}
#| label: summm4
summary(model4)
```

## Step 4: Generate Standardized Co-efficients
```{r}
#| label: betam4
lm.beta::lm.beta(model4)
```

## Step 5 (optional: Use Stargazer to produce a tidy output of ANOVA and Summary
```{r}
#| label: stargazerm4
stargazer(model4, type="text") #Tidy output of all the required stats
```

## Step 6: Report the model

## Step 7: Compare with the Regression Model incorporating Trouble Sleeping Differential Effect

```{r}
#| label: stargazercomp2
#Create the simple regression model
stargazer(model2, model4, type="text") #Quick model comparison
```