---
title: "MATH9102-W4-ttest"
author: "Deirdre Lawless"
format: 
  html:
      self-contained: true
      code-fold: true
editor: visual
---

```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}
quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("psych")   # Descriptive statistics by group
quiet_library("car")   # Includes Levene's test for homogeneity
quiet_library("effectsize")   # Includes calculations for effectsize quiet_library("pastecs")    # Tools for descriptive statistics, time series analysis, and data exploration (e.g., stat.desc)
quiet_library("FSA")        # Includes general tools for data analysis (e.g., nonparametric tests, bootstrapping)
quiet_library("semTools")   # Supplementary functions for Structural Equation Modeling (SEM), including model fit, reliability, and measurement invariance test


#Amend this to reflect your folder structure 
datalocation <- file.path("./")
datapath <- file.path(datalocation,'survey.dat')

#Read in the dataset
survey <- read.table(datapath)

#Setting the column names to be that used in the dataset but in lowercase to make life a bit easier
colnames(survey) <- tolower(colnames(survey))
```

# Investigate Difference in Perceived Stress Levels for Those With or Without Children

This analysis investigate the following hypothesis:

H0: There is no difference in perceived stress scores for respondents who have children and respondents who do not have children HA: There is a difference in perceived stress scores for respondents who have children and respondents who do not have children

The sample consisted of respondents who either had children or did not. Out of the total sample, 42.24% reported having children, while 57.76% did not have children, see Table 1 and Figure 1.

```{r}
# Create a frequency table for the numbers in each group 
# create table from child alue - just inlcude frequencies
child_freq <- table(survey$child)

# Percentage table
# her calculating the proportions - workign ut the %s
child_perc <- prop.table(child_freq) * 100

# Combine ing both  - joining frequencies and percentages from above
child_summary <- cbind(Frequency = child_freq, Percentage = round(child_perc, 2))
```

```{r}
#| fig.cap: "Table 1: Contingency Table Respondents Children Status"
# Print the combined table with frequencies and percentages
print(child_summary)

```

```{r}
# Filter out NA values from the 'child' variable only
# get rid of NAs
# already doen missing values analysis  - ctaete a bar chart to show that
cleaned_survey <- survey %>%
  filter(!is.na(child))
```

```{r}
#| fig.cap: "Figure 1: Bar Chart Showing Number of Respondents By Children Status"
# Base R bar chart
barplot(table(survey$child), 
        main = "Distribution Children Status", 
        xlab = "Children (YES/NO)", 
        ylab = "Count", 
        col = c("skyblue", "orange"))
```

```{r}
#| fig-cap: "Figure 2: Histogram of Total Perceived Stress"
#Create histogram
gs <- ggplot(survey, aes(x=tpstress)) 
gs <- gs + labs(x="Total Perceived Stress")
gs <- gs + geom_histogram(binwidth=2, colour="black", aes(y=..density.., fill=..count..))
gs <- gs + scale_fill_gradient("Count", low="#DCDCDC", high="#7C7C7C")
gs <- gs + stat_function(fun=dnorm, color="red",args=list(mean=mean(survey$tpstress, na.rm=TRUE), sd=sd(survey$tpstress, na.rm=TRUE)))
gs
```

```{r}
#| fig.cap: "Figure 3: QQPlot for Total Perceived Strees"
#Create a qqplot
qqnorm(survey$tpstress)
qqline(survey$tpstress, col=2) #show a line on theplot
```

```{r}
#Generate regular summary statistics 
pastecs::stat.desc(survey$tpstress, basic=F)

#Generate standardized scores for skewness and kurtosis
tpskew<-semTools::skew(survey$tpstress)
tpkurt<-semTools::kurtosis(survey$tpstress)
tpskew[1]/tpskew[2]
tpkurt[1]/tpkurt[2]

#Calculate the percentage of standardized scores outside acceptable limits
ztpstress<- abs(scale(survey$tpstress))

FSA::perc(as.numeric(ztpstress), 1.96, "gt")
FSA::perc(as.numeric(ztpstress), 3.29, "gt")
# Below we are reporting the normality assessment - embedding code within a paragraph to get the statistics generated ( you can just write it out)

```

Total Perceived Stress scrires were assessed for normality. Visual inspection of the histogram and QQ-Plot (see Figure 2 and Figure 3) identified some issues with skewness and kurtosis. While the standardized score for kurtosis (`r round(tpkurt[1]/tpkurt[2],2)`) could be considered acceptable using the criteria proposed by Curran, West and Finch (1996), but the standardized score for skewness (`r round(tpskew[1]/tpskew[2],2)`) was outside the acceptable range. However as over 99% of standardized scores fall within the bounds of +/- 3.29, using the guidance of Field, Miles and Field (2012), the data can be considered to approximate a normal distribution (m=`r round(mean(survey$tpstress, na.rm=TRUE),2)`, sd=`r round(sd(survey$tpstress, na.rm=TRUE),2)`, n=`r length(survey$tpstress)-sum(is.na(survey$tpstress))`).

```{r}
#| fig.cap: "Figure 5: Histogram of Total Perceivced Stress Grouped by Repondent's Children Status"

# Create a separate histogram showing the distribution for the two (potentially) separate populations (the groups)
# Create the plot with both histogram and density plot
ggplot(cleaned_survey, aes(x=tpstress, color=child, fill=child))  + 
  labs(x="Total Perceived Stress") + 
  # Histogram with alpha for transparency
  geom_histogram(aes(y = ..density..), position="identity", alpha=0.5, binwidth=1, color="black") +
  # Density plot with transparency for overlapping curves
  geom_density(alpha=0.4) +
  # Customize the color and fill scales
  scale_color_manual(values=c("blue", "red"), labels=c("No Children", "Children")) +
  scale_fill_manual(values=c("lightblue", "lightcoral"), labels=c("No Children", "Children"))

```

```{r}
#| fig.cap: "Table 2: Descriptive Statistics of Total Perceived Stress Grouped by Respndent's Children Status"
#Get descriptive stastitics by group - output as a matrix
psych::describeBy(survey$tpstress, survey$child, mat=TRUE)

```

## Differences - Parametric Tests

## Independent t-test

```{r}
#Conduct Levene's test for homogeneity of variance in library car - the null hypothesis is that variances in groups are equal so to assume homogeneity we would expect probability to not be statistically significant.
car::leveneTest(tpstress ~ child, data=survey)
#Pr(>F) is your probability - in this case it is not statistically significant so we can assume homogeneity


#Conduct the t-test from package stats - basic r package. Prefixing it to make sure that is the one used rather than t-test from another package.

#In this case we can use the var.equal = TRUE option to specify equal variances and a pooled variance estimate
stats::t.test(tpstress~child,var.equal=TRUE,data=survey)
#No statistically significant difference was found

res <- stats::t.test(tpstress~child,var.equal=TRUE,data=survey)
#Calculate Cohen's d
#artithmetically
effcd=round((2*res$statistic)/sqrt(res$parameter),2)

#Alternate approach -Using function from effectsize package to convert t to Cohen's d
effectsize::t_to_d(t = res$statistic, res$parameter)

#Calculate eta squared
effes=round((res$statistic*res$statistic)/((res$statistic*res$statistic)+(res$parameter)),3)
effes

```

### Reporting the results with Cohen's d effect

An independent-samples t-test was conducted to compare perceived stress scores for respondents who have children and those who do not. Levene’s test for equality of variances was not statistically significant, indicating that the assumption of homogeneity of variance was met (p \> .05). No significant difference in the scores for self-esteem was found (M=33.20, SD= 5.79, n=183 for respondents who have children, M= 33.64, SD= 5.30, n=249 for respondents who do not), (t(`r res$parameter`)= `r round(res$statistic,3)`, p = `r round(res$p.value,2)`. Cohen's d also indicated a very small effect size (`r effcd`).

### Reporting the results with eta squared effect

An independent-samples t-test was conducted to compare perceived stress scores for respondents who had children and those who do not. Levene’s test for equality of variances was not statistically significant, indicating that the assumption of homogeneity of variance was met (p \> .05). No significant difference in the scores for self-esteem was found (M=33.20, SD= 5.79 for respondents who have children, M= 33.64, SD= 5.30 for respondents who do not), (t(`r res$parameter`)= `r round(res$statistic,3)`, p = `r round(res$p.value,2)`. A very small effect size was also indicted by the eta squared value (`r effes`).
