---
title: "MATH9102-W4-Nominal"
author: "Deirdre Lawless"
format: 
  html:
      self-contained: true
      code-fold: true
editor: visual
---

```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```

This analysis is being conducted on a dataset accompanying Quantiative Data Analysis in Education a Critical Introduction Using SPSS by Paul Connolly available at https://www.routledge.com/Quantitative-Data-Analysis-in-Education-A-Critical-Introduction-Using-SPSS/Connolly/p/book/9780415372985. The dataset consists of a small number of variables selected from the Young Life and Times Survey (2005) that consisted of a random sample of 16 year olds in Northern Ireland (n=819). The survey runs annually and covers a wide variety of topics related to young people's attitudes and social activities. The variables selected for this dataset focus specifically on the young people's experiences of bullying in school.

This analysis investigates the question: “Is there a relationship between being bullied in school and participating in bullying of others in school?”. The associated hypothesis pair to be tested is: H0: There is no relationship between being a victim of bullying in school and participating in bullying others in school. HA: There is a relationship between being a victim of bullying in school and participating in bullying others in school.

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}
quiet_library("tidyverse") 
quiet_library("sjstats")  
quiet_library("gmodels")


#Amend this to reflect your folder structure 
datalocation <- file.path("./")
datapath <- file.path(datalocation,'bullying.dat')

#Load the data into a dataframe
bully<-read.table(datapath, header=TRUE)
```

## Approach 1 - use the Crosstable function from gmodels

```{r}
#CrossTable(predictor, outcome, fisher = TRUE, chisq = TRUE, expected = TRUE)
#Note: for this approach there must be the same number of values for both variables so
#missing data is important. 

# Remove rows where either variable is NA
bully_clean <- bully %>%
  filter(!is.na(ubullsch) & !is.na(ubulloth))

# Now run CrossTable requesting both Fisher and Chi Square so we have full information
# We request the format in SPSS to make it easier to interpret.
gmodels::CrossTable(
  bully_clean$ubullsch,
  bully_clean$ubulloth,
  fisher = TRUE,
  chisq = TRUE,
  expected = TRUE,
  sresid = TRUE,
  format = "SPSS"
)

```

## Approach 2: Create your own contingency table and use the chi square test from the stats package

```{r}
#Alternate approach: Create your own contingency table
mytable<-xtabs(~ubullsch+ubulloth, data=bully)
#You can use the chisq.test function in the stats package to do the test
#set the flag correct=TRUE to get Yates correction needed for 2x2 table
ctest<-stats::chisq.test(mytable, correct=TRUE)#chi square test

ctest#will give you the details of the test statistic and p-value
ctest$expected#expected frequencies
ctest$observed#observed frequencies
ctest$p.value
#Calculate effect sizes
sjstats::phi(mytable)
sjstats::cramer(mytable)

```

# Report your findings

A Chi-square test of independence was conducted to examine the relationship between being bullied in school and bullying others. The association was statistically significant (χ²(1, N = 799) = 22.23, p \< .001). Students who reported being bullied were significantly more likely to report bullying others (14.3%) than those who were not bullied (4.5%). The effect size, measured by Phi (φ = 0.17), indicates a small association. There is therefore evidence to support rejecting the null hypothesis in favour of the alternate hypothesis.

```{r}
# We report phi here becasue we are looking at a 2x 2 comparison
```

### For repeated measures for categorical variables use McNemars test mcnemars.test
