---
title: "MATH9102-W4-NonParam"
author: "Deirdre Lawless"
format: 
  html:
      self-contained: true
      code-fold: true
editor: visual
# same as parametric 
---

```{=html}
<style>
/* Center captions for figures and tables in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption,
figure > figcaption,
.table-caption,
table > caption {
  text-align: center;
  color: black;
  font-weight: bold;
}
</style>
```

!\<--# same as parametric --\>

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}
quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("psych")   # Descriptive statistics by group
quiet_library("car")   # Includes Levene's test for homogeneity
quiet_library("effectsize")   # Includes calculations for effectsize quiet_library("pastecs")    # Tools for descriptive statistics, time series analysis, and data exploration (e.g., stat.desc)
quiet_library("FSA")        # Includes general tools for data analysis (e.g., nonparametric tests, bootstrapping)
quiet_library("semTools")   # Supplementary functions for Structural Equation Modeling (SEM), including model fit, reliability, and measurement invariance test
quiet_library("coin") #	To calculate Z, to get rosenthal's r, we can use the Wilcox test from the ‘coin ‘package



#Amend this to reflect your folder structure 
datalocation <- file.path("./")
datapath <- file.path(datalocation,'Field-BDI-Non-parametric.csv')

#Load the data into a dataframe - it is a .csv file, we indicate that the first row is a header with colnames
drinkset<- read.csv(datapath, header=TRUE)
#Create data subsets for each drink
# dividin data into 2 subsets to make it easy for us when we're doing histograms
drinkxdata <- subset(drinkset, drink=='DrinkX')
drinkydata <-subset(drinkset, drink=='DrinkY')
```

# Differences in Depression Levels - Sunday

```{r}
#Note: I am not doing the inspection of the variables here as I know they are not normal since this data has been designed to illustrate non-parametric difference

#Get your descriptive statistics - we are using the by function to handle the need to group
by(drinkset$bdisun,drinkset$drink, median)
by(drinkset$bdisun,drinkset$drink, IQR)

#Create plots of these
# putting in density functin in red
#and removing missin g data
gs <- ggplot(drinkxdata, aes(x=bdisun))
gs <- gs + labs(x="Drink X")
gs <- gs + geom_histogram(binwidth=2, colour="black")
gs <- gs + scale_fill_gradient("Count", low="#DCDCDC", high="#7C7C7C")
gs <- gs + stat_function(fun=dnorm, color="red",args=list(mean=mean(drinkxdata$bdisun, na.rm=TRUE), sd=sd(drinkxdata$bdisun, na.rm=TRUE)))

# histogram has bin width issue, is entirely presentational 
# if put in a  value of '1' it will lok closer together
# I don't have an issue with what it looks like once the information is there
# its purely presentational
```

```{r}
#| fig.cap: "Figure 1: Histogram of Depression Levels for Drink X on Sunday"
gs
```

```{r}
# do the same for drink y
gs <- ggplot(drinkydata, aes(x=bdisun))
gs <- gs + ggtitle ("Depression Levels Sunday")
gs <- gs + labs(x="Drink Y")
gs <- gs + geom_histogram(binwidth=2, colour="black")
gs <- gs + scale_fill_gradient("Count", low="#DCDCDC", high="#7C7C7C")
gs <- gs + stat_function(fun=dnorm, color="red",args=list(mean=mean(drinkydata$bdisun, na.rm=TRUE), sd=sd(drinkydata$bdisun, na.rm=TRUE)))
```

```{r}
#| fig.cap: "Figure 2: Histogram of Depression Levels for Drink Y on Sunday"
gs
```

```{r}

# Test for differences on  Sunday to get our U statistic (reported as W)
stats::wilcox.test(bdisun~drink, data=drinkset) 
#by default wilcox.test() calculates an exact p-value if the samples contain less than 50 finite values and there are no ties. Otherwise, a normal approximation is used. In our example, the sample contains less than 50 finite values but there are ties. That’s why the warning message “cannot compute exact p-value with ties” is appearing.


#To calculate Z we can use the Wilcox test from the coin package 
coin::wilcox_test(bdisun~as.factor(drink), data=drinkset)

#Calculate the effect size R
reff=rstatix::wilcox_effsize(bdisun~drink, data=drinkset)

#Calculate the effect r
reff$effsize

```

## Reporting the results of the Sunday Finding

A Mann–Whitney test was used to test whether depression levels differed between those who drank DrinkX (Mdn = 17.50, IQR = 3.75) and those who drank DrinkY (Mdn = 16.00, IQR = 3.00) the day after the drinks were taken (Sunday). The results showed no significant difference (U = 64.5, z = 1.11, p = .29, r = .25).

#Note: We are reporting the W statistic output by the stats::wilcox_test as U here.

# Differences in Depression Levels - Wednesday

```{r}
#Note I am not doing the inspection of the variables here as I know they are not normal since this data has been designed to illustrate non-parametric difference

# doing the same thign for wed

#Get your descriptive statistcs
by(drinkset$bdiwed,drinkset$drink, median)
by(drinkset$bdiwed,drinkset$drink, IQR)
```

```{r}
#Create plots of these
gs <- ggplot(drinkxdata, aes(x=bdiwed))
gs <- gs + ggtitle ("Depression Levels Wednesday")
gs <- gs + labs(x="Drink X")
gs <- gs + geom_histogram(binwidth=2, colour="black")
gs <- gs + scale_fill_gradient("Count", low="#DCDCDC", high="#7C7C7C")
gs <- gs + stat_function(fun=dnorm, color="red",args=list(mean=mean(drinkxdata$bdiwed, na.rm=TRUE), sd=sd(drinkxdata$bdiwed, na.rm=TRUE)))
```

```{r}
#| fig.cap: "Figure 3: Histogram of Depression Levels for Drink X on Wednesday"
gs
```

```{r}

gs <- ggplot(drinkydata, aes(x=bdiwed))
gs <- gs + ggtitle ("Depression Levels Wednesday")
gs <- gs + labs(x="Drink Y")
gs <- gs + geom_histogram(binwidth=2, colour="black")
gs <- gs + scale_fill_gradient("Count", low="#DCDCDC", high="#7C7C7C")
gs <- gs + stat_function(fun=dnorm, color="red",args=list(mean=mean(drinkydata$bdiwed, na.rm=TRUE), sd=sd(drinkydata$bdiwed, na.rm=TRUE)))
```

```{r}
#| fig.cap: "Figure 4: Histogram of Depression Levels for Drink Y on Wednesday"
gs
```

```{r}
# Test for differences on  Wednesday to get our U statistic (reported as W)
stats::wilcox.test(bdiwed~drink, data=drinkset) 
#by default wilcox.test() calculates an exact p-value if the samples contain less than 50 finite values and there are no ties. Otherwise, a normal approximation is used. In our example, the sample contains less than 50 finite values but there are ties. That’s why the warning message “cannot compute exact p-value with ties” is appearing.

#To calculate Z we can use the Wilcox test from the coin package 
coin::wilcox_test(bdiwed~as.factor(drink), data=drinkset)

#Calculate the effect size R
reff=rstatix::wilcox_effsize(bdiwed~drink, data=drinkset)

#Calculate the effect r
reff$effsize

```

## Reporting the comparison

However, a Mann–Whitney test used to compare depression levels three days after the drinks were taken (Wednesday) indicated that those who drank DrinkX (Mdn = 33.50, IQR = 6.75) were significantly more depressed than those who drank DrinkY (Mdn = 7.50, IQR = 3.75) (U = 96.00, z = 3.48, p \< .001, r = .78).
