---
title: "MATH9102-Phase2-Task4"
format:
    html:
      self-contained: true
      toc: true      # Enables table of contents
      toc_float: true # Makes the TOC float on the side
      toc_depth: 3   # Specifies the depth of headers to include in TOC
---


# Introduction
Complete the PCA making relevant decisions and report your findings.
https://www.kaggle.com/datasets/hayriyigit/website-satisfaction-survey


These are the variables of interest to this task:

***Navigation:***

| **Variable** | **Meaning**                    |
| ------------ | ------------------------------ |
| **q1**       | Ease of navigating the website |
| **q2**       | Clarity of the menu/structure  |
| **q3**       | Ease of finding information    |
| **q4**       | Website is intuitive to use    |
| **q5**       | Layout helps navigation        |


***Typography:***

| **Variable** | **Meaning**                     |
| ------------ | ------------------------------- |
| **q9**       | Text readability and formatting |

***Colour:***

| **Variable** | **Meaning**                                   |
| ------------ | --------------------------------------------- |
| **q7**       | Colour scheme suitability                     |
| **q8**       | Layout attractiveness (partly colour-related) |

***Interactivity:***

| **Variable** | **Meaning**                                                   |
| ------------ | ------------------------------------------------------------- |
| **q15**      | Website loading speed (affects perceived interactivity)       |
| **q16**      | Reliability / absence of errors (supports smooth interaction) |
| **q17**      | Responsiveness on device (directly interactive behaviour)     |


**Note:** Throughout the code, placeholders have been included for some steps e.g. step 5 retains 8 components, step 6 uses all the data. However, you need to make decisions based on your interpretation of the outcomes of the relevant steps.

**WHAT YOU NEED TO DO:**

- **Step 2:** Review the outcomes and write a report on the suitability of the dataset for dimension reduction.

- **Step 3 and Step 4:** Review the outcomes and decide how many components you wish to retain. Adjust any visualizations you need to.

- **Step 5:** Adjust the code to retain the required number of components and apply rotation. Write a report on the outcome of the PCA.

- **Step 6:** Complete the code to conduct the reliability analysis. Write the report on the reliability of the data.

- **Conclusion:** Write a conclusion on the success and usefulness (or otherwise) of the PCA.


```{r echo=FALSE, include=FALSE}
# Set CRAN mirror for non-interactive sessions
options(repos = c(CRAN = "https://cloud.r-project.org"))

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}
# tidyverse: For data manipulation (dplyr), wrangling (tidyr), and plotting (ggplot2),
# essential for preparing datasets before running PCA/EFA.
quiet_library("tidyverse")

# psych: Core package for PCA, EFA, reliability analysis (Cronbach's alpha),
# factor extraction, rotation, and interpreting loadings.
quiet_library("psych")

# REdaS: Provides KMO (Kaiser-Meyer-Olkin) and Bartlett's test functions
# used to assess whether the dataset is suitable for dimension reduction.
quiet_library("REdaS")

# Hmisc: Contains correlation tools and functions for handling missing data,
# useful for computing correlation matrices required for PCA/EFA.
quiet_library("Hmisc")

# corrplot: Produces visual correlation matrix heatmaps,
# helping identify clusters of variables and collinearity patterns.
quiet_library("corrplot")

# ggcorrplot: ggplot-based correlation matrix visualisation,
# offering cleaner, publication-ready heatmaps.
quiet_library("ggcorrplot")

# factoextra: Visualisation tools for PCA and EFA outputs,
# including scree plots, biplots, eigenvalue charts, and clustering plots.
quiet_library("factoextra")

# nFactors: Helps determine the number of factors/components to retain
# using methods like parallel analysis and eigenvalue comparisons.
quiet_library("nFactors")

# knitr: Enables clean rendering of tables, outputs, and formatted results
# inside R Markdown or Quarto reports.
quiet_library("knitr")

```


```{r echo=FALSE, include=FALSE}
#| label: loaduxData
datalocation <- file.path("C:","Users", "deirdre.lawless", "datafiles")
# Load the dataset (update the path as needed)
datapath <- file.path(datalocation,'WebSatisfaction.csv')
websat <- read.csv(datapath)

# Extract all relevant UX attribute variables into one dataframe
uxData <- websat[, c(
  # Navigation
  "q1", "q2", "q3", "q4", "q5",
  
  # Typography
  "q9",
  
  # Colour
  "q7", "q8",
  
  # Interactivity
  "q15", "q16", "q17"
)]

# Check structure
str(uxData)

```


# STEP 1: Create and screen the correlation matrix for the uxData dataset
## Step 1a: Create the matrix
You can change this to use another approach if you wish.

```{r corscreen}
#create a correlation matrix  
uxMatrix<-cor(uxData)
round(uxMatrix, 2)
```

All the variables correlate with at least one other at .3 or above (absolute values) and nothing above .8:

| Variable | Highest Correlation | Meets ≥ .30 Rule  | Notes                    |
| -------- | ------------------- | ----------------- | ------------------------ |
| **q1**   | 0.73 (with q2)      |  Yes             | Strong                   |
| **q2**   | 0.74 (with q4)      |  Yes             | Strong                   |
| **q3**   | 0.65 (with q4)      |  Yes             | Strong                   |
| **q4**   | 0.74 (with q2)      |  Yes             | Strong                   |
| **q5**   | 0.61 (with q15)     |  Yes             | Strong                   |
| **q9**   | **0.58 (with q16)** | ️ Yes            |  Strong|
| **q7**   | 0.67 (with q17)     | ️ Yes            | Strong                   |
| **q8**   | 0.64 (with q17)     | ️ Yes            | Strong                   |
| **q15**  | 0.66 (with q17)     | ️ Yes            | Strong                   |
| **q16**  | 0.58 (with q9)      | ️ Yes            | Strong                   |
| **q17**  | 0.67 (with q7)      | ️ Yes            | Strong                   |



## Step 1b: Visualize the correlation matrix
You can change this to use another visualizatin if you wish.
```{r }
#| fig.cap:  "Figure 1: Basic Correlation Matrix"

# 1. Basic correlation matrix plot
ggcorrplot::ggcorrplot(uxMatrix)
```

```{r}
#| fig.cap: "Figure 1: Correlation Matrix - uxData Dataset"
# 5. Correlation matrix showing coefficients with labels
# Note: This can become cluttered with a high number of variables
#| fig-width: 16
#| fig-height: 16

ggcorrplot::ggcorrplot(
  uxMatrix,
  lab = TRUE,
  lab_size = 2.5,
  tl.cex = 8,
  tl.srt = 45,
  type = "lower"
)

```

# STEP 2: Check if the data is suitable for dimension reduction (there is enough variance of the right kind)

You want to do the following:

  * Check that your variables are correlated by doing Bartlett's test of Sphericity.
    + Check that the result is statistically significant.
  
  * Check that variables are not too highly correlated for PCA.
    + Calculate and inspect the determinant.
    + Check to see if it is greater than 0.00001.
  
  * Check that there is enough variance in your variables that could be caused by an underlying factor/dimension.
    + Use the Kaiser-Meyer-Olkin Measure of Sampling Adequacy (MSA) test.
    + High values close to 1 suggest that PCA/FA might be useful.
    + Values over 0.8 or over are considered strong.
    + Anything less than 0.5 suggests that PCA/FA won’t be useful.
    + 0.00 to 0.49 unacceptable, 0.50 to 0.59 miserable, 0.60 to 0.69 mediocre ,0.70 to 0.79 middling, 0.80 to 0.89 meritorious, 0.90 to 1.00 marvellous.

## Step 2a: Check Bartlett's test
```{r}
psych::cortest.bartlett(uxData)
```


## STEP 2b: Check the (KMO)

```{r}
#KMO 
REdaS::KMOS(uxData)
```


## STEP 2c:  Check the Determinant

```{r}
#Determinant:
det(uxMatrix)

```

# YOUR TASK: Write the Report on the Suitability of the Data for Dimension Reduction




# STEP 3: Execute the PCA
In this step, run the PCA using all of the remaining variables in the uxData dataset. The purpose here is simply to let R extract all possible components so we can see the complete structure of the data. You should focus on the loading matrix that is produced. 
A loading tells you how strongly each variable relates to a component. 
At this stage you are not trying to interpret all components; instead, you are trying to understand the overall structure so that we can make decisions in the next step.


```{r}
#do the reduction
pc1 <-  principal(uxData, nfactors = length(uxData), rotate = "none")#Conduct the PCA extracting the same number of components as there are variables

pc1#output all details of the PCA

```

# STEP 4:  Decide on which components to retain
Once the full PCA is run, you need to decide how many components to keep for interpretation. There are two main things to look at:

The scree plot – This helps you visually identify where the “elbow” or point of inflexion is. After this point, the components explain very little additional variance.

Eigenvalues – A common guideline is to keep components with eigenvalues greater than 1 (Kaiser’s criterion) or greater than 0.7 (Joliffe's criterion).


```{r}
#Create the scree plot
plot(pc1$values, type = "b") 

#Print the variance explained by each component
pc1$Vaccounted 

#Print the Eigenvalues
pc1$values


#Another way to look at the eigenvalues plus variance explained (need to use princomp function of PCA to get right class for use with factoextra functions)
pcf=princomp(uxData)
factoextra::get_eigenvalue(pcf)
factoextra::fviz_eig(pcf, addlabels = TRUE, ylim = c(0, 50))#Visualize the Eigenvalues
factoextra::fviz_pca_var(pcf, col.var = "black")
factoextra::fviz_pca_var(pcf, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE # Avoid text overlapping
             )

#Print the loadings above the level of 0.3
psych::print.psych(pc1, cut = 0.3, sort = TRUE)

#Create a diagram showing the components and how the manifest variables load
fa.diagram(pc1) 

#Show the loadings of variables on to components
fa.sort(pc1$loading)
```


Communalities - tell us how much of each variable’s variance is captured by the components we keep.
A communality close to 1 means the retained components represent that variable very well.
A communality below about 0.4 may indicate that the variable is not well represented.


```{r}
#Output the communalities of variables across components (will be one for PCA since all the variance is used)
pc1$communality 
#Visualize contribution of variables to each component
var <- factoextra::get_pca_var(pcf)
corrplot::corrplot(var$contrib, is.corr=FALSE)
```

***If you want to visualize the contributions of the variables to each component use this code**
```{r}
# Contributions of variables to PC?
factoextra::fviz_contrib(pcf, choice = "var", axes = 8, top = 10)# Replace the value for axes with the number of the component e.g. 1 for PC1, 2 for PC2, 8 for PC8

```


# YOUR TASK: Decide how many components to retain and complete the code for Step 5

# STEP 5: Repeat with reduced components and rotation

***You must:***
- Decide how may components to retain
- Re-run the PCA asking for only those components
- Apply a rotation — usually varimax when doing PC

The rotated solution gives you a clearer picture of the underlying dimensions in the dataset, and this is the version you would normally interpret.

```{r}
#Apply rotation to try to refine the component structure- you decide how may
pc2 <-  principal(uxData, nfactors = 8, rotate = "varimax") #replace the value of nfactors with the number of components to retain
#output the components
psych::print.psych(pc2, cut = 0.3, sort = TRUE)
#output the communalities
pc2$communality
#NOTE: you can do all the other things done for the model in step 4 if you want to

```



# YOUR TASK: Report the outcomes of the PCA


# YOUR TASK: Complete the Code for Reliability Analysis

# STEP 6: Reliability Analysis

Check whether the variables within each proposed group consistently measure the same underlying construct.

Cronbach's alpha is used to evaluate internal consistency:
   α > .80 = good
   α ~ .70 = acceptable
   α < .60 = poor

 Note: Dimension reduction does NOT guarantee reliability.
 A component/factor may be statistically valid but not reliable as a scale.
 If we are just trying to reduce the number of variables we may still be able to use. If however we want to use the component/factor as a reliable measure of a concept it will cause use problems if it is not considered reliable.
 

```{r}
#If you know that certain sections should be treated as groups, test these separately
#Output Cronbach alpha values - use pysch::alpha

# Compute Cronbach's Alpha for the entire dataset
psych::alpha(uxData)   
#Note: You should break this out into components.

```

# YOUR TASK: Report the outcomes of the Reliability Analysis


# YOUR TASK: Write a Summary Conclusion on the PCA


