---
title: "TU059/TU060/TU256/PhD PSI — Week 2: Creating Basic Plots in R"
author: "Deirdre Lawless"
date: 2025-08-22
format: html
---



# 1. Install and load packages

This first bit of code installs all the packages needed to run the various functions within this file.
We define a helper `quiet_library()` that installs (if needed) and loads a package quietly.

```{r}
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}

quiet_library("tidyverse")    # includes ggplot2, readr, dplyr, etc.
quiet_library("readr")        # explicit (for read_delim), though also in tidyverse

#Amend this to reflect your folder structure 
datalocation <- file.path("C:","Users", "deirdre.lawless", "datafiles")
#On Mac this would be datalocation <- file.path("/Users", "deirdre.lawless", "datafiles")

```



# 2. Reminder of some basics
A **dataframe** is a structure into which we load data from a file. It is a collection of rows and columns. 
We refer to the columns as **variables**. We can find out the names of all variables in the data frame using the **names** command.
We can reference a particular variable using the **$** symbol.
R is case sensitive so it will consider the variable salary to be different to a variable Salary or SALARY.




# 3. Some Basics with ggplot
Using the data file facebookNarcissim that contains data from a study that looked at ratings of Facebook profile pictures which were rated (on coolness, fashion, attractiveness and glamour, each of which is assigned a number). The intent was to try to predict (by building a statistical model) how high the person in the profile picture rated on a narcissism scale.

```{r loaddata}
datapath <- file.path(datalocation,'FacebookNarcissism.dat')

facebookData <- read.delim(datapath, header=TRUE)

#| label: load-data
#| message: false
#| warning: false

# Peek at columns available (uncomment if needed)
# names(facebookData)

```



## 3.1 Creating a basic ggplot
We are now going to step through a series of ggplot commands to create some basic plots.

First create the graph (pass the dataset to ggplot and we map the aesthetics (aes(x,y)) to use the variables for the narcissim scale total value (NPQC_R_Total) and the rating variable).
Then setup to use points and a title.
You can find more about aes at https://ggplot2.tidyverse.org/articles/ggplot2-specs.html

```{r}
#For each rating, show the values of NPQC_R_Total appearing in the dataset
graph <- ggplot(facebookData, aes(NPQC_R_Total, Rating))
graph + geom_point() + ggtitle("geompoint")
```



## 3.2 Changing the shape used
Changes the graph to use triangles
You can find the numbers to use for different shapes at this link http://www.sthda.com/english/wiki/ggplot2-point-shapes#point-shapes-in-r

```{r}
graph + geom_point(shape = 2) 

```



## 3.3 Change plot to use a theme and change the title to reflect that
```{r}
#using a theme theme, using Triangles
#you find more themes at https://ggplot2.tidyverse.org/reference/ggtheme.html

graph + theme_dark() + geom_point(shape = 2) + ggtitle ("theme_dark")

```

## 3.4. Change plot to us different colours for each rating type and change the title to reflect this
```{r}
graph + geom_point(aes(colour = Rating_Type)) + ggtitle("geom_point(aes(colour = Rating_Type))")
```


## 3.5 Change plot to add Jitter 
Jitter adds a small amount of random variation to the location of each point 
This spreads the data out
```{r}
graph + geom_point(aes(colour = Rating_Type), position = "jitter") + ggtitle ("geom_point(aes(colour = Rating_Type), position = jitter)")
```



## 3.6 Save a plot as an image
```{r}
ggsave("Week2 Example jitter.png", plot=last_plot())
```

## +-------------------------------END OF FILE ---------------------+