---
title: "MATH9102-W11-LogReg-Exercise"
author: "Deirdre Lawless"
format: 
  html:
    self-contained: true
    toc: true      # Enables table of contents
    toc_float: true # Makes the TOC float on the side
    toc_depth: 3   # Specifies the depth of headers to include in TOC
editor: source
---



```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}

/* Center and bold table captions, force them below tables */
.table-caption, 
caption {
  caption-side: bottom;  /* caption below table */
  text-align: center;
  color: black;
  font-weight: bold;
}
</style>
```

```{r echo=FALSE, include=FALSE}
# Set CRAN mirror for non-interactive sessions
options(repos = c(CRAN = "https://cloud.r-project.org"))

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}
# Data manipulation and plotting (includes dplyr, ggplot2, tidyr, etc.)
quiet_library("tidyverse")

# Epidemiology tools, including ROC curves used in logistic regression diagnostics
quiet_library("Epi")

# Effect size statistics, pseudo R-squared for logistic regression models
quiet_library("DescTools")

# Produces nicely formatted model output tables
quiet_library("stargazer")

# Likelihood ratio tests and other model comparison tools
quiet_library("lmtest")

# Variance Inflation Factor (VIF) for checking multicollinearity
quiet_library("car")

# Hosmer–Lemeshow goodness-of-fit test for logistic regression
quiet_library("generalhoslem")

# Classification diagnostics (e.g., collinearity checks, model assumptions)
quiet_library("regclass")

# Confusion matrices and classification metrics (sensitivity, specificity, etc.)
quiet_library("yardstick")

# Quick and tidy frequency tables for descriptive statistics
quiet_library("janitor")

# Additional model performance measures, including pseudo R-squared
quiet_library("performance")

# Extracts tidy model coefficients, odds ratios, and confidence intervals
quiet_library("parameters")

#Amend this to reflect your folder structure 
datalocation <- file.path("./")
datapath <- file.path(datalocation,'sleep.rdata')

sleepdf <- read.table(datapath)



sleepdf <- sleepdf %>%
  mutate(
    trubslep_num = case_when(
      trubslep == "yes" ~ 1,
      trubslep == "no"  ~ 0,
      TRUE ~ NA_real_
    )
  )

# remove NAs
sleepdf <- sleepdf %>% 
  filter(!is.na(trubslep_num) & !is.na(smoke) & !is.na(caffeine))

#Check your proportions of the outcome variable for bias - are these representative?
table(sleepdf$trubslep_num)

```

# Task 1: Build a Simple Logistic Regression: Can trouble sleeping be predicted by whether someone is a smoker?

## Step 1: Build the model
```{r buildmodel1}
#Make sure categorical data is used as factors

logmodel1 <- glm(trubslep_num~ smoke, data =sleepdf, na.action = na.exclude, family = binomial(link=logit))
```

## Step 2: Generate Summary Statistics to Assess Model Fit
### Step 2a: Summary of Model
```{r summarymodel1}
#Full summary of the model
summary(logmodel1)
```

### Step 2b: Omnibus Test of Model
```{r lrtestmodel1}
#Chi-square plus significance
lmtest::lrtest(logmodel1)
```

### Step 2c: Pseudo R2 Statistics
```{r pr2model1}
#Pseudo Squared 
DescTools::PseudoR2(logmodel1, which="CoxSnell")
DescTools::PseudoR2(logmodel1, which="Nagelkerke")
performance::r2(logmodel1) # Will return Tujur's Pseudo R2
```

## (Optional): Use Stargazer to Report
```{r stargmdl1}
#Summary of the model with co-efficients
stargazer(logmodel1, type="text", star.cutoffs = c(0.05, 0.01, 0.001))
```


### Your task: Interpret the fit and usefulness


## Step 3: Calculate Odds Ratios and Probability of Co-efficients
```{r predictorsmdl1}
# Odds ratios only 
parameters::parameters(logmodel1, exponentiate = TRUE) %>%
  as.data.frame() %>%
  dplyr::select(Parameter, OddsRatio = Coefficient)

# Predicted probabilities by smoke
emmeans::emmeans(logmodel1, ~ smoke, type = "response")

```

### Your task: Interpret the odds ratios and predicted probabilities



## Step 4: Assess Classification Performance
## Step 4a: Generate a ROC curve and Calculate Sensitivity, Specificity and AUC
```{r rocm1}
#| fig.cap: "Figure 1: ROC curve for Model1"

# Generate predicted probabilities for Model 1
p_hat <- stats::predict(logmodel1, type = "response")

# Build results1 data frame for Epi::ROC
results1 <- sleepdf %>%
  dplyr::mutate(
    trubslep_num = trubslep_num,   # already created earlier
    p_hat = p_hat                # predicted probabilities
  ) %>%
  dplyr::select(trubslep_num, p_hat) %>%     # only keep needed columns
  dplyr::filter(!is.na(trubslep_num))        # remove NA cases

# Generate the curve
roc1 <-Epi::ROC(
  form = trubslep_num ~ p_hat,
  data = results1,
  plot = "ROC"
)

```
### Your Task: Interpret the ROC curve, sensitivity, specificity, AUC



### Step 4b: Generate a Confusion Matrix
```{r}
#| tbl-cap: "Table 1: Confusion Matrix Model 1"
results1 <- sleepdf %>%
  dplyr::mutate(
    actual = factor(trubslep_num, levels = c(0,1)),
    pred_class = factor(
      ifelse(stats::predict(logmodel1, type = "response") > 0.5, 1, 0),
      levels = c(0,1)
    )
  )


cm <- yardstick::conf_mat(
  data = results1,
  truth = actual,
  estimate = pred_class
)

cm


```
### Your task: Interpret the Confusion Matrix


## Step 5: Check Model Meets Assumptions of Logistic Regression
### Step 5a: Linearity of Independent Variables and Log Odds
```{r hlml1}
#Check the assumption of linearity of independent variables and log odds using a Hosmer-Lemeshow test, if this is not statistically significant we are OK
generalhoslem::logitgof(sleepdf$trubslep_num, fitted(logmodel1))
```
### Step 5b: Collinearity 
```{r clmdl1}
#We would check for collinearity here but as we only have one predictor it doesn't make sense - check model 2 to see how to do this

```


### Step 5c: Outliers and Influential Cases
- **Cook’s Distance**: Measures how much each individual observation influences the overall model; large values indicate cases that may disproportionately affect the results.

- **Leverage (Hat values)**: Identifies observations with unusual predictor values; high-leverage cases have the potential to pull the model toward themselves.

- **Standardised (Pearson) residuals**: Show how poorly each observation is predicted by the model; large residuals (e.g., > |2|) indicate unusual or poorly fitted cases.

- **Deviance residuals**: Reflect how much each observation contributes to the model’s deviance; large values highlight observations the model struggles to accurately classify.

- **DFBetas**: Measure how much each observation influences individual regression coefficients; large DFBetas indicate that removing the case would substantially change a coefficient estimate.


```{r olmodel1}
# 1. Cook's Distance
cook <- cooks.distance(logmodel1)

# View largest Cook's D values
head(sort(cook, decreasing = TRUE))

# Flag influential cases (common threshold: D > 1)
which(cook > 1)


# 2. Leverage (Hat values)
lev <- hatvalues(logmodel1)

# View highest leverage cases
head(sort(lev, decreasing = TRUE))

# Flag high leverage cases
# Rule of thumb: h_i > 2*(k+1)/n
k <- length(coef(logmodel1)) - 1  # number of predictors
n <- nobs(logmodel1)
lev_threshold <- 2*(k+1)/n

#Count how many exceed threshold
sum(lev > lev_threshold)

#Count how many > .2
sum(lev > 0.2)


which(lev > lev_threshold)


# 3. DFBetas (influence on individual coefficients)
dfb <- dfbeta(logmodel1)

# Identify cases with large DFBetas (|DFB| > 1 is concerning)
which(abs(dfb) > 1, arr.ind = TRUE)


# 4. Standardised (Pearson) residuals
pearson_resid <- rstandard(logmodel1, type = "pearson")

# Flag large residuals (|residual| > 2 or > 3)
which(abs(pearson_resid) > 2)

# Optional: look at the largest ones
head(sort(abs(pearson_resid), decreasing = TRUE))


# 5. Deviance residuals
dev_resid <- residuals(logmodel1, type = "deviance")

# Flag extreme deviance residuals
which(abs(dev_resid) > 2)

```
### Your Task: Interpret the assumption checks



## Step 6: Report
### Your task: Write the report




## Conclusion Model 1

### Your Task: What can you conclude

# Task 2: Build a Simple Logistic Regression with Multiple Predictors: Can trouble sleeping be predicted by whether someone is a smoker and their caffeine intake?

## Step 1: Build the model
```{r buildmodel2}
#Make sure categorical data is used as factors

logmodel2 <- glm(trubslep_num~ smoke+caffeine, data = sleepdf, na.action = na.exclude, family = binomial(link=logit))
```

## Step 2: Generate Summary Statistics to Assess Model Fit
### Step 2a: Summary of Model
```{r summarymodel2}
#Full summary of the model
summary(logmodel2)
```

### Step 2b: Omnibus Test of Model
```{r lrtestmdl2}
#Chi-square plus significance
lmtest::lrtest(logmodel2)
```

### Step 2c: Pseudo R2 Statistics
```{r pr2mdl2}
#Pseudo Squared 
DescTools::PseudoR2(logmodel2, which="CoxSnell")
DescTools::PseudoR2(logmodel2, which="Nagelkerke")
performance::r2(logmodel2) # Will return Tujur's Pseudo R2
```

## (Optional): Use Stargazer to Report
```{r stargmdl2}
#Summary of the model with co-efficients
stargazer(logmodel2, type="text", star.cutoffs = c(0.05, 0.01, 0.001))
```


### Your task: Interpret the fit and usefulness



## Step 3: Calculate Odds Ratios and Probability of Co-efficients
```{r predictorsmdl2}
# Odds ratios only 
parameters::parameters(logmodel2, exponentiate = TRUE) %>%
  as.data.frame() %>%
  dplyr::select(Parameter, OddsRatio = Coefficient)

# Predicted probabilities by smoke
emmeans::emmeans(logmodel2, ~ smoke, type = "response")

# Predicted probabilities by caffeine
emmeans::emmeans(logmodel2, ~ caffeine, type = "response")

# Full grid of predictions for ALL combinations

pred_grid <- emmeans::emmeans(
  logmodel2,
  ~ smoke * caffeine,   # <-- interaction in predictions, even if not in model
  type = "response"
)
```

### Your task: Interpret the odds ratios and predicted probabilities


## Step 4: Assess Classification Performance
### Step 4a: Generate a ROC curve and Calculate Sensitivity, Specificity and AUC
```{r}
#| fig.cap: "Figure 2: ROC curve for Model 2"

# Generate predicted probabilities for Model 2
p_hat <- stats::predict(logmodel2, type = "response")

# Build results2 data frame for Epi::ROC
results2 <- sleepdf %>%
  dplyr::mutate(
    trubslep_num = trubslep_num,   # already created earlier
    p_hat = p_hat                # predicted probabilities
  ) %>%
  dplyr::select(trubslep_num, p_hat) %>%     # only keep needed columns
  dplyr::filter(!is.na(trubslep_num))        # remove NA cases

# Generate the curve
roc1 <-Epi::ROC(
  form = trubslep_num ~ p_hat,
  data = results2,
  plot = "ROC"
)

```
### Your Task: Interpret the ROC curve, sensitivity, specificity, AUC

### Step 4b: Generate a Confusion Matrix
```{r}
#| tbl-cap: "Table 2: Confusion Matrix Model 1"
results2 <- sleepdf %>%
  dplyr::mutate(
    actual = factor(trubslep_num, levels = c(0,1)),
    pred_class = factor(
      ifelse(stats::predict(logmodel2, type = "response") > 0.5, 1, 0),
      levels = c(0,1)
    )
  )


cm <- yardstick::conf_mat(
  data = results2,
  truth = actual,
  estimate = pred_class
)

cm


```
### Your task: Interpret the Confusion Matrix


## Step 5: Check Model Meets Assumptions of Logistic Regression
### Step 5a: Linearity of Independent Variables and Log Odds
```{r hlml2}
#Check the assumption of linearity of independent variables and log odds using a Hosmer-Lemeshow test, if this is not statistically significant we are OK
generalhoslem::logitgof(sleepdf$trubslep_num, fitted(logmodel2))
```

### Step 5b: Collinearity 
```{r clmdl2}
#Check collinearity

vifmodel<-car::vif(logmodel2)#You can ignore the warning messages, GVIF^(1/(2*Df)) is the value of interest
vifmodel
#Tolerance
1/vifmodel

```
### Step 5c: Outliers and Influential Cases
- **Cook’s Distance**: Measures how much each individual observation influences the overall model; large values indicate cases that may disproportionately affect the results.

- **Leverage (Hat values)**: Identifies observations with unusual predictor values; high-leverage cases have the potential to pull the model toward themselves.

- **Standardised (Pearson) residuals**: Show how poorly each observation is predicted by the model; large residuals (e.g., > |2|) indicate unusual or poorly fitted cases.

- **Deviance residuals**: Reflect how much each observation contributes to the model’s deviance; large values highlight observations the model struggles to accurately classify.

- **DFBetas**: Measure how much each observation influences individual regression coefficients; large DFBetas indicate that removing the case would substantially change a coefficient estimate.


```{r olmodel2}
# 1. Cook's Distance
cook <- cooks.distance(logmodel2)

# View largest Cook's D values
head(sort(cook, decreasing = TRUE))

# Flag influential cases (common threshold: D > 1)
which(cook > 1)


# 2. Leverage (Hat values)
lev <- hatvalues(logmodel2)

# View highest leverage cases
head(sort(lev, decreasing = TRUE))

# Flag high leverage cases
# Rule of thumb: h_i > 2*(k+1)/n
k <- length(coef(logmodel2)) - 1  # number of predictors
n <- nobs(logmodel2)
lev_threshold <- 2*(k+1)/n

#Count how many exceed threshold
sum(lev > lev_threshold)

#Count how many > .2
sum(lev > 0.2)

which(lev > lev_threshold)


# 3. DFBetas (influence on individual coefficients)
dfb <- dfbeta(logmodel2)

print(paste("Threshold is: ", lev_threshold))



# Identify cases with large DFBetas (|DFB| > 1 is concerning)
which(abs(dfb) > 1, arr.ind = TRUE)



# 4. Standardised (Pearson) residuals
pearson_resid <- rstandard(logmodel2, type = "pearson")

# Flag large residuals (|residual| > 2 or > 3)
which(abs(pearson_resid) > 2)

# Optional: look at the largest ones
head(sort(abs(pearson_resid), decreasing = TRUE))


# 5. Deviance residuals
dev_resid <- residuals(logmodel2, type = "deviance")

# Flag extreme deviance residuals
which(abs(dev_resid) > 2)

```

### Your Task: Interpret the assumption checks




## Step 6: Report
### Your task: Write the report

### Your task: Write a Conclusion about the model


## Your task: Write a comparison between model 1 and 2