---
title: "Multiple Linear Regression Report"
subtitle: "TU059 - Determining water quality metrics linear regression"
author: "STUDENT NUMBER: D24127620, STUDENT NAME: Eamonn Kelly, PROGRAMME CODE: TU059"
format: 
  html:
    self-contained: true
    toc: true      # Enables table of contents
    toc_float: true # Makes the TOC float on the side
    toc_depth: 3   # Specifies the depth of headers to include in TOC
editor: source

---


```{=html}
<style>
/* Center figure captions in Quarto HTML, make them black and bold */
.quarto-figure .figure-caption, 
figure > figcaption { 
  text-align: center; 
  color: black; 
  font-weight: bold; 
}
</style>
```


```{r}
source("MATH9102-W9-Functions.R")
#| label: install-and-load
#| message: false
#| warning: false

quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE)
    }
    library(pkg, character.only = TRUE)
  }))
}


quiet_library("tidyverse")  # Collection of R packages (dplyr, ggplot2, tidyr, etc.) for data wrangling, visualization, and analysis
quiet_library("lm.beta")   # For extracting standardized co-efficients from regression model
quiet_library("stargazer") # For formatting output of regression
quiet_library("performance")    # Loads tools for checking model assumptions and diagnostics (e.g., normality, collinearity)
quiet_library("car")            # Companion to Applied Regression – includes functions for influence tests and outlier detection
quiet_library("see")            # Needed for plots
quiet_library("ggResidpanel")   # Alternative for generating diagnostic plots for 4.5.x
quiet_library("stats")   # Core R statistical functions (cor.test, oneway.test, bartlett.test)
quiet_library("rstatix")   # User-friendly statistical tests (Tukey HSD post-hoc)
quiet_library("effectsize")   # Computes and interprets effect sizes (e.g., eta squared, Cohen’s d)
quiet_library("psych")  # Descriptive statistics for grouped data (describeBy)
quiet_library("ggplot2")   # Data visualization (histograms, QQ plots, diagnostic plots)
quiet_library("VIM")   # Visualization of missing data patterns (aggr)
quiet_library("kableExtra")   # Enhances tables created with knitr::kable (styling, layout)
quiet_library("knitr")   # Generates formatted tables and report outputs (kable)
quiet_library("performance")   # Regression diagnostics and model checks (check_model, check_normality)
quiet_library("GGally")   # ?????

# Code to load file
datalocation <- file.path("./")
datapath <- file.path(datalocation,"WQI Results on Dataset.csv")

#Read in the csv file dataset
wqi_df <- read.csv(datapath, header=TRUE)
#Setting the column names to be that used in the dataset but in lowercase to make life a bit easier
colnames(wqi_df) <- tolower(colnames(wqi_df))


# Complete any manipulation/cleaning needed
```


```{r}
#| label: craete_data_subset
#
# Define variables that we want to use in or analysis
# we'll craete a subset of the data with just those to keep our dataset as light as we can.
varsint <- c("waterbodyname",  "years",  "sampledate",  "alkalinity.total..as.caco3.",  "bod...5.days..total.", "ammonia.total..as.n.", "chloride", "conductivity..25.c",  "dissolved.oxygen",  "ph",  "temperature",  "total.hardness..as.caco3.", "true.colour", "ccme_wqi",  "wawqi",  "horton.wqi",  "brown.wqi",  "srdd.wqi")

# create data subset of variables we want to lok at
wqi_dsubset <- wqi_df[varsint]


#  human friendly labels
colnames(wqi_dsubset) <- c( "waterbodyname", "years", "month", "alkalinity",  "biochem_ox_dmnd", "ammonia", "chloride", "conductivity", "dissolved.oxygen", "ph",  "temperature", "hardness", "colour", "ccme_wqi",  "wawqi",  "horton.wqi",  "brown.wqi",  "srdd.wqi")

```


## 1. Introdction

This analysis uses water quality data which was taken over a period of 16 years from 2007 to 2023 across 16 locations in Ireland by the Environmental Protection Agency(EPA). The data was then used to classify water quality at these locations. The data set was sourced from the following location 
[Water Quality Monitoring Dataset (Ireland)](https://figshare.com/articles/dataset/Water_Quality_Monitoring_Dataset_Ireland_/25002131), and is a merging of multiple csv files from the EPA into a single source. 

Water samples were taken from 160 sites within these 16 geographical areas, at various times throughout the year, tested in labs, and calculations were then carried out to determine Water Quality according to different water quality standards, a number of which exist.

### Purpose of Analysis
The analysis will consist of two parts

#### Task 1 - MLR Overview

The purpose of this part analysis is to investigate the extent to which key water quality measurements can explain variation in the **biochemical oxygen demand (BOD)** variable, which is a widely used metric to indicate water quality and will be used as our **outcome** variable.
- High 'BOD' values indicating increased oxygen consumption by microbes breaking down organic material in the water, and thus less oxygen being available for living organisms in the water such as fish and aquatic plants, meaning lower water quality. It can be seen as a measure of microbe activity, the higher the BOD value the more microbe activity.
- Lower values of 'BOD', indicating more oxygen being available in the water, providing a higher level of water quality. 

For *predictor variables* in Task 1, we will use the following continuous integer or ratio variables:
- **dissolved oxygen (DO)**: a key water quality measurable providing the amount of oxygen present in water. We expect to see the more dissolved oxygen present, the better the water quality, thus the lower the BOD value.
- **temperature**: influences how soluble substances are within water, including oxygen and organic matter. This should affect BOD value at the extremes higher and lower.
- **ph**: acidity  / alkaline indicator which affects support for various processes, both chemical and organic. Similar to temperature we expect to see poorer water quality at extreme values of **ph**.

A **differential (interaction) effect* will be included in our MLR model. The impact of temperature on the BOD value will depend on the amount of DO present in the water. 
- At low DO, increased temperature will increase the microbe activity, i.e. resulting in a higher BOD value, 
- At High DO, the temperature effect will less of an impact on BOD

#### Task 3 - MLR Comparison Overview

The purpose of this comparison model analysis is to determine if the relationship between the outcome variable, **biochemical oxygen demand (BOD)**, and the predictor variables, **dissolved oxygen**,**temperature** and  **ph**, changes per month. This will be achieved by adding a predictor variable, a categorical predictor. **month**, which is the month in which the data was taken. Month will be treated as a 'nominal' variable, with *Jan* used as the reference category and dummy variables used or the remaining months, allowing us to determine the differential effects in relation to that month throughout the year.

### Task 1-MLR-Hypothesis

- H1: **dissolved oxygen**,**temperature** and  **ph** significantly predict **biochemical oxygen demand (BOD)** and the effect of temperature on BOD is significantly moderated by the interaction effect between temperature and DO, present in the water.


### Task 3-MLR Comparison - Hypothesis

H2 - **month** as a categorical predictor significantly explains seasonal variations in the model predictions, indicating that BOD levels vary per month, relative to January.


Throughout the analysis, a significance threshold of 0.05 was used, and effect sizes were reported according to Cohen's conventions(1988).

Before interpreting the multiple linear regression results, diagnostic checks were conducted to ensure compliance with the key statistical assumptions underpinning the analysis. These included tests for linearity, normality of residuals, homoscedasticity, multicollinearity, and the presence of outliers, leverage points, or influential observations (Field, Miles, & Field, 2021; Tabachnick & Fidell, 2019). Linearity and homoscedasticity were assessed through examination of residual plots, where a random distribution of residuals around the zero line indicated that the assumptions were met. The normality of residuals was evaluated both visually, using Q–Q plots, and statistically, with small deviations tolerated given the robustness of regression to minor non-normality in larger samples (Field et al., 2021). Standardized residuals were inspected to identify potential outliers, with absolute values exceeding ±3.29 considered cause for further investigation (Tabachnick & Fidell, 2019). To assess influence, Bonferroni-adjusted tests of studentized residuals and Cook’s distance statistics were applied; Cook’s distance values greater than 1.0 were regarded as indicative of influential outliers, while leverage values were examined to identify cases exerting disproportionate influence on the regression coefficients (Belsley, Kuh, & Welsch, 1980; Field et al., 2021). Multicollinearity was assessed using Variance Inflation Factor (VIF) and Tolerance values. Following Field, Miles, and Field (2021), VIF values greater than 5 or Tolerance values below 0.2 were considered indicative of potential multicollinearity, while more extreme thresholds (VIF \> 10, Tolerance \< 0.1) would indicate serious violations. Collectively, these diagnostic procedures confirmed that the data met the necessary assumptions for valid multiple regression analysis, providing confidence in the reliability of the model estimates.


## 2. Data Description

The dataset contains '29,159' rows and '18' variables typically included in water quality analysis such as **ph**, **temperature**, **biochemical oxygen demand *BOD**, **dissolved oxygen**, **alkalinity** and others. The concepts of interest, their measurement and statistical type are shown in Tables 1 and 2 below.


| variable         | r_type  | is_factor | factor_summary                           | stat_type             |
| ---------------- | ------- | --------- | ---------------------------------------------------------------- | 
| waterbodyname    | factor  | TRUE      | 160 levels (e.g., ABBEYTOWN_010, Allua,) | Categorical (Nominal) |
| years            | integer | FALSE     | NA                                       | Discrete (Ratio)      |
| month            | factor  | TRUE      | 12 levels (e.g., Jan, Feb, …)            | Categorical (Ordinal) |
| alkalinity       | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| biochem_ox_dmnd  | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| ammonia          | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| chloride         | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| conductivity     | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| dissolved.oxygen | double  | FALSE     | NA                                       | Continuous (Ratio)    |
| ph               | double  | FALSE     | NA                                       | Continuous (Interval) |
| temperature      | double  | FALSE     | NA                                       | Continuous (Interval) |
| hardness         | double  | FALSE     | NA                                       | Continuous (Ratio) |
| colour           | double  | FALSE     | NA                                       | Continuous (Ratio) |
| ccme_wqi         | factor  | TRUE      | 5 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| wawqi            | factor  | TRUE      | 6 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| horton.wqi       | factor  | TRUE      | 6 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| brown.wqi        | factor  | TRUE      | 5 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |
| srdd.wqi         | factor  | TRUE      | 6 levels (e.g., Excellent, Fair, …)      | Categorical (Ordinal) |

Table 1. Variable and Statistical Datatypes present in the data set


We will create a reduced dataset to include primarily just the variables required for our two models, namely, **month**, **dissolved.oxygen**, **ph** and **temperature**.


| Variable | Range (Min–Max) | Mean | SD | Median | IQR (1st Qu. - 3rd Qu| 
|-----------|-----------------|------|----|---------|--------------|
| **Biochemical Oxygen Demand (BOD, mg/L)** | 0.0–16.0 | 1.28 | 0.65 | 1.20 | 1.2–1.2 | 
| **Dissolved Oxygen (mg/L)** | 0–198 | 58.29 | 22.5 | 54.9 | 49.4–75.0 |
| **pH** | 4.7–9.8 | 7.55 | 0.7 | 7.7 | 7.2–8.0 | 
| **Temperature (°C)** | 0.6–637.0* | 10.85 | 2.5 | 10.7 | 8.2–13.1 |


Table 2. Continuous (Interval/Ratio) Data Descriptions

**Some General observations:**
- **Biochemical Oxygen Demand (BOD, mg/L)** - very stable data, 75% of data below 1.2 value, there is a potential outlier present though
- **ph** - mean values
- **Temperature (°C)** - range from just above freezing and 3rd quantile approx 13 C. There is an outlier here.

There do appear to be some outliers, particularly in temperature, with a temperature of 637 C not be a real value. This ms be due to a broken thermometer, or perhaps data entry error. Conductivity, has a 3rd quantile value of 501 yet has a max value of 42000. This could be an outlier however we need to determine if extreme values are outliers or actual real data due to extreme environmental conditions

Visual inspection of the histograms and Q–Q plots for all predictor and outcome variables wre preformed. The standardized scores for skewness and kurtosis, as well as the percentages of absolute z-scores exceeding ±3.29, were inspected.

Also, **month** values are listed as categorical ordinal in the earlier table, we will treat month as nominal categorical data in or model. This will allow us do comparison of monthly differences without have an implied linear relationship. We have sufficient data per month to allow us include as a categorical predictor.

As per the assignment phase 1 analysis the following steps were applied to create our working data set for this task.


### Clean and Prepare the Data

```{r}
#| label: clean_prepare_dataset
# Temperature variable -  delete these rows 
wqi_dsubset_clean_t <- wqi_dsubset %>% filter(temperature <= 40 | is.na(temperature))

# Biochemical Oxygen Demand (BOD) variable -  transform to the 'log10'.
wqi_dsubset_clean_t_bod <- wqi_dsubset_clean_t %>%
  # add +1 to handle zeros safely and rename the variables
  mutate(bod_log10 = log10(biochem_ox_dmnd + 1))


```


```{r}
#| label: create_mlr_dataframe
mlr_t1_t3_df <- wqi_dsubset_clean_t_bod %>% select(month, bod_log10, dissolved.oxygen, temperature, ph)

# # Dataset for LogReg
# logr_t2_df <- wqi_dsubset_clean_t_bod %>%
#   mutate(high_bod = ifelse(biochem_ox_dmnd > 5, 1, 0)) %>%
#   select(high_bod, dissolved.oxygen, temperature, conductivity)
# 
# # Dataset for Dimension Reduction
# dimr_t3_df <- wqi_dsubset_clean_t %>%

# cat("number of rows = ", nrow(mlr_t1_t3_df), "\n")
# cat("number of cols = ", ncol(mlr_t1_t3_df))
summary(mlr_t1_t3_df)
```

Our final working dataset contains '29,154' rows and '5' variables

As per the assignment phase 1 analysis, the following steps were applied to create our working data set for this task along with any additional considerations mentioned below.
- **BOD**
  - To address the skewness we applied a transform to the 'log10'.
- **Temperature**
  - removed any temp value above 40 degrees, to ensure we have realistic temp values and remove outliers.
- **Dissolved Oxygen** 
  - Analysis showed the DO data to be non normal and bimodal or multi-modal.  
  - We will not apply a transformation to this variable data, rather consider it non-parametric.
  - As predictor variables are not required to be normally distributed in linear regression, we leave this in its
  original state for this variable. 
- **ph**
  - No Changes made
- **month**
  - no changes 


## Exploratory Data Analysis

Exploratory data analysis was carried out to seek eveidcen supporting and to help shape and validate decisions for predictor ad outcome variables prior to the building of the model

### 2.1 Missing Data

```{r}
# Prepare data for missingness check
mdata <- mlr_t1_t3_df %>% select(bod_log10, dissolved.oxygen, temperature, ph) %>% as.data.frame()

# Reset row names (this fixes the error)
row.names(mdata) <- NULL

# Missing data inspection
VIM::aggr(mdata,sortVar = TRUE, combined = FALSE)

```

There was no missing data present in the dataset.

### 2.2 Visualizations of variables


```{r assess-p1}
#| label: assess-p2
# Generate plots using function plot_ratio_variable
plot_ratio_variable(mlr_t1_t3_df, "bod_log10", "BOD", 1)
plot_ratio_variable(mlr_t1_t3_df, "dissolved.oxygen","DO",2)
plot_ratio_variable(mlr_t1_t3_df, "temperature", "temp", 3)
plot_ratio_variable(mlr_t1_t3_df, "ph", "ph", 4)

```

The above Historgam and QQ plots of our variables provide evidence supporting our decisions to transform **BOD** to Log10, to address positive skewness and non normality. We can visually see the presence of multi modal distribution in **DO** data. No transformation was applied as is not required to fit to linear regression model, and we retain our ability to interpret the data more easily. The**Temp** data being our most evenly distributed and closest to normality, after havign removed some outliers. And finally, our ph values havign a near normal distribution

The QQ plot for BOD also shows a horizontal plateau region in the data, this could be an expected effect de to environmental factors, or also potentially due to a testing practice or methodology and could be a limiting factor 

As ws assessed in the previous assignment, using the guidance of Field, Miles and Field (2012), if 98% of the data fell between +/- 3.29 the distribution was considered to be approximately normal, and Skewness and Kurtosis values between –2 and +2 ≈ were acceptable


### 2.2 Colineaity of variables 
The relationships between the concepts of interest were investigated before linear regression models were constructed to ensure that a linear relationship could be considered to exist between the proposed dependent variable and proposed predictors.


```{r}
#| label: cortest
r1=stats::cor.test(mlr_t1_t3_df$dissolved.oxygen, mlr_t1_t3_df$bod_log10)
r2=stats::cor.test(mlr_t1_t3_df$temperature, mlr_t1_t3_df$bod_log10)
r3=stats::cor.test(mlr_t1_t3_df$ph, mlr_t1_t3_df$bod_log10)

```

Pearson correlations were conducted to asses the level of relationship between **BOD** and each of the predictor variables,  **dissolved.oxygen (DO)**, **temperature (Temp)**, and **ph**. There is a statistically significant, positive, weak correlation between **BOD** and **DO**. (p < 0.05, Corr.Co-eff(r) = 0.02, ). There is a statistically significant, positive, weak correlation between **BOD** and **Temp**. (p < 0.05, Corr.Co-eff(r) = 0.1, ). There is a statistically significant, positive, very weak correlation between **BOD** and **ph**. (p < 0.05, Corr.Co-eff(r) = 0.07, ). 

All correlations are statistically significant at a weak level (see Table 5). However we do have a large sample size, and as each predictor value increase we expect the **BOD** value to also increase, thus providing supporting evidence for our intended inclusion of these predictor variables in our linear regression model.

```{r}

#| tbl-cap: "Table 5: Correlations between BOD and Precictor Variables"

# Create a data frame for the correlation results

corr_table <- data.frame(
Dependent = c("bod_log10", "bod_log10", "bod_log10"),
Predictor = c("dissolved.oxygen", "temperature", "ph"),
`Correlation Coefficient (r)` = c(
round(r1$estimate, 2),
round(r2$estimate, 2),
round(r3$estimate, 2)
),
Significance = c("p < 0.001", "p < 0.001", "p < 0.001")
)

knitr::kable(
corr_table)
```

### 2.4 Visualitaions of linear relationships between outcoe and predictor variables

Scatterplots of the outcome variabel and predictor variabels demonstrate linear trends for each of thbe predictor variables, aai supportig their use in our mdoel craetion. 

```{r}

#| label: fig-corrplot
#| fig-cap: "Visual Relationship between Predictors and BOD"

# Create a small subset for the plot
plot_df <- mlr_t1_t3_df[, c("bod_log10", "dissolved.oxygen", "temperature", "ph")]

# Create the plot matrix
ggpairs(plot_df, 
        lower = list(continuous = wrap("points", alpha = 0.3, size=0.5, color="blue")),
        diag = list(continuous = wrap("barDiag", fill="skyblue")))

```


```{r}

mlr_long <- mlr_t1_t3_df %>%
  select(bod_log10, dissolved.oxygen, temperature, ph) %>%
  pivot_longer(cols = -bod_log10, names_to = "predictor", values_to = "value")

#| label: scatter_faceted
#| fig.cap: "Figure 15: Scatterplots of BOD against predictors"
#| fig.width: 12
#| fig.height: 6

ggplot(mlr_long, aes(x = value, y = bod_log10)) +
  geom_point(alpha = 0.4, colour = "steelblue") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(x = "Predictor value", y = "BOD (log₁₀ mg/L)") +
  theme_minimal()
```
### Anova - Bartley test for variance on cat data  - month

A Bartlettes test of homogeneity of variances was carried out on thecategorical variable we are considering using as a predictor, i.e. **month** against **BOD**, our outcome variable. The test returned values of (K^2 = 1038, df=11,  p-value < 2.2e-16. The effect size is quite small at Eta squared = 8.08*10^-3 (i.e. approx = 0.008), meaning while it is statistically significant (p < 0.05) it only explains > 1% of the total BOD variance. 

Overall this provides support to the decision to include month as a categorical differential (interraction ) effect predictor and we can perform further analysis once our model has been created.


```{r}
#| label: variance-test
#Get descriptive statistics by group - output as a matrix
# bOd by nonth
monthdesc<-psych::describeBy(mlr_t1_t3_df$bod_log10, group=mlr_t1_t3_df$month, mat=TRUE)

#Conduct Bartlett's test for homogeneity of variance  
stats::bartlett.test(bod_log10 ~ month, data=mlr_t1_t3_df)

#Conduct ANOVA using the one-way ANOVA from the stats package
# we change var.equal to FALSE as environment data will typically have variances across month
# i.e. they will not be the same
res <- stats::oneway.test(bod_log10 ~ as.factor(month), data = mlr_t1_t3_df, var.equal = TRUE)

#We use Tukey for post-hoc testing
tukeyres <-rstatix::tukey_hsd(mlr_t1_t3_df, bod_log10 ~ month)

#Compute our effect size - Eta squared
# gives magntiude of the month effect on BOD
effes=effectsize::effectsize(res)
effes

#Store the relevant pieces of the output from ANOVA in variables to use for reporting
#Degrees of freedom
df1=res$parameter[1]
df2=res$parameter[2]
#F statistic
Fstat=round(res$statistic, 3)
# Pvalue <- round(res$p.value)
pval=round(res$p.value,2)
eta2=round(effes$Eta2, 3)
```


# Step 1: Build the model
```{r}
#| label: clean_prepare_dataset
#Create the model 1 for task 1
# Thsi is our baseline model
model1_t1<-lm(bod_log10~dissolved.oxygen+temperature+ph+(temperature*dissolved.oxygen), data=mlr_t1_t3_df)
```

##  Step 2: Generate the Analysis of Variance Table
This provides the information needed to calculate the F statistic.
```{r}
#| label: anovam1
anova(model1_t1)
```

## Step 3: Generate the summary statistics needed to assess model fit
```{r}
#| label: summm1
summary(model1_t1)
```

## Step 4: Generate Standardized Co-efficients
In this example as the data is standardized these will not be different to the co-efficients produced by the summary.
```{r}
#| label: betam1
lm.beta::lm.beta(model1_t1)
```

## Step 5 (optional): Use Stargazer to produce a tidy output of ANOVA and Summary
```{r stargazer}
#| label: stargazer
stargazer(model1_t1, type="text") #Tidy output of all the required stats
```

## Step 6: Report the results
Replace the content in italics with content relevant to your analysis (does not need to be italicised). 


A multiple linear regression model was created to _explain what the model was created for_.  A _statistically/non statitsically_ significant regression equation was found (F(_df1,df2_)=_FRatio_, p = _pvalue (if not significant)/p < relevant cut-of (if significant < 0.5, 0.1 or 0.001)_ with an adjusted R^2^ of _adjusted R2 from summary_. The regression equation is therefore:

  _Outcome Concept = intercept + b1 * pred1 + b2 * pred2 + b3 * pred3_
  
This indicates _explain what increases/changes in each predictor mean for the outcome_. _Write a statement about what this means for the hypothesis_.
 
 
Remember for categorical predictors you are comparing to the reference category.



| Statistic needed in written report                      | Where to find it in R output                                                    | Code chunk label |     |          |
| ------------------------------------------------------- | ------------------------------------------------------------------------------- | ---------------- | --- | -------- |
| **F-statistic**                                         | In the **ANOVA table** (`F value` column)                                       | `anovam1`        |     |          |
| **Model degrees of freedom (df1, df2)**                 | In **ANOVA** output (`Df`) <br> *df1 = model predictors* <br> *df2 = residuals* | `anovam1`        |     |          |
| **p-value for the model (overall test)**                | In **ANOVA** (`Pr(>F)`) — same row as F-statistic                               | `anovam1`        |     |          |
| **Adjusted R²**                                         | In the **Regression Summary** (`Adjusted R-squared`)                            | `summm1`         |     |          |
| **Unstandardized coefficients** (Intercept, b1, b2, b3) | In **Regression Summary** (`Estimate`)                                          | `summm1`         |     |          |
| **p-values for predictors**                             | In **Regression Summary** (`Pr(>|t| )`) | `summm1` |
| **Standardized coefficients** (β values)                | In standardized regression output                                               | `betam1`         |     |          |
| **Optional formatted version of the above**             | Stargazer formatted summary output                                              | `stargazer`      |     |          |



# Residual Analysis

## Step 1: Generate Summary Statistics/Non-Zero Variance

```{r}
#Generate summary statistics
#If using   pastecs::stat.desc for continous variables - make sure you include the basic switch=F to ensure you don't get scienfitic notation


```

## Step 2: Residual Analysis

### Step 2a: Conduct a Normality Test

```{r}
# Check normality - conducts a Shapiro-Wilks test and checks the standardized residuals for normal distribution
# A p-value < 0.05 indicates a significant deviation from normal distribution.

performance::check_normality(model1)

```

### Step 2b: Visualize residual distribution

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Residual Distribution Plots for Model 1"
# Visualize residual distribution


# Histogram of Residuals
ggplot(data.frame(residuals = resid(model1)), aes(x = residuals)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 20, 
                 fill = "skyblue", 
                 color = "white") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(resid(model1)), sd = sd(resid(model1))), 
                color = "red", 
                linewidth = 1) +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Density") +
  theme_minimal()

# A transformed version of a Q–Q plot displayed horizontally
plot(check_normality(model1))

# Q-Q plot of standardized residuals
qqnorm(rstandard(model1))
qqline(rstandard(model1), col = "red")

```

### Step 2c: Check outliers or influential points

```{r}
# Check for outliers or influential points

performance::check_outliers(model1)

```

### Step 2d: Examine the residuals

```{r}
# Extract residuals
residuals <- resid(model1)

# Standardize the residuals
std_residuals <- residuals / sd(residuals)

# Calculate max and min of the standardized residuals
max_std_residual <- max(std_residuals)
min_std_residual <- min(std_residuals)

# Print the results
cat("Max standardized residual:", max_std_residual, "\n")
cat("Min standardized residual:", min_std_residual, "\n")

#Identify potential outliers in the dataset based on the studentized residuals of the model.
car::outlierTest(model1) # Bonferonni p-value for most extreme observations - observations with a p value lower than the threshold are consider outliers.
# In this case there is only one row 

```

## Step 3: Check homoscedasticity (equal variance)

```{r}
performance::check_heteroscedasticity(model1)

```

## Step 4: Check multi-colinearity

```{r}
# Check multicollinearity

performance::check_collinearity(model1)

```

## Step 5: Report
Replace the content in italics with content relevant to your analysis (does not need to be italicised).

The data met the assumption of non-zero variances _report the variance for your continous predictors_. Examination of diagnostic plots indicated that the residuals were _approximately normally distributed/not normally distributed_, with a density plot _resembling normality/deviating from normality_, although the formal test suggested _what the test suggested)_ _(p=?)_. The standardized residuals ranged from _report max and min_  with _x/no_ observations exceeding ±3 standard deviations. The Bonferroni-adjusted outlier test (p = ?) indicated that _what this meant_, as Cook’s distance _remained below/exceeded_ the conservative threshold of 1.0. The residuals-versus-fitted values plot displayed _describe the scatter_  around the zero line, with _some/no_ evidence of heteroscedasticity _(p = ?)_, _supporting/deviating from_ the assumption of constant variance. Collinearity statistics showed _low/moderate/hig_ intercorrelations between predictors _(VIF and Tolerance for each predictor)_, confirming that multicollinearity _was/was not_ a concern. Overall, the model _met/did not meet_ the key assumptions of multiple linear regression—linearity, independence, homoscedasticity, and low multicollinearity—with only a minor departure from normality and a single non-influential outlier. The model can therefore be considered _to/not to_ provide a reliable estimate of the relationship between _list the outcome and the predictor concepts_.


| Assumption / Statistic required in report     | What to look for                                             | Where to find it                                                          | Code chunk label             |
| --------------------------------------------- | ------------------------------------------------------------ | ------------------------------------------------------------------------- | ---------------------------- |
| **Normality of residuals**                    | p-value from Shapiro–Wilk or check_normality result          | Output of `performance::check_normality(model1)`                          | first in Step 2a |
|                                               | Visual inspection: *distribution resembles normality or not* | Histogram + Q-Q plots                                                     | Step 2b plots                |
| **Standardized residuals range**              | Max and min standardized residual values                     | Printed result of `cat(...)`                                              | Step 2d                      |
| **Outlier test**                              | Bonferroni p-value for most extreme case                     | Output of `car::outlierTest(model1)`                                      | Step 2d                      |
| **Influence**                                 | Cook’s Distance < 1.0                                        | In diagnostic plots (`plot(model1)` → Cook’s) or within `check_model()`   | Step 6                       |
| **Homoscedasticity (Equal variance)**         | p-value from heteroscedasticity test                         | `performance::check_heteroscedasticity(model1)` Scale-Location Plot                           | Step 3                       |
| **Linearity**                                 | Evidence of non-random scatter around fitted line            | Residuals vs Fitted plot in `plot(model1, which = 1)` or `check_model()`  | Step 6                       |
| **Multicollinearity**                         | VIF and Tolerance values                                     | `performance::check_collinearity(model1)`                                 | Step 4                       |



## Step 6: Generate Summary Diagnostic Plots

```{r}
#| fig.width: 12
#| fig.height: 10
#| fig.cap: "Diagnostic plots for Model 1"
# Overall diagnostic plots

# If using R 4.4.x use
performance::check_model(model1,
  check = c(
    "linearity",
    "homoscedasticity",
    "outliers",
    "collinearity",
    "normality"
  )
)

# If using R 4.5.x Show all 4 diagnostic plots in one figure using R basic plots
par(mfrow = c(2, 2))   # 2x2 grid layout
plot(model1)            # Generates: Residuals vs Fitted, Normal Q-Q, Scale-Location, Cook’s distance
par(mfrow = c(1, 1))    # Reset layout

```

```{r}
#| fig.cap: " Diagnostic plots for Model - Alternative approaches"
# Alternative for 4.5.x
car::residualPlots(model1)



ggResidpanel::resid_panel(model1)


```
